### 2014-02-06 inst/doc/article.Rnw
version 0.8-0

<<echo=FALSE, results=hide>>=
# This should be set to FALSE in case you want to redo the sampling as well.
# Default is TRUE to save CRAN some time.
usePreCalcResults <- TRUE
@

%\VignetteIndexEntry{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol}
%\VignetteKeyword{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}

\documentclass[article, nojss]{jss}

\usepackage{amsmath}
\usepackage{amssymb}
\usepackage{bm}

\input{macros.tex}

\graphicspath{{extrafig/}}

\interfootnotelinepenalty=10000

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%% declarations for jss.cls %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%% almost as usual
\author{Gregor Kastner\\WU Vienna University of Economics and Business}
%\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol} \\ (DRAFT VERSION)}
\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
%\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting
%\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol (DRAFT VERSION)} %% without formatting

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Gregor Kastner} %% comma-separated

%% an abstract and keywords
\Abstract{
Heteroskedasticity in financial and economic datasets is a commonly observed feature, and the need to model this feature properly has for a long time been of great interest to researchers and practitioners alike. The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modelling by means of the \emph{stochastic volatility} framework. Being a state-space formulation, it describes contemporaneous volatilities as latent random variables as opposed to deterministic values. The software described in this paper utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables, which can then be used for predicting future volatilities. The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers. Main focus of the paper is to show the functionality of \pkg{stochvol}, nevertheless it also provides a brief mathematical description of the model, an overview of the sampling schemes used, and an in-depth example using exchange rate data.
}
\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
%% at least one keyword must be supplied

%% publication information
%% NOTE: Typically, this can be left commented and will be filled out by the technical editor
%% \Volume{50}
%% \Issue{9}
%% \Month{June}
%% \Year{2012}
%% \Submitdate{2012-06-04}
%% \Acceptdate{2012-06-04}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Gregor Kastner\\
  Institute for Statistics and Mathematics\\
  Department of Finance, Accounting and Statistics\\
  WU Vienna University of Economics and Business\\
  Welthandelsplatz 1\\
  1020 Vienna, Austria\\
  E-mail: \email{gregor.kastner@wu.ac.at}
%  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
}
%% It is also possible to add a telephone and fax number
%% before the e-mail in the following format:
%% Telephone: +43/512/507-7103
%% Fax: +43/512/507-2851

%% for those who use Sweave please include the following line (with % symbols):
%% need no \usepackage{Sweave.sty}

%% end of declarations %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\shortcites{r:mvt}
%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = 'R> ', continue = '+  ')
@

\section{Introduction}
When analyzing (financial) returns, focus is often laid on estimating and predicting potentially time varying volatilities. This interest has a long history, dating at least back to \cite{mar:por}, who investigated portfolio construction with optimal expected return-variance trade-off. In his article, he proposes rolling-window-type estimates for the instantaneous volatilities, but already then recognizes the potential for ``better methods, which take into account more information''.

One way of doing so is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models. After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and were applied to a vast amount of real-world problems. As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one. Over time, this specification became known as the \emph{stochastic volatility (SV)} model. Even though several papers \citep[e.g.][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} find early evidence in favor of using SV, these models have found comparably little use in applied work. Reviewing this discrepancy, \cite{bos:rel} states:
\begin{quote}
 While there are literally thousands of applications of GARCH, for SV, this number is far lower. Two reasons for this relative lack of applied work using SV are apparent. First, there are as of yet no standard packages for estimating SV models, whereas for GARCH, most statistical packages have a wealth of options for incorporating GARCH effects. A second difference seems to be that GARCH has many variants of the model \citep{bol:glo}, with basically a single estimation method for all of them. For SV, there are few variants of the model, but a full series of estimation methods.
\end{quote}
In \cite{kas-fru:anc}, the latter issue has been thoroughly investigated, and an efficient MCMC estimation scheme
%using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
has been proposed. The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} were crafted to cope with the first problem: the apparent lack of standard packages for efficiently estimating SV models.

\section{Model specification and estimation}

We briefly introduce the model and specify the notation used in the remainder of the paper. Furthermore, a quick overview over Bayesian parameter estimation via Markov chain Monte Carlo (MCMC) methods is given.

\subsection{The SV model}

Let $\yb = (y_1, y_2, \dots, y_T)^\top$ be a vector of returns with mean zero. The intrinsic feature of the SV model is that each observation $y_t$ is assumed to have its ``own'' contemporaneous variance $\e^{\hsv_t}$, thus relaxing the usual assumption of homoskedasticity. In order to make estimation of such a model feasible (note that there are just as many data points as there are variances!), this variance is not allowed to vary unrestrictedly with time. Rather, its logarithm is assumed to follow a parametric stochastic process, more specifically: an autoregressive process of order one. Note that feature is fundamentally different to GARCH-type models, where the time-varying volatility is assumed to follow a deterministic rather than stochastic evolution.

The SV model (in its centered parameterization) is thus given through
\begin{eqnarray}  
 \label{c1}
 y_{t}|\hsv_t &\sim&  \Normal{0, \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \label{c2}
 \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapar^2},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{c3}
\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapar^2/(1-\phi^2)},
\end{eqnarray}

where $\Normal{\mu, \sigmapar^2}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapar^2$. We will refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} $\mupar$, the \emph{persistence} $\phipar$, and the \emph{volatility} $\sigmapar$ of log-variance. The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{T})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely: the log-variance process). Note that the initial state $\hsv_0$ is distributed according to the stationary distribution of the autoregressive process of order one.

\subsection{Prior distributions}
\label{priors}

To complete the model setup, a prior distribution for the parameter vector $\allpara$ needs to be specified. Following \cite{kim-etal:sto}, we choose independent components for each parameter, i.e., $p(\allpara) = p(\mupar)p(\phipar)p(\sigmapar)$.

The level $\mupar \in \mathbb{R}$ is equipped with the usual normal prior $\mupar \sim \Normal{b_{\mupar}, B_{\mupar}}$. In practical applications, this prior is usually chosen to be rather uninformative, e.g., through setting $b_\mupar = 0$ and $B_\mupar \geq 100$ for daily log-returns. Our experience with empirical data is that the exact choice is usually not very influential; see also Section~\ref{priorconfig}.

For the persistence parameter $\phipar \in (-1,1)$, we choose $(\phi+1)/2 \sim \Betadis{a_0, b_0}$, implying 
\begin{eqnarray}
 \label{beta_transformed}
 p(\phi) = \frac{1}{2\Betafun{a_0,b_0}}\left ( \frac{1+\phi}{2}\right ) ^{a_0-1}\left ( \frac{1-\phi}{2}\right ) ^{b_0-1},
\end{eqnarray}
where $a_0$ and $b_0$ are positive hyperparameters and $\Betafun{x,y} = \int_0^1t^{x-1}(1-t)^{y-1}\,dt$ denotes the beta function.
Clearly, the support of this distribution is the unit interval $(-1, 1)$; thus, stationarity of the autoregressive volatility process is guaranteed. Its expected value and variance are given through the expressions
\begin{eqnarray*}
 E(\phipar) &=& \frac{2a_0}{a_0+b_0}-1,\\
 V(\phipar) &=& \frac{4a_0b_0}{(a_0+b_0)^2(a_0+b_0+1)}.
\end{eqnarray*}
This obviously implies that the prior expectation of $\phipar$ depends only on the ratio $a_0:b_0$. It is greater than zero if and only if $a_0 > b_0$ and smaller than zero if and only if $a_0 < b_0$. For a fixed ratio $a_0:b_0$, the prior variance decreases with larger values of $a_0$ and $b_0$. The uniform distribution on the unit interval arises as a special case when $a_0=b_0=1$. For financial datasets with not too many observations (i.e., $T \lesssim 1000$), the choice of the hyperparameters $a_0$ and $b_0$ can be quite influential on the shape of the posterior distribution of $\phipar$. In fact, note that in the case when the underlying data-generating process is (near-)homoskedastic, the volatility of log-variance $\sigmapar$ will be (very close to) zero and thus the likelihood will contain little to no information about $\phipar$. Consequently, the posterior distribution of $\phipar$ will be (almost) equal to its prior, no matter how many data points are observed. For some discussion about this issue, see e.g., \cite{kim-etal:sto}, who choose $a_0=20$ and $b_0=1.5$, implying a prior mean of $0.86$ with a prior standard deviation of $0.11$ and thus very little mass for nonpositive values of $\phi$.

For the volatility of log-variance $\sigmapar \in \mathbb{R}^+$, we choose $\sigmapar^2 \sim B_{\sigmapar}\times \chi^2_1=\Gammad{1/2,1/2B_{\sigmapar}}$, which is motivated by \cite{fru-wag:sto}, who equivalently stipulate the prior for $\pm \sqrt { \sigmapar^2}$ to follow a centered normal distribution, i.e., $\pm \sqrt { \sigmapar^2} \sim \Normal{0,B_{\sigmapar}}$. As opposed to the usual Inverse-Gamma prior, this prior is not conjugate in the usual sampling scheme, however, does not bound $\sigmapar$ away from zero a priori. The choice of the hyperparameter $B_{\sigmapar}$ turns out to be of minor influence in empirical applications, as long as it is not picked to be too small.

\subsection{MCMC Sampling}

An MCMC algorithm such as the one implemented in the package \pkg{stochvol} will provide its user with draws from the posterior distribution of the desired random variables: in our case the latent log-variances $\hsvm$ and the parameter vector $\allpara$. Because these draws are usually dependent, Bayesian inference via MCMC may require careful design of the algorithm and attentive investigation of the draws obtained.

One key feature of the algorithm used in this package is the joint sampling of all instantaneous volatilities ``all without a loop'' (AWOL), a technique going back at least to \cite{rue:fas} and discussed in more detail in \cite{mcc-etal:sim}. Doing so reduces correlation of the draws significantly and requires auxiliary finite mixture approximation of the errors as in \cite{kim-etal:sto} or \cite{omo-etal:sto}.

In order to avoid the cost of code interpretation within each MCMC iteration, the core computations of the sampler are implemented in \proglang{C}, interfaced to \proglang{R} via the \pkg{Rcpp} package \citep{edd-fra:rcp}, where the convenience functions and the user-interface are implemented. This combination allows to make use of the well-established and widely accepted ease-of-use of \proglang{R} and its underlying functional programming paradigm. Moreover, existing frameworks for analyzing MCMC output such as \pkg{coda} \citep{plu-etal:cod} as well as high-level visualization tools can easily be used. Last but not least, users with a basic knowledge of \proglang{R} can use the package in a familiar surrounding with a very small entry cost. Nevertheless, despite all these convenience features, the package profits from highly optimized machine code generated by a compiler at package build time, thus providing acceptable runtime even for larger datasets.

A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of an ``ancillarity-sufficiency interweaving strategy'' (ASIS), which has been brought forward in the general context of state-space models by \cite{yu-men:cen}. ASIS exploits the fact that for certain parameter constellations, sampling efficiency improves substantially when considering a non-centered version of a state-space model.
This fact is commonly known as a reparameterization issue with an entire body of literature attached to it; for an early reference see e.g., \cite{hil-smi:par}.
In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility $\sigmapar$ of log-variance from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$. However, in the case of the SV model, it turns out that no single superior parameterization exists. Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better. To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered, and once in a noncentered parameterization. This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm. For details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.

\section[The stochvol package]{The \pkg{stochvol} package}

The usual stand-alone approach to fitting SV models with \pkg{stochvol} follows the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results. All these steps will be described in more detail below, along with a worked example. For a stepwise incorporation of SV effects into other MCMC samplers, please see Section \ref{other}.

\subsection{Preparing the data}
\label{prep}

The core sampling function \code{svsample} expects its input data \code{y} to be a numeric vector of returns without any missing values (\code{NA}s), and will throw an error if provided with anything else. In case that \code{y} contains zeros, a warning will be issued and a small offset constant of size \code{sd(y)/10000} will be added to the squared returns before doing the auxiliary mixture sampling \citep[cf.][]{omo-etal:sto}. A common and recommended way of avoiding zero returns is to de-mean the returns beforehand. Below is an example how to prepare data, illustrated with the \code{exrates} data set\footnote{The data set, which as been obtained from the European Central Bank's Statistical Data Warehouse, contains the daily bilateral prices of one Euro in 23 currencies from January 3, 2000, until April 4, 2012. Conversions to New Turkish Lira and Fourth Romanian Leu have been incorporated. See \code{?exrates} for more information.} which is included in the package. A visualization of one of these time series is displayed in Figure~\ref{fig1}.

<<usd1, fig=TRUE, include=FALSE, width=11, height=6>>=
set.seed(123)
library("stochvol")
data("exrates")
ret <- logret(exrates$USD, demean=TRUE)
par(mfrow=c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log-returns")
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd1}
 \caption{Visualization of EUR-USD exchange rates included in the \pkg{stochvol} package.}
 \label{fig1}
\end{center}
\end{figure}

Additionally to real-world data, \pkg{stochvol} has also a built-in data generator \code{svsim}. This function simply produces realizations of an SV process and returns an object of class \code{svsim}, which has its own \code{print}, \code{summary}, and \code{plot} methods. Exemplary code using \code{svsim} is given below, and the particular instance of this simulated series is displayed in Figure~\ref{fig2}.

<<usd2, fig=TRUE, include=FALSE, width=11, height=6>>=
sim <- svsim(500, mu = -9, phi = 0.99, sigma = 0.1)
par(mfrow=c(2, 1))
plot(sim)
@
\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd2}
 \caption{Visualization of a simulated time series as provided by the default \code{plot} method.}
 \label{fig2}
\end{center}
\end{figure}

 \subsection{Specifying prior distributions and configuration parameters}
 \label{priorconfig}

 After preparing the data vector \code{y}, the user needs to specify the prior hyperparameters for the parameter vector $\allpara = (\mupar, \phipar, \sigmapar)^\top$ -- see also Section~\ref{priors} -- and some configuration parameters. The appropriate values are passed to the main sampling function \code{svsample} as arguments, which are described below.

 The argument \code{priormu} is a vector of length $2$, containing mean and standard deviation of the normal prior for the level of the log-variance $\mupar$. A common strategy is to choose a vague prior here, e.g., \code{c(0, 100)}, because the likelihood usually carries enough information about this parameter. If one prefers to use (slightly) informative priors, e.g., to avoid outlier draws of $\mupar$, care must be taken whether actual log-returns or \emph{percentage} log-returns are analyzed. Assuming daily data, the former commonly have an unconditional variance of $0.0001$ or less and thus the level on the log-scale $\mupar$ lies around $\log(0.0001) \approx -9$, while the latter have the $100^2$-fold unconditional variance (around $1$), which implies a level of $\log(1) = 0$. Choices in the literature include \code{c(0, 10)} \citep{jac-etal:bayJE}, \code{c(0, 5)} \citep{yu:on}, \code{c(0, sqrt(10))} \citep{kim-etal:sto, mey-yu:bug} or \code{c(0, 1)} \citep{omo-etal:sto}. Note that most of these choices are quite informative and clearly designed for \emph{percentage} log-returns!

 For specifying the prior hyperparameters for the persistence of log-variance, $\phipar$, the argument \code{priorphi} may be used. It is again a vector of length $2$, containing $a_0$ and $b_0$ specified in Equation~\ref{beta_transformed}. As elaborated in Section~\ref{priors}, these values can possibly be quite influential, thus we advise to put some thought into choosing them and study the effect of different choices carefully. The default is currently given through \code{c(5, 1.5)}, implying a prior mean of $0.54$ and a prior standard deviation of $0.31$.

 The prior variance of log-variance hyperparameter $B_\sigmapar$ may be controlled through \code{priorsigma}. This argument defaults to \code{1} if not provided by the user. As discussed in Section~\ref{priors}, the exact choice of this value is usually not very influential in typical applications. In general, it should not be picked too small unless there is a very good reason, e.g., explicit prior knowledge, to do so.

For specifying the size of the burn-in, the parameter \code{burnin} is provided. It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain. The current default value for this parameter is $1000$, which has proved to suffice in most situations. Nevertheless, the user is encouraged to check convergence carefully, see Section \ref{check} for more details. The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$. Consequently, the sampler will be run for a total of \code{burnin + draws} iterations.

 Three thinning parameters are available, which all are $1$ if not specified otherwise. The first one, \code{thinpara}, specifies the denominator of the fraction of parameter draws (i.e., draws of $\allpara$) that are stored. E.g., if \code{thinpara} equals $10$, every $10$th draw is kept. The default parameter thinning value of $1$ means that all draws are saved. The second thinning parameter, \code{thinlatent}, acts in the same way for the latent variables $\hsvm$. The third thinning parameter, \code{thintime}, refers to thinning with respect to the time dimension of the latent volatility.
In the case that \code{thintime} is greater than 1, not all elements of $\hsvm$ are stored, e.g., for \code{thintime} equaling $10$, only the draws of $\hsv_1, \hsv_{11}, \hsv_{21}, \dots$ (and $\hsv_0$) are kept.

Another configuration argument is \code{quiet}, which defaults to \code{FALSE}. If set to \code{TRUE}, all output during sampling (progress bar, status messages) is omitted. The arguments \code{startpara} and \code{startlatent} are optional starting values for the parameter vector $\allpara$ and the latent variables $\hsvm$, respectively. All other configuration parameters are summarized in the argument \code{expert}, because it is not very likely that the end-user needs to mess with the defaults.\footnote{Examples of configurations that can be changed with the \code{expert}-argument include the specification of the (baseline) parameterization ({\it centered} or {\it noncentered}) and the possibility to turn off interweaving. Moreover, some algorithmic details such as the number of blocks used for the parameter updates or the possibility of using a random walk Metropolis-Hastings proposal (instead of the default independence proposal) can be found here.} Please refer to the package documentation in combination with \cite{kas-fru:anc} for details.

Any further arguments (\code{\dots}) will be forwarded to \code{updatesummary}, controlling the type of summary statistics that are calculated for the posterior draws.

\subsection{Running the sampler}

At the heart of the package \pkg{stochvol} lies the function \code{svsample}, which serves as an \proglang{R}-wrapper for the actual sampler coded in \proglang{C}. Exemplary usage of this function is given in the code snipped below, along with the default output.

<<eval=FALSE>>=
res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
                priorsigma = .1)
@

<<results=hide, echo=FALSE>>=
res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
@

\begin{CodeOutput}
Calling GIS_C MCMC sampler with 11000 iter. Series length T = 3139.

  0% [+++++++++++++++++++++++++++++++++++++++++++++++++++] 100%

Timing (elapsed): 17.935 seconds.
613 iterations per second.

Converting results to coda objects... Done!
Summarizing posterior draws... Done!
\end{CodeOutput}

As can be seen, this function calls the main MCMC sampler and converts its output to \pkg{coda}-compatible objects. Moreover, some summary statistics for the posterior draws are calculated. The return value of \code{svsample} is an object of type \code{svdraws}, which is a named list with eight elements, holding (1) the parameter draws in \code{para}, (2) the latent log-volatilities in \code{latent}, (3) the initial latent log-volatility draw in \code{latent0}, (4) the data provided in \code{y}, (5) the sampling runtime in \code{runtime}, (6) the prior hyperparameters in \code{priors}, (7) the thinning values in \code{thinning}, and (8) summary statistics of these draws, alongside some common transformations thereof, in \code{summary}.

 \subsection{Assessing the output and displaying the results}
 \label{check}

 Following common practice, \code{print} and \code{summary} methods are available for \code{svdraws} objects. Each of these has two optional parameters, \code{showpara} and \code{showlatent}, specifying which output should be displayed. If \code{showpara} is \code{TRUE} (the default), values/summaries of the parameter draws are shown. If \code{showlatent} is \code{TRUE} (the default), values/summaries of the latent variable draws are shown. In the example below, the summary for the parameter draws only is displayed.

<<>>=
summary(res, showlatent = FALSE)
@

There are several plotting functions specifically designed for objects of class \code{svsample}, which will be described in the following paragraphs.
 \begin{enumerate}
  \item[(1)] \code{volplot}: Plots posterior quantiles of the latent volatilities in percent, i.e., $100\exp(h_t/2)$, over time. Apart from the mandatory \code{svsample}-object itself, this function takes several optional arguments. Only some will be mentioned here; for an exhaustive list please see the corresponding help document accessible through \code{?volplot} or \code{help(volplot)}. Selected optional arguments that are commonly used include \code{forecast} for $n$-step-ahead volatility prediction, \code{dates} for date-labels on the $x$-axis, alongside some graphical parameters. The code snipped below shows a typical example and Figure~\ref{fig3} displays its output.

<<usd3, fig=TRUE, include=FALSE, width=11, height=7>>=
volplot(res, forecast = 100, dates = exrates$date[-1])
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.9\textwidth]{article-usd3}
 \caption{Visualization of estimated contemporaneous volatilities of EUR-USD exchange rates, as provided by \code{volplot}. If not specified otherwise, posterior medians and 5\%/95\% quantiles are plotted. The dotted lines at the right side indicate predicted future volatilities.}
 \label{fig3}
\end{center}
\end{figure}

In case the user wants to display different posterior quantiles, the \code{updatesummary} function has to be called first. See the code below for an example and Figure~\ref{fig4} for the corresponding plot.

<<usd4, fig=TRUE, include=FALSE, width=11, height=7>>=
res <- updatesummary(res, quantiles = c(.01, .1, .5, .9, .99))
volplot(res, forecast = 100, dates = exrates$date[-1])
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.9\textwidth]{article-usd4}
 \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
 \label{fig4}
\end{center}
\end{figure}

 \item[(2)] \code{paratraceplot}: Displays trace plots for the parameters contained in $\allpara$. Note that the burn-in has already been discarded. Figure~\ref{fig5} shows an example.

<<usd5, fig=TRUE, include=FALSE, width=8, height=5>>=
par(mfrow = c(3, 1))
paratraceplot(res)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-usd5}
 \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
 %The underlying data set is again EUR-USD exchange rates.}
 \label{fig5}
\end{center}
 \end{figure}

\item[(3)] \code{paradensplot}: Displays a kernel density estimate for the parameters contained in $\allpara$. If the argument \code{showobs} is \code{TRUE} (which is the default), individual posterior draws are indicated through a \emph{rug}, i.e., short vertical lines at the $x$-axis. For quicker drawing of large posterior samples, this argument should be set to \code{FALSE}. If the argument \code{showprior} is \code{TRUE} (which is the default), the prior distribution is indicated through a dashed gray line. Figure~\ref{fig6} shows an example output for the EUR-USD exchange rates obtained from the \code{exrates} dataset.

<<usd6, fig=TRUE, include=FALSE, width=8, height=3.5>>=
par(mfrow = c(1, 3))
paradensplot(res)
@

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.9\textwidth]{article-usd6}
 \caption{Posterior density estimates (black solid lines) along with prior densities (dashed gray lines). Individual posterior draws are indicated by the underlying rug.}
 \label{fig6}
\end{center}
\end{figure}
\end{enumerate}

The generic \code{plot} method for \code{svdraws} objects combines all above plots into one plot. All arguments described above can be used. See \code{?plot.svsample} for an exhaustive summary of possible arguments and Figure~\ref{fig7} for an example.

<<usd7, fig=TRUE, include=FALSE, width=9, height=7>>=
plot(res)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=.9\textwidth]{article-usd7}
 \caption{Illustration of the default \code{plot} method for \code{svdraws}-objects. This visualization combines \code{volplot} (Figure~\ref{fig4}), \code{traceplot} (Figure~\ref{fig5}), and \code{paradensplot} (Figure~\ref{fig6}) into one single plot.}
 \label{fig7}
\end{center}
\end{figure}

For extracting standardized residuals, the \code{residuals}/\code{resid} method can be used on a given \code{svdraws} object. With the optional argument \code{type}, the type of summary statistic may be specified. Currently, \code{type} is allowed to be either \code{"mean"} or \code{"median"}, where the former corresponds to the default value. This method returns a real vector of class \code{svresid}, which contains the requested summary statistic of standardized residuals for each point in time. There is also a \code{plot} method available, providing the option of comparing the standardized residuals to the original data when given through the argument \code{origdata}. See the code below for an example and Figure~\ref{fig8} for the corresponding output.

<<usd8, fig=TRUE, include=FALSE, width=9, height=7>>=
myresid <- resid(res)
plot(myresid, ret)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=.9\textwidth]{article-usd8}
 \caption{Mean standardized residual plots for assessing the model fit, as provided by the corresponding \code{plot} method. The dashed lines in the bottom left panel indicate the 2.5\%/97.5\% quantiles of the standard normal distribution.}
 \label{fig8}
\end{center}
\end{figure}

\section[Using stochvol within other samplers]{Using \pkg{stochvol} within other samplers}
\label{other}
We demonstrate how the \pkg{stochvol} package can be used to incorporate stochastic volatility into any given MCMC sampler. For the sake of simplicity, we explain this procedure with the help of the Bayesian normal linear model with $T$ observations and $k=p-1$ predictors, given through
\begin{eqnarray}
 \label{reg}
\yb|\betab,\Sigmab \sim \Normal{\Xb\betab, \Sigmab}.
\end{eqnarray}
Here, $\yb$ denotes the $T\times1$ vector of responses, $\Xb$ is the $T \times p$ design matrix containing ones in the first column and the predictors in the others, and $\betab$ stands for the $p\times 1$ vector of regression coefficients. In the following sections, we will discuss two specifications of the $T\times T$ error covariance matrix $\Sigmab$.

\subsection{The Bayesian normal linear model with homoskedastic errors}
\label{homo}
The arguably simplest specification of the error covariance matrix in Equation~\ref{reg} is given by $\Sigmab\equiv\sigmaepsextrasuper2\Ib$, where $\Ib$ denotes the $T$-dimensional unit matrix. This specification is used in many applications and commonly referred to as the linear regression model with homoskedastic errors. To keep things simple, let model parameters $\betab$ and $\sigmaepsextrasuper2$ be equipped with the usual conjugate prior $p(\betab, \sigmaepsextrasuper2) = p(\betab|\sigmaepsextrasuper2)p(\sigmaepsextrasuper2)$, where
\begin{eqnarray*}
\betab|\sigmaepsextrasuper2 &\sim& \Normal{\bb0, \sigmaepsextrasuper2\Bb0},\\
 \sigmaepsextrasuper2&\sim& \Gammainv{c_0, C_0}.
\end{eqnarray*}

A commonly used Gibbs-sampler for drawing from the posterior distribution of this model is given by sampling in turn from the full conditional distributions $\betab|\yb,\sigmaepsextrasuper2 \sim \Normal{\bpostb, \Bpostb}$ with
\[
\bpostb = \left(\XtX + \Bb0^{-1} \right)^{-1}\left( \Xb^\top\yb + \Bb0^{-1}\bb0 \right), \qquad
\Bpostb = \sigmaepsextrasuper2\left(\XtX + \Bb0^{-1} \right)^{-1},
\]

and $\sigmaepsextrasuper2|\yb,\betab \sim \Gammainv{c_T, C_T}$ with
\[
c_T = c_0 + \frac{T}{2} + \frac{p}{2}, \qquad
C_T = C_0 + \frac{1}{2}
 \left(
 (\yb - \Xb\betab)^\top(\yb - \Xb\betab) +
 (\betab - \bb0)^\top\Bb0^{-1}(\betab - \bb0)
 \right).
\]

In \proglang{R}, this can straightforwardly be coded as follows:
\begin{itemize}
 \item Set seed to make results reproducible and simulate some data:
<<>>=
set.seed(123456)
T <- 1000
beta.true <- c(.1, .5)
sigma.true <- 0.01
X <- matrix(c(rep(1, T), rnorm(T, sd = sigma.true)), nrow = T)
y <- rnorm(T, X %*% beta.true, sigma.true)
@

\item Specify configuration parameters and prior values:
<<>>=
draws <- 5000
burnin <- 100
b0 <- matrix(c(0, 0), nrow = ncol(X))
B0inv <- diag(c(10^-10, 10^-10))
c0 <- 0.001
C0 <- 0.001
@
 \item Pre-calculate some values outside the main MCMC loop:
<<>>=
p <- ncol(X)
preCov <- solve(crossprod(X) + B0inv)
preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
preDf <- c0 + T/2 + p/2
@
\item Assign some storage space for holding the draws and set an initial value for $\sigmaepsextrasuper2$:
<<>>=
draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
colnames(draws1) <- c(paste("beta", 0:(p-1), sep='_'), "sigma")
sigma2draw <- 1
@
\item Run the main sampler: Iteratively draw from the conditional bivariate Gaussian distribution $\betab|\yb,\sigmaepsextrasuper2$, e.g., through the use of \pkg{mvtnorm} \citep{r:mvt}, and the conditional Inverse Gamma distribution $\sigmaepsextrasuper2|\yb,\betab$.
<<eval=FALSE>>=
for (i in -(burnin-1):draws) {
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))
 tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
 		crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
 sigma2draw <- 1/rgamma(1, preDf, rate = tmp)
 if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
}
@
<<echo=FALSE>>=
#if (file.exists('~/tmprary/article_draws1.RData')) {
# load('~/tmprary/article_draws1.RData')
#} else {
for (i in -(burnin-1):draws) {
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))
 tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
 		 crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
 sigma2draw <- 1/rgamma(1, preDf, rate = tmp)
 if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
}
# save(draws1, file = '~/tmprary/article_draws1.RData')
#}
@
 \item Finally, visualize the posterior draws:
<<homo, fig=TRUE, include=FALSE, width=7, height=6>>=
par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
plot(coda::mcmc(draws1))
@
\begin{figure}[!htp]
\begin{center}
\includegraphics[width=.85\textwidth]{article-homo}
\caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors. Underlying data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $\sigmaepsextrasuper{\text{true}} = 0.01$, $T=1000$.}
\label{homodraws}
\end{center}
\end{figure}
<<>>=
colMeans(draws1)
@
\end{itemize}

\subsection{The Bayesian normal linear model with SV errors}
\label{hetero}

Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1/2}, \dots, \e^{h_T/2}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$. Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package. To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution by a sampling step of $\allpara$ and $\hb$ through a call to \code{.svsample}. This ``dotted'' function is a minimal-overhead version of the regular \code{svsample}. It provides the full sampling functionality of the ``non-dotted'' version, but has slightly different default values, a simplified return value structure and does not perform costly input checks. Thus, it is optimized for repeated calls but needs to be used with proper care. Note that the current draws of the variables need to be passed to the function through \code{startpara} and \code{startlatent}.

\begin{itemize}
 \item Simulate some data:
<<>>=
mu.true <- log(sigma.true^2)
phi.true <- 0.97
volvol.true <- 0.3
simresid <- svsim(T, mu = mu.true, phi = phi.true, sigma = volvol.true)
y <- X %*% beta.true + simresid$y
@
 \item Specify configuration parameters and prior values:
<<>>=
draws <- 50000
burnin <- 1000
thinning <- 10
priormu <- c(-10, 2)
priorphi <- c(20, 1.5)
priorsigma <- 1
@
 \item Assign some storage space for holding the draws and set initial values:
<<>>=
draws2 <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + T + p)
colnames(draws2) <- c("mu", "phi", "sigma", paste("beta", 0:(p-1), sep='_'),
                      paste("h", 1:T, sep='_'))
betadraw <- c(0, 0)
svdraw <- list(para = c(mu = -10, phi = .9, sigma = .2), latent = rep(-10, T))
@
 \item Run the main sampler: Iteratively draw the latent volatilities (and AR-parameters) through conditioning on the regression parameters and calling \code{.svsample}, and the regression parameters through conditioning on the latent volatilities.
<<eval=FALSE>>=
for (i in -(burnin-1):draws) {
 ytilde <- y - X %*% betadraw
 svdraw <- .svsample(ytilde, startpara=para(svdraw),
                     startlatent=latent(svdraw), priormu=priormu,
                     priorphi=priorphi, priorsigma=priorsigma)
 normalizer <- as.numeric(exp(-latent(svdraw)/2))
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))
 if (i > 0 & i %% thinning == 0) { 
  draws2[i/thinning, 1:3] <- para(svdraw)
  draws2[i/thinning, 4:5] <- betadraw
  draws2[i/thinning, 6:(T+5)] <- latent(svdraw)
 }
}
@

<<echo=FALSE>>=
if (usePreCalcResults) {
 load('vignette_sampling_draws2.RData')
} else {
for (i in -(burnin-1):draws) {

 # draw latent volatilities and AR-parameters:
 ytilde <- y - X %*% betadraw
 svdraw <- .svsample(ytilde, startpara=para(svdraw),
                     startlatent=latent(svdraw), priormu=priormu,
                     priorphi=priorphi, priorsigma=priorsigma)

 # draw the betas:
 normalizer <- as.numeric(exp(-latent(svdraw)/2))
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))

 # store the results:
 if (i > 0 & i %% thinning == 0) { 
  draws2[i/thinning, 1:3] <- para(svdraw)
  draws2[i/thinning, 4:5] <- betadraw
  draws2[i/thinning, 6:(T+5)] <- latent(svdraw)
 }
}
 draws2selection <- draws2[,4:8]
 #save(draws2selection, file = 'vignette_sampling_draws2.RData')
}
@
\item Finally, visualize (some) posterior draws:
<<eval=FALSE>>=
plot(coda::mcmc(draws2[,4:7]))
@

<<hetero, fig=TRUE, include=FALSE, echo=FALSE, results=hide>>=
par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
plot(coda::mcmc(draws2selection[,1:4]))
@

\begin{figure}[!ht]
 \begin{center}
\includegraphics[width=.8\textwidth]{article-hetero}
\caption{Trace plots and kernel density estimates
%for draws from the marginal posterior distributions of the parameters
in the regression model with heteroskedastic errors. Data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $h_1^\text{true} = -8.28$, $h_2^\text{true} = -8.50$, $T=1000$.}
\label{heterodraws}
 \end{center}
\end{figure}
<<eval=FALSE>>=
colMeans(draws2[,4:8])
@

<<echo=FALSE>>=
colMeans(draws2selection)
@
\end{itemize}

\section{Real-world example}
In the following, we aim for a comparison of the performance of the Bayesian normal linear model with homoskedastic errors from Section~\ref{homo} with the Bayesian normal linear model with SV errors from Section~\ref{hetero} using the \code{exrates} data set introduced in Section~\ref{prep}.

\subsection{Model setup}

We again use the daily price of 1 EUR in USD from January 3, 2000, until April 4, 2012, denoted by $\pb = (p_1, p_2, \dots, p_T)^\top$. This time however, instead of using log-returns, we investigate the development of raw prices by regression: Let $\yb$ contain all raw observations except the very first one, and let $\Xb$ denote the design matrix containing ones in the first column and lagged raw prices in the second, i.e.
\[
\yb = 
\begin{pmatrix}
 p_2\\
 p_3\\
 \vdots\\
 p_T
\end{pmatrix}, \quad
\Xb =
\begin{pmatrix}
1&p_1\\
1&p_2\\
\vdots&\vdots\\
1&p_{T-1}
\end{pmatrix}.
\]
%
<<scatter, fig=TRUE, include=FALSE, results=hide, echo=FALSE, width=9>>=
x <- exrates$USD[-length(exrates$USD)]
y <- exrates$USD[-1]
X <- matrix(c(rep(1, length(x)), x), nrow = length(x))
par(mfrow=c(1,1), mar = c(2.9, 2.9, 2.7, .5), mgp = c(1.8,.6,0), tcl = -.4)
plot(x,y, xlab=expression(p[t]), ylab=expression(p[t+1]),
    main="Scatterplot of lagged daily raw prices of 1 EUR in USD",
    col="#00000022", pch=16, cex=1)
abline(0,1)
@
%
Clearly, we expect the posterior distribution of $\betab$ to spread around $(0,1)^\top$, which corresponds to a random walk. A scatterplot of $p_t$ against $p_{t+1}$, displayed in Figure~\ref{datascatter}, confirms this.
\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=.75\textwidth]{article-scatter}
\end{center}
\caption{Scatterplot of daily raw prices at time $t$ against daily raw prices at time $t+1$. The solid line indicates the identity function $f(x) = x$.}
\label{datascatter}
\end{figure}

\subsection{Posterior inference}

We run both samplers for 100\,000 iterations and discard the first 10\,000 draws as burn-in. Prior hyperparameters are chosen as follows: $\bb{0} = (0, 0)^\top$, $\Bb{0} = \diag{10^{10}, 10^{10}}$, $c_0 = C_0 = 0.001$, $b_\mu = -10$, $B_\mu=2$, $a_0=20$, $b_0 = 1.5$, $B_\sigmapar=1$. However, due to the length of the dataset (and its obvious heteroskedasticity), the exact prior specification is not very influential. The two samplers yield slightly different posteriors for $\betab$, visualized in Figure~\ref{betapost}, where both the marginal posterior densities $p(\beta_0|\yb)$ and $p(\beta_1|\yb)$ as well as a scatterplot of draws from the joint posterior $\betab|\yb$ are displayed.

<<results=hide, echo=FALSE>>=
if (usePreCalcResults) {
 load('vignette_sampling_realworld.RData')
} else {
set.seed(34567890)

#configuration parameters
draws <- 100000
thinning <- 100
burnin <- 10000
priormu <- c(-10, 1)
priorphi <- c(20, 1.5)
priorsigma <- .1
p <- 2

n <- length(x)

## design matrix:
X <- matrix(c(rep(1, n), x), nrow=n)

## prior for beta (or beta|sigma in homoskedastic regression):
b0 <- matrix(c(0, 0), nrow=ncol(X))
B0 <- diag(c(10^10, 10^10))

## prior for sigma^2 (only used in homoskedastic regression)
c0 <- 0.001
C0 <- 0.001

B0inv <- solve(B0)

## initialize some space to hold results:
realres <- vector('list', 2)

realres[[1]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
colnames(realres[[1]]) <- c("mu", "phi", "sigma", paste("beta", 0:(p-1), sep='_'),
			   paste("h", 1:n, sep='_'))

## some indicators:
paras <- 1:3
betas <- 3+(1:p)
latents <- 3+p+(1:n)

## starting values:
betadraw <- rep(.1, p)
svdraw <- list(para = c(mu = -10, phi = .8, sigma = .2),
	       latent = rep(-10, n))

## sampler:
tim <- system.time(
for (i in -(burnin-1):draws) {
 if (i%%1000 == 0) cat("Iteration", i, "done.\n")
 # draw latent volatilities and AR-parameters:
 ytilde <- y - X %*% betadraw
 svdraw <- .svsample(ytilde, startpara=para(svdraw),
		      startlatent=latent(svdraw), priormu=priormu,
		      priorphi=priorphi, priorsigma=priorsigma)

 # draw the betas:
 normalizer <- as.numeric(exp(-latent(svdraw)/2))
 Xnew <- X * normalizer
 ynew <- y * normalizer
 Sigma <- solve(crossprod(Xnew) + B0inv)
 mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))

 # store the results:
 if (i > 0 & i %% thinning == 0) { 
  realres[[1]][i/thinning, paras] <- para(svdraw)
  realres[[1]][i/thinning, latents] <- latent(svdraw)
  realres[[1]][i/thinning, betas] <- betadraw
 }
}
)[["elapsed"]]

## standard sampler:
## pre-calculate some values:
preCov <- solve(crossprod(X) + B0inv)
preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
preDf <- c0 + n/2 + p/2

## allocate space:
realres[[2]] <- matrix(as.numeric(NA), nrow=floor(draws/thinning), ncol=p+1)
colnames(realres[[2]]) <- c(paste("beta", 0:(p-1), sep='_'), "sigma")

## starting values:
betadraw <- rep(0, p)
sigma2draw <- var(y)

## sampler:
tim2 <- system.time(
for (i in -(burnin-1):draws) {
 if (i%%1000 == 0) cat("Iteration", i, "done.\n")
 # draw beta:
 betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))

 # draw sigma^2:
 tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
		crossprod((betadraw-b0), B0inv) %*% (betadraw-b0))
 sigma2draw <- 1/rgamma(1, preDf, rate=tmp)

 # store the results:
 if (i > 0 & i %% thinning == 0) { 
  realres[[2]][i/thinning, 1:p] <- betadraw
  realres[[2]][i/thinning, p+1] <- sqrt(sigma2draw)
 }
}
)[["elapsed"]]

#Standardized residuals for model checking:
#homoskedastic
predmeans2 <- tcrossprod(X, realres[[2]][,1:2])
standresidmeans2 <- rowMeans((y - predmeans2)/realres[[2]][,3])

#heteroskedastic
predmeans <- tcrossprod(X, realres[[1]][,4:5])
standresidmeans <- rowMeans((y - predmeans)/exp(t(realres[[1]][,6:ncol(realres[[1]])])/2))

realresselection <- vector('list', 2)
realresselection[[1]] <- realres[[1]][,c('beta_0', 'beta_1')]
realresselection[[2]] <- realres[[2]][,c('beta_0', 'beta_1')]
#save(realresselection, standresidmeans, standresidmeans2, file = 'vignette_sampling_realworld.RData')
}
@

<<betapost, echo=FALSE, result=hide, fig=TRUE, include=FALSE, width=9>>=
smootherfactor <- 1.8
par(mar = c(2.9, 2.9, 2.7, .5), mgp = c(1.8,.6,0), tcl = -.4)
layout(matrix(c(1,2,3,3), byrow=TRUE, nrow=2))
plot(density(realresselection[[1]][,"beta_0"], bw="SJ", adjust=smootherfactor),
    main=bquote(paste("p(",beta[0],"|y)", sep="")),
    xlab="", ylab="", xlim=c(-.0025, .0045))
lines(density(realresselection[[2]][,"beta_0"], bw="SJ", adjust=smootherfactor), lty=2, col=2)
#points(ols$coefficients[1],0, col=2)
#lines(confint(ols)[1,], rep(0,2), col=2)
legend("topleft", c("SV", "homosked."), col=1:2, lty=1:2)

plot(density(realresselection[[1]][,"beta_1"], bw="SJ", adjust=smootherfactor),
    main=bquote(paste("p(",beta[1],"|y)", sep="")),
    xlab="", ylab="", xlim=c(0.9965, 1.0022))
lines(density(realresselection[[2]][,"beta_1"], bw="SJ", adjust=smootherfactor), lty=2, col=2)
#points(ols$coefficients[2],0, col=2)
#lines(confint(ols)[2,], rep(0,2), col=2)
legend("topright", c("SV", "homosked."), col=1:2, lty=1:2)

plotorder <- sample.int(2*nrow(realresselection[[1]]))
cols <- rep(c("#00000055", "#ff000055"), each=nrow(realresselection[[1]]))[plotorder]
pchs <- rep(1:2, each=nrow(realresselection[[1]]))[plotorder]
beta0 <- c(realresselection[[1]][,"beta_0"], realresselection[[2]][,"beta_0"])[plotorder]
beta1 <- c(realresselection[[1]][,"beta_1"], realresselection[[2]][,"beta_1"])[plotorder]

plot(beta0, beta1, col=cols, pch=pchs,
    xlab=bquote(paste("p(",beta[0],"|y)")),
    ylab=bquote(paste("p(",beta[1],"|y)")),
    main="Scatterplot of posterior draws")
legend("topright", c("SV", "homosked."), col=1:2, pch=1:2)
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{article-betapost}
 \caption{Visualization of the posterior distributions $\betab|\yb$ for the model with SV residuals and the model with homoskedastic errors. Top panels: kernel density estimate of the univariate posterior marginal distributions. Bottom panel: bivariate scatterplot of posterior draws.}
  \label{betapost}
\end{center}
 \end{figure}

To assess the model fit, mean standardized resdiduals for both samplers are depicted in Figure~\ref{qqplot}. It stands out that the model with homoskedastic errors shows deficiencies in terms of heavy correlation amongst the residuals. This can clearly be seen in the top left panel, where mean standardized residuals are plotted against time. The bottom left panel shows the same plot for the model with SV errors, where this effect practically vanishes. Moreover, in the model with homoskedastic errors, the normality assumption about the unconditional error distribution is clearly violated. This can be seen by inspecting the quantile-quantile plot in the top right panel, where observed residuals show much heavier tails than one would expect from a normal distribution. On the contrary, standardized residuals obtained from the model with SV errors align almost perfectly.

<<qqplot, echo=FALSE, result=hide, fig=TRUE, include=FALSE, width=9, height=7>>=
par(mfrow=c(2,2), mar = c(3.1, 3.3, 2.0, .5), mgp = c(1.7,.5,0),
    tcl = -.4)
plot(standresidmeans2, main="Residual scatterplot (homoskedastic errors)",
     xlab="Time", ylab="Standardized residuals")
qqplot(qnorm(ppoints(length(standresidmeans2)), mean=0, sd=1),
       standresidmeans2, main="Residual Q-Q plot (homoskedastic errors)",
       xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
abline(0,1)
plot(standresidmeans, main="Residual scatterplot (SV errors)",
     xlab="Time", ylab="Standardized residuals")
qqplot(qnorm(ppoints(length(standresidmeans)), mean=0, sd=1),
       standresidmeans, main="Residual Q-Q plot (SV errors)",
       xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
abline(0,1)
@

\begin{figure}[!ht]
 \begin{center}
  \includegraphics[width=.9\textwidth]{article-qqplot}
 \end{center}
 \caption{Visualization of mean standardized residuals. Top left panel shows a scatterplot against time for the model with homoskedastic errors, bottom left panel shows this plot for the model with SV errors. Quantile-Quantile plots of empirical quantiles against expected quantiles from a $\Normal{0,1}$-distribution are displayed on the panels on the right-hand side.}
 \label{qqplot}
\end{figure}

\subsection{Predictive performance and model fit}
Within a Bayesian framework, a natural way of assessing the predictive performance of a given model is through its \emph{predictive density} (sometimes also referred to as \emph{posterior predictive distribution}). Collecting all \emph{unobservables}, i.e., parameters and possible latent variables, in a single vector $\unobs$, it is given through
\begin{eqnarray}
 \label{preddens}
p(\y_{t+1}|\yb^o_{[1:t]}) = \int\limits_{\bm K} \!
 p(y_{t+1}|\yb^o_{[1:t]}, \unobs) \times
 p(\unobs|\yb^o_{[1:t]}) 
 \, \dif \unobs.
\end{eqnarray}
Note that for the model with homoskedastic errors, $\unobs=(\betab,\sigmaeps)^\top$, while for the model with SV errors, $\unobs=(\betab, \allpara, \hb)^\top$. By using the superscript $o$ in $\yb^o_{[1:t]}$, we follow \cite{gew-ami:com} and denote \emph{ex post} realizations (observations) for the set of points in time $\{1,2,\dots ,t\}$ of the \emph{ex ante} random values $\yb_{[1:t]} = (\y_1, \y_2, \dots,\y_t)^\top$. Integration is carried out over ${\bm K}$, which simply stands for the space of all possible values for $\unobs$. Equation~\ref{preddens} can be viewed as the integral of the likelihood function over the joint posterior distribution of the unobservables $\unobs$. Thus, it can be interpreted as the predictive density for a future value $\y_{t+1}$ after integrating out the uncertainty about $\unobs$, conditional on the history $\yb^o_{[1:t]}$.

In the SV errors case, Equation~\ref{preddens} is a $(T+p+3)$-dimensional integral which cannot be solved analytically. Nevertheless, it may be evaluated at an arbitrary point $x$ through Monte Carlo integration:
\begin{equation}
 \label{preddensMC}
 p(x|\yb^o_{[1:t]}) \approx \frac{1}{M}\sum_{m=1}^M p(x|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]}),
\end{equation}
%
where $\unobs^{(m)}_{[1:t]}$ stands for the $m$\textsuperscript{th} draw from the respective posterior distribution up to time $t$. If Equation~\ref{preddensMC} is evaluated at $x=y_{t+1}^o$, we refer to it as the (one-step-ahead) \emph{predicitve likelihood} (at time $t+1$) denoted $\PL_{t+1}$.
Moreover, draws from (\ref{preddens}) can be obtained by simulating values $\y_{t+1}^{(m)}$ from the distribution given through the density $p(\y_{t+1}|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]})$, the summands of Equation~\ref{preddensMC}.

For model at hand, the predictive density and likelihood can thus be computed through the following

\begin{alg}[Predictive density and likelihood evaluation at time $t+1$]
 \label{predlikalg}
 \
 \begin{enumerate}
  \item Reduce the data set to a training set $\yb^o_{[1:t]} = (\y_1^o,\dots,y_t^o)^\top$.
  \item Run the posterior sampler using data from the training set only to obtain $M$ posterior draws $\unobs^{(m)}_{[1:t]}$.
  \item[(3.)] Needed for the SV model only: Simulate $M$ values from the conditional distribution $\hsv_{t+1,[1:t]}|\yb^o_{[1:t]},\unobs_{[1:t]}$ by drawing $\hsv_{t+1,[1:t]}^{(m)}$ from a normal distribution with mean $\mupar_{[1:t]}^{(m)} + \phipar_{[1:t]}^{(m)}(\hsv^{(m)}_{t,[1:t]}-\mupar_{[1:t]}^{(m)})$ and standard deviation $\sigmaparextrasub{[1:t]}^{(m)}$ for $m=1,\dots,M$.
  \item[4a.] To obtain $\PL_{t+1}$, average over $M$ densities of normal distributions with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model) or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model), each evaluated at $y^o_{t+1}$.
  \item[4b.] To obtain $M$ draws from the predictive distribution, simulate from a normal distribution with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model) or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model) for $m=1,\dots,M$.
 \end{enumerate}
\end{alg}

To avoid strong dependence on the prior, the first 1000 days are used as training set only and the evaluation of the predictive distribution starts at $t=1001$, corresponding to December 4, 2003. The results for both models are displayed in Figure~\ref{predlik1}. In the top panel, the observed series along with the 98\% one-day-ahead predictive intervals are displayed. The bottom panel shows the log one-day-ahead predictive likelihood.

Figure~\ref{predlik2} displays a zoomed-in version, depicting only the time span from January 2008 until August 2009. Note that while at the beginning of 2008 both credible intervals are very similar, there is a substantial difference one year later, where SV intervals become around twice as large compared to the corresponding homoskedastic analogs. According to the values of the log predictive likelihoods, SV errors can handle the inflated volatility during that time substantially better. Throughout 2009, the width of the intervals as well as the predictive likelihoods consolidate again.

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{predlik1}
\end{center}
\caption{Top panel: Observed series (green) and symmetrical 98\% one-day-ahead predictive intervals for the model with homoskedastic errors (red) and the model with SV errors (black). Bottom panel: Log one-day-ahead predictive likelihoods for both models.}
\label{predlik1}
\end{figure}

\begin{figure}[!htp]
\begin{center}
 \includegraphics[width=\textwidth]{predlik2}
\end{center}
\caption{Zoomed version of Figure~\ref{predlik1}, showing only the period from January 2008 until August 2009. This time span is chosen to includes the beginning of the financial crisis. During that phase, predictive performance of the model with homoskedastic errors deteriorates substantially, while SV errors can capture the inflated volatility much better.}
\label{predlik2}
\end{figure}

It is worth pointing out that log predictive likelihoods also carry an intrinsic connection to the log \emph{marginal likelihood}, defined through
\begin{eqnarray}
 \label{ML}
\log \ML = \log p(\yb^o) = \log \int\limits_{\bm K} \!
 p(\yb^o| \unobs) \times
 p(\unobs) 
 \, \dif \unobs.
\end{eqnarray}
This real number corresponds to the logarithm of the normalizing constant in the denominator of Bayes' law and is often
used for evaluating model evidence. It can straightforwardly be decomposed into the sum of the one-step-ahead logarithms
of the predictive likelihoods:
\[
\log \ML = \log p(\yb^o) = \log \prod_{t=1}^T p(\y_t^o|\yb_{[1:t-1]}^o) = \sum_{t=1}^T \log \PL_{t}.
\]
Thus, Algorithm \ref{predlikalg} provides a conceptually simple way of computing the marginal likelihood. However, these computations are quite costly in terms of CPU time, as they require an individual model fit for each of the $T$ points in time. On the other hand, due to the embarrassingly parallel nature of the task and because of today's comparably easy access to parallel computing environments, this burden becomes easily manageable. E.g., the computations for the analysis in this paper have been conducted in less than one hour, using 25 IBM dx360M3 nodes within a cluster of workstations. Implementation in \proglang{R} was achieved through the packages \pkg{parallel} \citep{r:r} and \pkg{snow} \citep{r:sno}.

Cumulative sums of $\log \PL_t$ also allow for model comparison through cumulative log predictive Bayes factors. Letting $\PL_t(A)$ denote the predictive likelihood of model $A$ at time $t$, and $\PL_t(B)$ the corresponding value of model $B$, the cumulative log predictive Bayes factor at time $u$ (and starting point $S$) in favor of model $A$ over model $B$ is simply given through
\begin{equation}
 \label{clpbf}
\log \left[ \frac{p_A(\yb^o_{[S+1:u]}|\yb^o_{[1:S]})}{p_B(\yb^o_{[S+1:u]}|\yb^o_{[1:S]})} \right] =  \sum_{t=S+1}^u \log \left[ \frac{\PL_t(A)}{\PL_t(B)} \right] =  \sum_{t=S+1}^u [\log \PL_t(A) - \log \PL_t(B)].
\end{equation}

If the cumulative log predictive Bayes factor is positive at a given point in time, we have evidence in favor of model $A$, and vice versa. In this context, information up to time $S$ is regarded as prior information, while out-of-sample predictive evaluation starts at time $S+1$. Note that the usual log Bayes factor is a special case of Equation~\ref{clpbf} for $S=0$ and $u=T$.

The top panel of Figure~\ref{predlik3} gives an overview of the model performance through visual inspection of observed residuals against their predicted distributions. For the purpose of this paper, residuals are simply defined as the deviation from the median of the predicted distribution. It stands out that predictive quantiles arising from the model with SV errors exhibit much more flexibility to ``adapt'' to the ``current state of the world'', while the simpler homoskedastic model barely exhibits this feature. While there is little difference in predicted residuals until the beginning of 2007, the model with SV residuals performs substantially better during the pre-crisis era (less volatility) and during the actual crisis (more volatility). This picture is confirmed through the cumulative log predictive Bayes factors displayed in the bottom panel. Note that the last point plotted equals around $223$, providing overwhelming overall evidence in favor of the model with SV errors, clearly rejecting the assumption of homoskedastic error terms.

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{predlik3}
\end{center}
\caption{Top panel: Observed residuals with respect to the median of the one-day-ahead predictive distribution along with 1\% and 99\% quantiles of the respective predictive distributions. It can clearly be seen that the variance of the predictive distribution in the SV model adjusts to heteroskedasticity, while the model with homoskedastic errors is much more restrictive. Bottom panel: Cumulative log predictive Bayes factors in favor of the model with SV residuals. Values greater than zero mean that the model with SV residuals performs better out of sample up to this point in time.}
\label{predlik3}
\end{figure}

\section{Conclusion}
Aim of this paper was to introduce to the functionality of the \proglang{R} package \pkg{stochvol}, which provides a fully Bayesian simulation-based approach for inference in stochastic volatility models. The typical workflow occurring when using \pkg{stochvol} was presented by analyzing exchange rate data in the package's \code{exrates} data set. Furthermore, it was shown how the package can be used as a ``plug-in'' tool for other MCMC samplers. This was illustrated by estimating a Bayesian linear model with SV errors.

In the real-world example, raw exchange rates from EUR to USD were analyzed. Out-of-sample analysis through cumulative predictive Bayes factors clearly showed that already for a simple linear autoregressive model for the raw exchange rates, SV residuals substantially improve prediction performance, especially in turbulent times.

\bibliography{mybib}
\end{document}

### 2014-02-06 src/sampler.cpp

@@ -2,6 +2,12 @@
 
 #include "sampler.h"
 
+// registering "update" to be available for other packages
+
+RcppExport void R_init_stochvol(DllInfo *dll) {
+ R_RegisterCCallable("stochvol", "update", (DL_FUNC) &update);
+}
+
 using namespace Rcpp; // avoid to type Rcpp:: every time
 RNGScope scope;       // just in case no seed has been set at R level
 
@@ -78,8 +84,7 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
  for (int i = 0; i < T; i++) h[i] = startvol[i];  // need to make a manual copy (!)
  if (!centered_baseline) h = (h-mu[0])*sqrt(sigma2inv[0]);
  
- NumericVector h_alter(T);  // contains h1 to hT in alternative para
- double h0, h0_alter;
+ double h0;
  
  IntegerVector r(T);  // mixture indicators
  
@@ -89,35 +94,71 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
 
  NumericMatrix mixprob(10, T);  // mixture probabilities
  NumericVector data = log(y*y + offset);  // commonly used transformation
- NumericVector omega_diag(T);  // contains diagonal elements of precision matrix
- double omega_offdiag;  // contains off-diag element of precision matrix (const)
- NumericVector chol_offdiag(T-1), chol_diag(T);  // Cholesky-factor of Omega
- NumericVector covector(T);  // holds covector (see McCausland et al. 2011)
- NumericVector htmp(T);  // intermediate vector for sampling h
- NumericVector newvals(3);  // holds mu, phi, sigma in every iteration
- newvals[0] = mu[0];
- newvals[1] = phi[0];
- newvals[2] = 1/sqrt(sigma2inv[0]);
-
- double sigmainvtmp, phi2tmp;
+ NumericVector curpara(3);  // holds mu, phi, sigma in every iteration
+ curpara[0] = mu[0];
+ curpara[1] = phi[0];
+ curpara[2] = 1/sqrt(sigma2inv[0]);
 
  // initializes the progress bar
  // "show" holds the number of iterations per progress sign
  int show;
  if (verbose) show = progressbar_init(N);
  
[…]

### 2014-02-06 src/sampler.h

@@ -4,9 +4,14 @@
 // Main sampling steps and helper functions
 
 #include <Rcpp.h>
+#define COMPILING_STOCHVOL
+#include <update.h>  // this header declares update
+                     // (which is also available to other packages)
 #include "auxmix.h"
 #include "progutils.h"
 #include "densities.h"
+
+// "update" is declared in /inst/include/update.h
 
 // Main sampler (as called from R):
 RcppExport SEXP sampler(const SEXP, const SEXP, const SEXP,
@@ -17,7 +22,7 @@ RcppExport SEXP sampler(const SEXP, const SEXP, const SEXP,
 
 // Step (b): sample mu, phi, sigma - __CENTERED__ version:
 Rcpp::NumericVector regressionCentered(
-       double h0, Rcpp::NumericVector h,
+       double h0, const Rcpp::NumericVector &h,
        double mu, double phi, double sigma,
        double C0, double cT, double Bsigma,
        double a0, double b0,
@@ -27,8 +32,9 @@ Rcpp::NumericVector regressionCentered(
 
 // Step (b): sample mu, phi, sigma - __NONCENTERED__ version:
 Rcpp::NumericVector regressionNoncentered(
-       Rcpp::NumericVector data,
-       double h0, Rcpp::NumericVector h, Rcpp::IntegerVector r,
+       const Rcpp::NumericVector &data,
+       double h0, const Rcpp::NumericVector &h,
+       const Rcpp::IntegerVector &r,
        double mu, double phi, double sigma,
        double Bsigma, double a0, double b0,
        double bmu, double Bmu,

### 2014-02-07 src/progutils.cpp
version 0.8-1

@@ -39,7 +39,7 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
 
 // sets up the progress bar
 int progressbar_init(int N) {
- int show, tmp;
+ int show;
  REprintf("\n      ");
  if (N >= 2500) {
   for (int i = 0; i < 50+1; i++) REprintf(" ");

### 2014-02-07 src/sampler.cpp

@@ -65,7 +65,7 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
  double C0 = 1.5*Bsigma;
  
  // pre-calculation of a posterior parameter
- double cT; 
+ double cT = -10000; 
  if (Gammaprior) {  
   if (MHsteps == 2 || MHsteps == 3) cT = T/2.0; // we want IG(-.5,0) as proposal
   else if (MHsteps == 1) cT = (T-1)/2.0; // we want IG(-.5,0) as proposal
@@ -265,9 +265,12 @@ Rcpp::NumericVector regressionCentered(
        bool Gammaprior, bool truncnormal, double MHcontrol, int MHsteps) {
  
  int T = h.length();
- double R, z, CT, sigma2_prop, sigma_prop, sum1, sum2, sum3, sum4, tmp1,
-       	tmp2, BT11, BT12, BT22, bT1, bT2, chol11, chol12, chol22, phi_prop,
+ double z, CT, sum1, sum2, sum3, sum4, tmp1,
+       	BT11, BT12, BT22, bT1, bT2, chol11, chol12, chol22, phi_prop,
        	gamma_prop, tmpR, tmpR2, logR;
+ double R = -10000;
+ double sigma2_prop = -10000;
+ double sigma_prop = -10000;
  Rcpp::NumericVector innov(2);
  Rcpp::NumericVector quant(2);
 

### 2014-07-01 inst/doc/article.Rnw
version 0.8-4

@@
 This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm.
-For details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.
+For more details about the algorithm and empirical results concerning sampling efficiency, see \cite{kas-fru:anc}.
 
@@
 
-Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1/2}, \dots, \e^{h_T/2}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$.
+Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1}, \dots, \e^{h_T}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$.
 Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package.
@@
 
+For an example of using \pkg{stochvol} for univariate SV updates within a factor SV framework, see \cite{kas-etal:ana}.
+
 \section{Conclusion}

### 2014-07-01 R/wrapper.R

@@ -243,7 +243,7 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
  res <- .Call("sampler", y, draws, burnin, priormu[1], priormu[2]^2,
 	      priorphi[1], priorphi[2], priorsigma, thinlatent, thintime,
 	      startpara, startlatent, quiet, 3L, 2L, 10^8, 10^12,
-	      -1, TRUE, FALSE, 0, PACKAGE = "stochvol")
+	      -1, TRUE, FALSE, 0, FALSE, PACKAGE = "stochvol")
 
  res$para <- t(res$para[-1,,drop=FALSE])
  rownames(res$para) <- names(res$para) <- c("mu", "phi", "sigma")

### 2014-10-16 R/wrapper.R
version 0.9-0

@@ -190,7 +190,7 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
  }
 
  if (!quiet) {
-  cat(paste("\nCalling ", parameterization, " MCMC sampler with ", draws+burnin, " iter. Series length T = ", length(y), ".\n",sep=""), file=stderr())
+  cat(paste("\nCalling ", parameterization, " MCMC sampler with ", draws+burnin, " iter. Series length is ", length(y), ".\n",sep=""), file=stderr())
   flush.console()
  }
 
@@ -238,7 +238,7 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
 
 # This function does not check input nor converts the result to coda objects
 
-.svsample <- function(y, draws = 1, burnin = 0, priormu = c(0, 100), priorphi = c(5, 1.5), priorsigma = 1, thinpara = 1, thinlatent = 1, thintime = 1, quiet = TRUE, startpara, startlatent) {
+svsample2 <- function(y, draws = 1, burnin = 0, priormu = c(0, 100), priorphi = c(5, 1.5), priorsigma = 1, thinpara = 1, thinlatent = 1, thintime = 1, quiet = TRUE, startpara, startlatent) {
 
  res <- .Call("sampler", y, draws, burnin, priormu[1], priormu[2]^2,
 	      priorphi[1], priorphi[2], priorsigma, thinlatent, thintime,
@@ -249,3 +249,7 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
  rownames(res$para) <- names(res$para) <- c("mu", "phi", "sigma")
  res
 }
+
+# Note: .svsample is deprecated and will be faded out, use svsample2 instead
+
+.svsample <- svsample2

### 2014-10-16 inst/doc/article.R

@@ -1,11 +1,14 @@
 ### R code from vignette source 'article.Rnw'
 
 ###################################################
-### code chunk number 1: article.Rnw:1-4
+### code chunk number 1: article.Rnw:1-7
 ###################################################
 # This should be set to FALSE in case you want to redo the sampling as well.
 # Default is TRUE to save CRAN some time.
 usePreCalcResults <- TRUE
+
+# Default here is FALSE because the object created is rather large
+usePreCalcResultsExtra <- FALSE 
 
 
 ###################################################
@@ -20,35 +23,40 @@ options(prompt = 'R> ', continue = '+  ')
 set.seed(123)
 library("stochvol")
 data("exrates")
-ret <- logret(exrates$USD, demean=TRUE)
-par(mfrow=c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
+ret <- logret(exrates$USD, demean = TRUE)
+par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
 plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
-plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log-returns")
+plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log returns")
 
 
 ###################################################
 ### code chunk number 4: usd2
 ###################################################
 sim <- svsim(500, mu = -9, phi = 0.99, sigma = 0.1)
-par(mfrow=c(2, 1))
+par(mfrow = c(2, 1))
 plot(sim)
 
 
 ###################################################
-### code chunk number 5: article.Rnw:219-221 (eval = FALSE)
+### code chunk number 5: article.Rnw:227-229 (eval = FALSE)
 ###################################################
 ## res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
 ##                 priorsigma = .1)
 
 
 ###################################################
-### code chunk number 6: article.Rnw:224-225
+### code chunk number 6: article.Rnw:232-238
 ###################################################
-res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
+if (usePreCalcResultsExtra) {
+ load('vignette_sampling_draws1.RData')
+} else {
+ res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
[…]

### 2014-10-16 src/sampler.cpp

@@ -21,7 +21,9 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
   const SEXP truncnormal_in, const SEXP offset_in,
   const SEXP dontupdatemu_in) {
 
- RNGScope scope;       // just in case no seed has been set at R level
+ //RNGScope scope;       // just in case no seed has been set at R level
+ GetRNGstate(); // "by hand" because RNGScope isn't save if return
+                // variables are declared afterwards
 
  // convert SEXP into Rcpp-structures (no copy at this point)
  NumericVector y(y_in), startvol(startvol_in);
@@ -131,21 +133,21 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
   update(data, &curpara(0), &h(0), h0, &mixprob(0,0), &r(0), centered_baseline, C0, cT,
          Bsigma, a0, b0, bmu, Bmu, B011inv, B022inv, Gammaprior,
 	 truncnormal, MHcontrol, MHsteps, parameterization, dontupdatemu);
-  /* 
-  Rprintf("mixind  after:  %i\n", r(2));
+   
+ /* Rprintf("mixind  after:  %i\n", r(2));
    Rprintf("mixprob after:  %f\n", mixprob(5,0));
    Rprintf("curpara after:  %f\n", curpara(1));
    Rprintf("h0      after: %f\n", h0);
    Rprintf("h       after:  %f\n", h(2));
    Rprintf("data    after:  %f\n\n", data(5));
-*/
+   */
 
- 
   // storage:
   if (!((i+1) % thin)) if (i >= burnin) {  // this means we should store h
    store_h(&h(0), &hstore(0, (i-burnin)/thin), timethin, hstorelength,
             h0, &h0store((i-burnin)/thin), curpara, centered_baseline);
   }
+//  Rprintf("hSTORE  after:  %f\n", hstore(2, (i-burnin)/thin));
   mu[i+1] = curpara[0];
   phi[i+1] = curpara[1];
   sigma2inv[i+1] = 1/(curpara[2]*curpara[2]);
@@ -636,5 +638,6 @@ Rcpp::NumericVector regressionNoncentered(
  if (Rcpp::as<double>(Rcpp::runif(1)) < expR) phi = phi_prop;
 
  Rcpp::NumericVector ret = Rcpp::NumericVector::create(mu, phi, fabs(sigma));
+ PutRNGstate();
  return ret;
 }

### 2014-12-27 inst/doc/article.Rnw
version 0.9-1

@@
 usePreCalcResults <- TRUE
+
+# Default here is FALSE because the object created is rather large
+usePreCalcResultsExtra <- FALSE 
 @
@@
 \documentclass[article, nojss]{jss}
+%\documentclass[article]{jss}
 
@@
 \usepackage{bm}
+\usepackage{textcomp}
+\usepackage{tablefootnote}
 
@@
 \author{Gregor Kastner\\WU Vienna University of Economics and Business}
-%\title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol} \\ (DRAFT VERSION)}
 \title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
@@
 \Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting
-%\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol (DRAFT VERSION)} %% without formatting
 
@@
 \Abstract{
-Heteroskedasticity in financial and economic datasets is a commonly observed feature, and the need to model this feature properly has for a long time been of great interest to researchers and practitioners alike.
-The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modelling by means of the \emph{stochastic volatility} framework.
-Being a state-space formulation, it describes contemporaneous volatilities as latent random variables as opposed to deterministic values.
-The software described in this paper utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables, which can then be used for predicting future volatilities.
+The \proglang{R} package \pkg{stochvol} provides a fully Bayesian implementation of heteroskedasticity modeling within the framework of stochastic volatility.
+It utilizes Markov chain Monte Carlo (MCMC) samplers to conduct inference by obtaining draws from the posterior distribution of parameters and latent variables which can then be used for predicting future volatilities.
 The package can straightforwardly be employed as a stand-alone tool; moreover, it allows for easy incorporation into other MCMC samplers.
-Main focus of the paper is to show the functionality of \pkg{stochvol}, nevertheless it also provides a brief mathematical description of the model, an overview of the sampling schemes used, and an in-depth example using exchange rate data.
+The main focus of this paper is to show the functionality of \pkg{stochvol}.
+In addition, it provides a brief mathematical description of the model, an overview of the sampling schemes used, and several illustrative examples using exchange rate data.
 }
@@
 \section{Introduction}
-When analyzing (financial) returns, focus is often laid on estimating and predicting potentially time varying volatilities.
-This interest has a long history, dating at least back to \cite{mar:por}, who investigated portfolio construction with optimal expected return-variance trade-off.
+Returns -- in particular financial returns -- are commonly analyzed by estimating and predicting potentially time-varying volatilities.
+This focus has a long history, dating back at least to \cite{mar:por} who investigates portfolio construction with optimal expected return-variance trade-off.
 In his article, he proposes rolling-window-type estimates for the instantaneous volatilities, but already then recognizes the potential for ``better methods, which take into account more information''.
 
-One way of doing so is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models.
-After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and were applied to a vast amount of real-world problems.
+One approach is to model the evolution of volatility deterministically, i.e., through the (G)ARCH class of models.
+After the groundbreaking papers of \cite{eng:aut} and \cite{bol:gen}, these models have been generalized in numerous ways and applied to a vast amount of real-world problems.
 As an alternative, \cite{tay:fin} proposes in his seminal work to model the volatility probabilistically, i.e., through a state-space model where the logarithm of the squared volatilities -- the latent states -- follow an autoregressive process of order one.
 Over time, this specification became known as the \emph{stochastic volatility (SV)} model.
-Even though several papers \citep[e.g.][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} find early evidence in favor of using SV, these models have found comparably little use in applied work.
-Reviewing this discrepancy, \cite{bos:rel} states:
-\begin{quote}
- While there are literally thousands of applications of GARCH, for SV, this number is far lower.
+Even though several papers \citep[e.g.,][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} provide early evidence in favor of using SV, these models have found comparably little use in applied work.
+This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard packages implementing these methods.
+
+%While there are literally thousands of applications of GARCH, for SV, this number is far lower.
 Two reasons for this relative lack of applied work using SV are apparent.
@@
 For SV, there are few variants of the model, but a full series of estimation methods.
-\end{quote}
-In \cite{kas-fru:anc}, the latter issue has been thoroughly investigated, and an efficient MCMC estimation scheme
+
+ In \cite{kas-fru:anc}, the former issue is thoroughly investigated and an efficient MCMC estimation scheme
 %using techniques such as sampling AWOL \citep[``all without a loop'', adapted from][]{rue:fas, mcc-etal:sim} and ASIS \citep[``ancillarity-sufficiency interweaving strategy'', adapted from][]{yu-men:cen}, an
-has been proposed.
-The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} were crafted to cope with the first problem: the apparent lack of standard packages for efficiently estimating SV models.
+is proposed.
+The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use packages for efficiently estimating SV models.
 
@@
 
-We briefly introduce the model and specify the notation used in the remainder of the paper.
-Furthermore, a quick overview over Bayesian parameter estimation via Markov chain Monte Carlo (MCMC) methods is given.
+We begin by briefly introducing the model and specifying the notation used in the remainder of the paper.
+Furthermore, an overview of Bayesian parameter estimation via Markov chain Monte Carlo (MCMC) methods is given.
 
@@
 
-Let $\yb = (y_1, y_2, \dots, y_T)^\top$ be a vector of returns with mean zero.
+Let $\yb = (y_1, y_2, \dots, y_n)^\top$ be a vector of returns with mean zero.
 The intrinsic feature of the SV model is that each observation $y_t$ is assumed to have its ``own'' contemporaneous variance $\e^{\hsv_t}$, thus relaxing the usual assumption of homoskedasticity.
-In order to make estimation of such a model feasible (note that there are just as many data points as there are variances!), this variance is not allowed to vary unrestrictedly with time.
-Rather, its logarithm is assumed to follow a parametric stochastic process, more specifically: an autoregressive process of order one.
-Note that feature is fundamentally different to GARCH-type models, where the time-varying volatility is assumed to follow a deterministic rather than stochastic evolution.
+In order to make the estimation of such a model feasible,
+%(note that there are just as many data points as there are variances!)
+this variance is not allowed to vary unrestrictedly with time.
+Rather, its logarithm is assumed to follow an autoregressive process of order one.
+Note that this feature is fundamentally different to GARCH-type models where the time-varying volatility is assumed to follow a deterministic instead of a stochastic evolution.
 
@@
  \label{c2}
- \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapar^2},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
+ \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapartwo},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
  \label{c3}
-\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapar^2/(1-\phi^2)},
+\hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapartwo/(1-\phi^2)},
 \end{eqnarray}
 
-where $\Normal{\mu, \sigmapar^2}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapar^2$.
-We will refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} $\mupar$, the \emph{persistence} $\phipar$, and the \emph{volatility} $\sigmapar$ of log-variance.
-The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{T})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely: the log-variance process).
+where $\Normal{\mu, \sigmapartwo}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapartwo$.
+We refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} of log-variance $\mupar$, the \emph{persistence} of log-variance $\phipar$, and the \emph{volatility} of log-variance $\sigmapar$.
+The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process).
 Note that the initial state $\hsv_0$ is distributed according to the stationary distribution of the autoregressive process of order one.
@@
 The level $\mupar \in \mathbb{R}$ is equipped with the usual normal prior $\mupar \sim \Normal{b_{\mupar}, B_{\mupar}}$.
-In practical applications, this prior is usually chosen to be rather uninformative, e.g., through setting $b_\mupar = 0$ and $B_\mupar \geq 100$ for daily log-returns.
+In practical applications, this prior is usually chosen to be rather uninformative, e.g., through setting $b_\mupar = 0$ and $B_\mupar \geq 100$ for daily log returns.
 Our experience with empirical data is that the exact choice is usually not very influential; see also Section~\ref{priorconfig}.
@@
 where $a_0$ and $b_0$ are positive hyperparameters and $\Betafun{x,y} = \int_0^1t^{x-1}(1-t)^{y-1}\,dt$ denotes the beta function.
-Clearly, the support of this distribution is the unit interval $(-1, 1)$; thus, stationarity of the autoregressive volatility process is guaranteed.
+Clearly, the support of this distribution is the interval $(-1, 1)$; thus, stationarity of the autoregressive volatility process is guaranteed.
 Its expected value and variance are given through the expressions
@@
 For a fixed ratio $a_0:b_0$, the prior variance decreases with larger values of $a_0$ and $b_0$.
-The uniform distribution on the unit interval arises as a special case when $a_0=b_0=1$.
-For financial datasets with not too many observations (i.e., $T \lesssim 1000$), the choice of the hyperparameters $a_0$ and $b_0$ can be quite influential on the shape of the posterior distribution of $\phipar$.
-In fact, note that in the case when the underlying data-generating process is (near-)homoskedastic, the volatility of log-variance $\sigmapar$ will be (very close to) zero and thus the likelihood will contain little to no information about $\phipar$.
-Consequently, the posterior distribution of $\phipar$ will be (almost) equal to its prior, no matter how many data points are observed.
-For some discussion about this issue, see e.g., \cite{kim-etal:sto}, who choose $a_0=20$ and $b_0=1.5$, implying a prior mean of $0.86$ with a prior standard deviation of $0.11$ and thus very little mass for nonpositive values of $\phi$.
-
-For the volatility of log-variance $\sigmapar \in \mathbb{R}^+$, we choose $\sigmapar^2 \sim B_{\sigmapar}\times \chi^2_1=\Gammad{1/2,1/2B_{\sigmapar}}$, which is motivated by \cite{fru-wag:sto}, who equivalently stipulate the prior for $\pm \sqrt { \sigmapar^2}$ to follow a centered normal distribution, i.e., $\pm \sqrt { \sigmapar^2} \sim \Normal{0,B_{\sigmapar}}$.
-As opposed to the usual Inverse-Gamma prior, this prior is not conjugate in the usual sampling scheme, however, does not bound $\sigmapar$ away from zero a priori.
-The choice of the hyperparameter $B_{\sigmapar}$ turns out to be of minor influence in empirical applications, as long as it is not picked to be too small.
+The uniform distribution on $(-1,1)$ arises as a special case when $a_0=b_0=1$.
+For financial datasets with not too many observations (i.e., $n \lesssim 1000$), the choice of the hyperparameters $a_0$ and $b_0$ can be quite influential on the shape of the posterior distribution of $\phipar$.
+In fact, when the underlying data-generating process is (near-)homoskedastic, the volatility of log-variance $\sigmapar$ is (very close to) zero and thus the likelihood contains little to no information about $\phipar$.
+Consequently, the posterior distribution of $\phipar$ is (almost) equal to its prior, no matter how many data points are observed.
+For some discussion about this issue, see, e.g., \cite{kim-etal:sto} who choose $a_0=20$ and $b_0=1.5$, implying a prior mean of $0.86$ with a prior standard deviation of $0.11$ and thus very little mass for nonpositive values of $\phi$.
+
+For the volatility of log-variance $\sigmapar \in \mathbb{R}^+$, we choose $\sigmapartwo \sim B_{\sigmapar}\times \chi^2_1=\Gammad{1/2,1/2B_{\sigmapar}}$.
+This choice is motivated by \cite{fru-wag:sto} who equivalently stipulate the prior for $\pm \sqrt { \sigmapartwo}$ to follow a centered normal distribution, i.e., $\pm \sqrt { \sigmapartwo} \sim \Normal{0,B_{\sigmapar}}$.
+As opposed to the more common Inverse-Gamma prior for $\sigmapartwo$, this prior is not conjugate in the usual sampling scheme.
+However, it does not bound $\sigmapartwo$ away from zero a priori.
+The choice of the hyperparameter $B_{\sigmapar}$ turns out to be of minor influence in empirical applications as long as it is not set too small.
 
@@
 
-An MCMC algorithm such as the one implemented in the package \pkg{stochvol} will provide its user with draws from the posterior distribution of the desired random variables: in our case the latent log-variances $\hsvm$ and the parameter vector $\allpara$.
+An MCMC algorithm such as the one implemented in the package \pkg{stochvol} provides its user with draws from the posterior distribution of the desired random variables; in our case with the latent log-variances $\hsvm$ and the parameter vector $\allpara$.
 Because these draws are usually dependent, Bayesian inference via MCMC may require careful design of the algorithm and attentive investigation of the draws obtained.
@@
 
-In order to avoid the cost of code interpretation within each MCMC iteration, the core computations of the sampler are implemented in \proglang{C}, interfaced to \proglang{R} via the \pkg{Rcpp} package \citep{edd-fra:rcp}, where the convenience functions and the user-interface are implemented.
+In order to avoid the cost of code interpretation within each MCMC iteration, the core computations are implemented in \proglang{C}.
+Their output is interfaced to \proglang{R} via the \pkg{Rcpp} package \citep{edd-fra:rcp}; there, the convenience functions and the user interface are implemented.
 This combination allows to make use of the well-established and widely accepted ease-of-use of \proglang{R} and its underlying functional programming paradigm.
 Moreover, existing frameworks for analyzing MCMC output such as \pkg{coda} \citep{plu-etal:cod} as well as high-level visualization tools can easily be used.
-Last but not least, users with a basic knowledge of \proglang{R} can use the package in a familiar surrounding with a very small entry cost.
-Nevertheless, despite all these convenience features, the package profits from highly optimized machine code generated by a compiler at package build time, thus providing acceptable runtime even for larger datasets.
-
-A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of an ``ancillarity-sufficiency interweaving strategy'' (ASIS), which has been brought forward in the general context of state-space models by \cite{yu-men:cen}.
-ASIS exploits the fact that for certain parameter constellations, sampling efficiency improves substantially when considering a non-centered version of a state-space model.
-This fact is commonly known as a reparameterization issue with an entire body of literature attached to it; for an early reference see e.g., \cite{hil-smi:par}.
-In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility $\sigmapar$ of log-variance from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$.
+Last but not least, users with a basic knowledge of \proglang{R} can use the package with a very low entry cost.
+Nevertheless, despite all these convenience features, the package profits from a highly optimized machine code generated by a compiler at package build time, thus providing acceptable runtime even for larger datasets.
+
+A novel and crucial feature of the algorithm implemented in \pkg{stochvol} is the usage of a variant of the ``ancillarity-sufficiency interweaving strategy'' (ASIS) which has been brought forward in the general context of state-space models by \cite{yu-men:cen}.
+ASIS exploits the fact that for certain parameter constellations sampling efficiency improves substantially when considering a non-centered version of a state-space model.
+This is commonly referred to as a reparameterization issue and dealt with extensively in scholarly literature; for an early reference see, e.g., \cite{hil-smi:par}.
+In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$.
 However, in the case of the SV model, it turns out that no single superior parameterization exists.
 Rather, for some underlying processes, the standard parameterization yields superior results, while for other processes non-centered versions are better.
-To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered, and once in a noncentered parameterization.
+To overcome this issue, the parameter vector $\allpara$ is sampled twice: once in the centered and once in a noncentered parameterization.
 This method of ``combining best of different worlds'' allows for efficient inference regardless of the underlying process with one algorithm.
@@
 
-The usual stand-alone approach to fitting SV models with \pkg{stochvol} follows the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results.
-All these steps will be described in more detail below, along with a worked example.
-For a stepwise incorporation of SV effects into other MCMC samplers, please see Section \ref{other}.
+The usual stand-alone approach to fitting SV models with \pkg{stochvol} exhibits the following workflow: (1) Prepare the data, (2) specify the prior distributions and configuration parameters, (3) run the sampler, (4) assess the output and display the results.
+All these steps are described in more detail below, along with a worked example.
+For a stepwise incorporation of SV effects into other MCMC samplers, see Section \ref{other}.
 
@@
 
-The core sampling function \code{svsample} expects its input data \code{y} to be a numeric vector of returns without any missing values (\code{NA}s), and will throw an error if provided with anything else.
-In case that \code{y} contains zeros, a warning will be issued and a small offset constant of size \code{sd(y)/10000} will be added to the squared returns before doing the auxiliary mixture sampling \citep[cf.][]{omo-etal:sto}.
-A common and recommended way of avoiding zero returns is to de-mean the returns beforehand.
-Below is an example how to prepare data, illustrated with the \code{exrates} data set\footnote{The data set, which as been obtained from the European Central Bank's Statistical Data Warehouse, contains the daily bilateral prices of one Euro in 23 currencies from January 3, 2000, until April 4, 2012.
-Conversions to New Turkish Lira and Fourth Romanian Leu have been incorporated.
-See \code{?exrates} for more information.} which is included in the package.
-A visualization of one of these time series is displayed in Figure~\ref{fig1}.
+The core sampling function \code{svsample} expects its input data \code{y} to be a numeric vector of returns without any missing values (\code{NA}s) and throws an error if provided with anything else.
+In the case that \code{y} contains zeros, a warning is issued and a small offset constant of size \code{sd(y)/10000} is added to the squared returns before doing the auxiliary mixture sampling \citep[cf.][]{omo-etal:sto}.
+However, we generally recommend to avoid zero returns altogether, e.g., by demeaning them beforehand.
+
+Below is an illustration of how to prepare data by using the sample dataset \code{exrates}\footnote{The dataset -- obtained from the European Central Bank's Statistical Data Warehouse -- contains the daily bilateral prices of one euro in 23 currencies from January 3, 2000 until April 4, 2012.
+Conversions to New Turkish lira and Fourth Romanian leu have been incorporated.
+See \code{?exrates} for more information.} included in the package.
+Figure~\ref{fig1} provides a visualization of a time series from this dataset.
 
@@
 data("exrates")
-ret <- logret(exrates$USD, demean=TRUE)
-par(mfrow=c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
+ret <- logret(exrates$USD, demean = TRUE)
+par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
 plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
-plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log-returns")
+plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log returns")
 @
@@
 
-Additionally to real-world data, \pkg{stochvol} has also a built-in data generator \code{svsim}.
-This function simply produces realizations of an SV process and returns an object of class \code{svsim}, which has its own \code{print}, \code{summary}, and \code{plot} methods.
-Exemplary code using \code{svsim} is given below, and the particular instance of this simulated series is displayed in Figure~\ref{fig2}.
+Additionally to real-world data, \pkg{stochvol} also has a built-in data generator \code{svsim}.
+This function simply produces realizations of an SV process and returns an object of class \code{svsim} which has its own \code{print}, \code{summary}, and \code{plot} methods.
+Exemplary code using \code{svsim} is given below and the particular instance of this simulated series is displayed in Figure~\ref{fig2}.
 
@@
 sim <- svsim(500, mu = -9, phi = 0.99, sigma = 0.1)
-par(mfrow=c(2, 1))
+par(mfrow = c(2, 1))
 plot(sim)
@@
  After preparing the data vector \code{y}, the user needs to specify the prior hyperparameters for the parameter vector $\allpara = (\mupar, \phipar, \sigmapar)^\top$ -- see also Section~\ref{priors} -- and some configuration parameters.
-The appropriate values are passed to the main sampling function \code{svsample} as arguments, which are described below.
+The appropriate values are passed to the main sampling function \code{svsample} as arguments which are described below.
 
@@
 A common strategy is to choose a vague prior here, e.g., \code{c(0, 100)}, because the likelihood usually carries enough information about this parameter.
-If one prefers to use (slightly) informative priors, e.g., to avoid outlier draws of $\mupar$, care must be taken whether actual log-returns or \emph{percentage} log-returns are analyzed.
-Assuming daily data, the former commonly have an unconditional variance of $0.0001$ or less and thus the level on the log-scale $\mupar$ lies around $\log(0.0001) \approx -9$, while the latter have the $100^2$-fold unconditional variance (around $1$), which implies a level of $\log(1) = 0$.
+If one prefers to use (slightly) informative priors, e.g., to avoid outlier draws of $\mupar$, one must pay attention to whether \emph{log returns} or \emph{percentage log returns} are analyzed.
+Assuming daily data, log returns commonly have an unconditional variance of $0.0001$ or less and thus the level on the log scale $\mupar$ lies around $\log(0.0001) \approx -9$.
+Percentage log returns, on the other hand, have the $100^2$-fold unconditional variance (around $1$) which implies a level of $\log(1) = 0$.
 Choices in the literature include \code{c(0, 10)} \citep{jac-etal:bayJE}, \code{c(0, 5)} \citep{yu:on}, \code{c(0, sqrt(10))} \citep{kim-etal:sto, mey-yu:bug} or \code{c(0, 1)} \citep{omo-etal:sto}.
-Note that most of these choices are quite informative and clearly designed for \emph{percentage} log-returns!
+Note that most of these choices are quite informative and clearly designed for percentage log returns.
 
@@
 It is again a vector of length $2$, containing $a_0$ and $b_0$ specified in Equation~\ref{beta_transformed}.
-As elaborated in Section~\ref{priors}, these values can possibly be quite influential, thus we advise to put some thought into choosing them and study the effect of different choices carefully.
+As elaborated in Section~\ref{priors}, these values can possibly be quite influential, thus we advise to choose them carefully and study the effects of different choices.
 The default is currently given through \code{c(5, 1.5)}, implying a prior mean of $0.54$ and a prior standard deviation of $0.31$.
@@
 As discussed in Section~\ref{priors}, the exact choice of this value is usually not very influential in typical applications.
-In general, it should not be picked too small unless there is a very good reason, e.g., explicit prior knowledge, to do so.
+In general, it should not be set too small unless there is a very good reason, e.g., explicit prior knowledge, to do so.
 
@@
 It is the amount of MCMC iterations that are run but discarded to ensure convergence to the stationary distribution of the chain.
-The current default value for this parameter is $1000$, which has proved to suffice in most situations.
-Nevertheless, the user is encouraged to check convergence carefully, see Section \ref{check} for more details.
+The current default value for this parameter is $1000$ which has turned out to suffice in most situations.
+Nevertheless, the user is encouraged to check convergence carefully; see Section \ref{check} for more details.
 The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$.
-Consequently, the sampler will be run for a total of \code{burnin + draws} iterations.
-
- Three thinning parameters are available, which all are $1$ if not specified otherwise.
-The first one, \code{thinpara}, specifies the denominator of the fraction of parameter draws (i.e., draws of $\allpara$) that are stored.
+Consequently, the sampler is run for a total of \code{burnin + draws} iterations.
+
+ Three thinning parameters are available which all are $1$ if not specified otherwise.
+The first one, \code{thinpara}, is the denominator in the fraction of parameter draws (i.e., draws of $\allpara$) that are stored.
 E.g., if \code{thinpara} equals $10$, every $10$th draw is kept.
@@
 
-Another configuration argument is \code{quiet}, which defaults to \code{FALSE}.
+Another configuration argument is \code{quiet} which defaults to \code{FALSE}.
 If set to \code{TRUE}, all output during sampling (progress bar, status messages) is omitted.
 The arguments \code{startpara} and \code{startlatent} are optional starting values for the parameter vector $\allpara$ and the latent variables $\hsvm$, respectively.
-All other configuration parameters are summarized in the argument \code{expert}, because it is not very likely that the end-user needs to mess with the defaults.\footnote{Examples of configurations that can be changed with the \code{expert}-argument include the specification of the (baseline) parameterization ({\it centered} or {\it noncentered}) and the possibility to turn off interweaving.
-Moreover, some algorithmic details such as the number of blocks used for the parameter updates or the possibility of using a random walk Metropolis-Hastings proposal (instead of the default independence proposal) can be found here.} Please refer to the package documentation in combination with \cite{kas-fru:anc} for details.
-
-Any further arguments (\code{\dots}) will be forwarded to \code{updatesummary}, controlling the type of summary statistics that are calculated for the posterior draws.
+All other configuration parameters are summarized in the argument \code{expert}, because it is not very likely that the end-user needs to mess with the defaults.\footnote{Examples of configurations that can be changed with the \code{expert}-argument include the specification of the (baseline) parameterization (either \emph{centered} or \emph{noncentered}) and the possibility to turn off interweaving.
+Moreover, some algorithmic details such as the number of blocks used for the parameter updates or the possibility of using a random walk Metropolis-Hastings proposal (instead of the default independence proposal) can be found here.} Please refer to the package documentation and \cite{kas-fru:anc} for details.
+
+Any further arguments (\code{\dots}) are forwarded to \code{updatesummary}, controlling the type of summary statistics that are calculated for the posterior draws.
 
@@
 
-At the heart of the package \pkg{stochvol} lies the function \code{svsample}, which serves as an \proglang{R}-wrapper for the actual sampler coded in \proglang{C}.
+At the heart of the package \pkg{stochvol} lies the function \code{svsample} which serves as an \proglang{R}-wrapper for the actual sampler coded in \proglang{C}.
 Exemplary usage of this function is given in the code snipped below, along with the default output.
@@
 <<results=hide, echo=FALSE>>=
+if (usePreCalcResultsExtra) {
+ load('vignette_sampling_draws1.RData')
+} else {
  res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
+ #save(res, file = 'vignette_sampling_draws1.RData')
+}
 @
@@
 \begin{CodeOutput}
-Calling GIS_C MCMC sampler with 11000 iter. Series length T = 3139.
+Calling GIS_C MCMC sampler with 11000 iter. Series length is 3139.
 
@@
 As can be seen, this function calls the main MCMC sampler and converts its output to \pkg{coda}-compatible objects.
+The latter is done mainly for reasons of compatibility and in order to have straightforward access to the convergence diagnostics checks implemented there.
 Moreover, some summary statistics for the posterior draws are calculated.
-The return value of \code{svsample} is an object of type \code{svdraws}, which is a named list with eight elements, holding (1) the parameter draws in \code{para}, (2) the latent log-volatilities in \code{latent}, (3) the initial latent log-volatility draw in \code{latent0}, (4) the data provided in \code{y}, (5) the sampling runtime in \code{runtime}, (6) the prior hyperparameters in \code{priors}, (7) the thinning values in \code{thinning}, and (8) summary statistics of these draws, alongside some common transformations thereof, in \code{summary}.
+The return value of \code{svsample} is an object of type \code{svdraws} which is a named list with eight elements, holding (1) the parameter draws in \code{para}, (2) the latent log-volatilities in \code{latent}, (3) the initial latent log-volatility draw in \code{latent0}, (4) the data provided in \code{y}, (5) the sampling runtime in \code{runtime}, (6) the prior hyperparameters in \code{priors}, (7) the thinning values in \code{thinning}, and (8) summary statistics of these draws, alongside some common transformations thereof, in \code{summary}.
 
@@
 
-There are several plotting functions specifically designed for objects of class \code{svsample}, which will be described in the following paragraphs.
+There are several plotting functions specifically designed for objects of class \code{svsample} which are described in the following paragraphs.
  \begin{enumerate}
@@
 Apart from the mandatory \code{svsample}-object itself, this function takes several optional arguments.
-Only some will be mentioned here; for an exhaustive list please see the corresponding help document accessible through \code{?volplot} or \code{help(volplot)}.
-Selected optional arguments that are commonly used include \code{forecast} for $n$-step-ahead volatility prediction, \code{dates} for date-labels on the $x$-axis, alongside some graphical parameters.
+Only some are mentioned here; for an exhaustive list please see the corresponding help document accessible through \code{?volplot} or \code{help(volplot)}.
+Selected optional arguments that are commonly used include \code{forecast} for $n$-step-ahead volatility prediction, \code{dates} for labels on the $x$-axis, alongside some graphical parameters.
 The code snipped below shows a typical example and Figure~\ref{fig3} displays its output.
@@
 
-\begin{figure}[!htp]
-\begin{center}
- \includegraphics[width=.9\textwidth]{article-usd3}
+\begin{figure}[!ht]
+\begin{center}
+ \includegraphics[width=\textwidth]{article-usd3}
  \caption{Visualization of estimated contemporaneous volatilities of EUR-USD exchange rates, as provided by \code{volplot}.
 If not specified otherwise, posterior medians and 5\%/95\% quantiles are plotted.
-The dotted lines at the right side indicate predicted future volatilities.}
+The dotted lines on the right indicate predicted future volatilities.}
  \label{fig3}
@@
 \begin{center}
- \includegraphics[width=.9\textwidth]{article-usd4}
+ \includegraphics[width=.96\textwidth]{article-usd4}
  \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
@@
 
-\begin{figure}[!ht]
-\begin{center}
- \includegraphics[width=\textwidth]{article-usd5}
+\begin{figure}[!htp]
+\begin{center}
+ \includegraphics[width=.96\textwidth]{article-usd5}
  \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
- %The underlying data set is again EUR-USD exchange rates.}
+ %The underlying dataset is again EUR-USD exchange rates.}
  \label{fig5}
@@
 \item[(3)] \code{paradensplot}: Displays a kernel density estimate for the parameters contained in $\allpara$.
-If the argument \code{showobs} is \code{TRUE} (which is the default), individual posterior draws are indicated through a \emph{rug}, i.e., short vertical lines at the $x$-axis.
+If the argument \code{showobs} is \code{TRUE} (which is the default), individual posterior draws are indicated through a rug, i.e., short vertical lines along the $x$-axis.
 For quicker drawing of large posterior samples, this argument should be set to \code{FALSE}.
 If the argument \code{showprior} is \code{TRUE} (which is the default), the prior distribution is indicated through a dashed gray line.
-Figure~\ref{fig6} shows an example output for the EUR-USD exchange rates obtained from the \code{exrates} dataset.
+Figure~\ref{fig6} shows a sample output for the EUR-USD exchange rates obtained from the \code{exrates} dataset.
 
@@
 \begin{center}
- \includegraphics[width=.9\textwidth]{article-usd6}
+ \includegraphics[width=\textwidth]{article-usd6}
  \caption{Posterior density estimates (black solid lines) along with prior densities (dashed gray lines). Individual posterior draws are indicated by the underlying rug.}
@@
 <<usd7, fig=TRUE, include=FALSE, width=9, height=7>>=
-plot(res)
+plot(res, showobs = FALSE)
 @
@@
 \begin{center}
- \includegraphics[width=.9\textwidth]{article-usd7}
+ \includegraphics[width=\textwidth]{article-usd7}
  \caption{Illustration of the default \code{plot} method for \code{svdraws}-objects.
@@
 Currently, \code{type} is allowed to be either \code{"mean"} or \code{"median"}, where the former corresponds to the default value.
-This method returns a real vector of class \code{svresid}, which contains the requested summary statistic of standardized residuals for each point in time.
+This method returns a real vector of class \code{svresid} which contains the requested summary statistic of standardized residuals for each point in time.
 There is also a \code{plot} method available, providing the option of comparing the standardized residuals to the original data when given through the argument \code{origdata}.
@@
 \begin{center}
- \includegraphics[width=.9\textwidth]{article-usd8}
+ \includegraphics[width=\textwidth]{article-usd8}
  \caption{Mean standardized residual plots for assessing the model fit, as provided by the corresponding \code{plot} method.
@@
 We demonstrate how the \pkg{stochvol} package can be used to incorporate stochastic volatility into any given MCMC sampler.
-For the sake of simplicity, we explain this procedure with the help of the Bayesian normal linear model with $T$ observations and $k=p-1$ predictors, given through
+This is particularly easy when the sampler itself is coded in \proglang{R} or \proglang{C}/\proglang{C++} and applies, e.g., to many of the specialized procedures listed in the CRAN task view about Bayesian inference, available at \url{http://cran.r-project.org/web/views/Bayesian.html}.
+For the sake of simplicity, we explain the procedure using a ``hand-coded'' Gibbs-sampler for the Bayesian normal linear model with $n$ observations and $k=p-1$ predictors, given through
 \begin{eqnarray}
@@
 \end{eqnarray}
-Here, $\yb$ denotes the $T\times1$ vector of responses, $\Xb$ is the $T \times p$ design matrix containing ones in the first column and the predictors in the others, and $\betab$ stands for the $p\times 1$ vector of regression coefficients.
-In the following sections, we will discuss two specifications of the $T\times T$ error covariance matrix $\Sigmab$.
+Here, $\yb$ denotes the $n\times1$ vector of responses, $\Xb$ is the $n \times p$ design matrix containing ones in the first column and the predictors in the others, and $\betab = (\beta_0, \beta_1, \dots, \beta_{p-1})^\top$ stands for the $p\times 1$ vector of regression coefficients.
+In the following sections, we discuss two specifications of the $n\times n$ error covariance matrix $\Sigmab$.
 
@@
 \label{homo}
-The arguably simplest specification of the error covariance matrix in Equation~\ref{reg} is given by $\Sigmab\equiv\sigmaepsextrasuper2\Ib$, where $\Ib$ denotes the $T$-dimensional unit matrix.
+The arguably simplest specification of the error covariance matrix in Equation~\ref{reg} is given by $\Sigmab\equiv\sigmaepsextrasuper2\Ib$, where $\Ib$ denotes the $n$-dimensional unit matrix.
 This specification is used in many applications and commonly referred to as the linear regression model with homoskedastic errors.
@@
 
-A commonly used Gibbs-sampler for drawing from the posterior distribution of this model is given by sampling in turn from the full conditional distributions $\betab|\yb,\sigmaepsextrasuper2 \sim \Normal{\bpostb, \Bpostb}$ with
+Commonly, samples from the posterior distribution of this model are obtained through a Gibbs-algorithm, where draws are generated in turn from the full conditional distributions $\betab|\yb,\sigmaepsextrasuper2 \sim \Normal{\bpostb, \Bpostb}$ with
 \[
@@
 
-and $\sigmaepsextrasuper2|\yb,\betab \sim \Gammainv{c_T, C_T}$ with
+and $\sigmaepsextrasuper2|\yb,\betab \sim \Gammainv{c_n, C_n}$ with
 \[
-c_T = c_0 + \frac{T}{2} + \frac{p}{2}, \qquad
-C_T = C_0 + \frac{1}{2}
+c_n = c_0 + \frac{n}{2} + \frac{p}{2}, \qquad
+C_n = C_0 + \frac{1}{2}
  \left(
@@
 \begin{itemize}
- \item Set seed to make results reproducible and simulate some data:
+ \item Set the seed to make results reproducible and simulate some data from the underlying model:
 <<>>=
 set.seed(123456)
-T <- 1000
+n <- 1000
 beta.true <- c(.1, .5)
 sigma.true <- 0.01
-X <- matrix(c(rep(1, T), rnorm(T, sd = sigma.true)), nrow = T)
-y <- rnorm(T, X %*% beta.true, sigma.true)
-@
-
-\item Specify configuration parameters and prior values:
+X <- matrix(c(rep(1, n), rnorm(n, sd = sigma.true)), nrow = n)
+y <- rnorm(n, X %*% beta.true, sigma.true)
+@
+
+\item Specify the size of the burn-in and the number of draws thereafter; set the prior parameters:
 <<>>=
+burnin <- 100
 draws <- 5000
-burnin <- 100
 b0 <- matrix(c(0, 0), nrow = ncol(X))
@@
 preMean <- preCov %*% (crossprod(X, y) + B0inv %*% b0)
-preDf <- c0 + T/2 + p/2
+preDf <- c0 + n/2 + p/2
 @
@@
 <<eval=FALSE>>=
+for (i in -(burnin-1):draws) {
+ betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean,
+					 sigma2draw * preCov))
+ tmp <- C0 + .5 * (crossprod(y - X %*% betadraw) +
+ 		  crossprod((betadraw - b0), B0inv) %*%
+		   (betadraw - b0))
+ sigma2draw <- 1 / rgamma(1, preDf, rate = tmp)
+ if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
+}
+@
+<<echo=FALSE>>=
 for (i in -(burnin-1):draws) {
@@
 @
-<<echo=FALSE>>=
-#if (file.exists('~/tmprary/article_draws1.RData')) {
-# load('~/tmprary/article_draws1.RData')
-#} else {
-for (i in -(burnin-1):draws) {
- betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))
- tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
- 		 crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
- sigma2draw <- 1/rgamma(1, preDf, rate = tmp)
- if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
-}
-# save(draws1, file = '~/tmprary/article_draws1.RData')
-#}
-@
- \item Finally, visualize the posterior draws:
-<<homo, fig=TRUE, include=FALSE, width=7, height=6>>=
+\item Finally, summarize and visualize the posterior draws:
+<<>>=
+colMeans(draws1)
+@
+
+<<homo, echo=FALSE, fig=TRUE, include=FALSE, width=7.5, height=6>>=
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
@@
 @
-\begin{figure}[!htp]
-\begin{center}
-\includegraphics[width=.85\textwidth]{article-homo}
+
+<<eval=FALSE>>=
+plot(coda::mcmc(draws1))
+@
+
+\begin{figure}[!ht]
+\begin{center}
+\includegraphics[width=.98\textwidth]{article-homo}
 \caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors.
-Underlying data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $\sigmaepsextrasuper{\text{true}} = 0.01$, $T=1000$.}
+Underlying data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $\sigmaepsextrasuper{\text{true}} = 0.01$, $n=1000$.}
 \label{homodraws}
@@
 \end{figure}
-<<>>=
-colMeans(draws1)
-@
+
 \end{itemize}
@@
 
-Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1}, \dots, \e^{h_T}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$.
+Instead of homoskedastic errors, we now specify the error covariance matrix in Equation~\ref{reg} to be $\Sigmab \equiv \diag{\e^{h_1}, \dots, \e^{h_n}}$, thus introducing nonlinear dependence between the observations due to the AR(1)-nature of $\hb$.
 Instead of cooking up an entire new sampler, we adapt the code from above utilizing the \pkg{stochvol} package.
-To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution by a sampling step of $\allpara$ and $\hb$ through a call to \code{.svsample}.
-This ``dotted'' function is a minimal-overhead version of the regular \code{svsample}.
-It provides the full sampling functionality of the ``non-dotted'' version, but has slightly different default values, a simplified return value structure and does not perform costly input checks.
-Thus, it is optimized for repeated calls but needs to be used with proper care.
-Note that the current draws of the variables need to be passed to the function through \code{startpara} and \code{startlatent}.
-
+To do so, we simply replace the sampling step of $\sigmaepsextrasuper{2}$ from an Inverse-Gamma distribution with a sampling step of $\allpara$ and $\hb$ through a call to \code{svsample2}.\footnote{In earlier version of the \pkg{stochvol} package, this function was called \code{.svsample}.} This function is a minimal-overhead version of the regular \code{svsample}.
+It provides the full sampling functionality of the original version but has slightly different default values, a simplified return value structure, and it does not perform costly input checks.
+Thus, it executes faster and is more suited for repeated calls.
+The drawback is that it needs to be used with proper care.\footnote{Erroneous or incompatible input values most likely result in run-time errors of compiled code, often implying a segmentation fault and the consecutive abnormal termination of \proglang{R}.
+This can render debugging tedious.} Note that the current draws of the variables need to be passed to the function through \code{startpara} and \code{startlatent}.
+
+Here is how it goes:
 \begin{itemize}
@@
 phi.true <- 0.97
-volvol.true <- 0.3
-simresid <- svsim(T, mu = mu.true, phi = phi.true, sigma = volvol.true)
+vv.true <- 0.3
+simresid <- svsim(n, mu = mu.true, phi = phi.true, sigma = vv.true)
 y <- X %*% beta.true + simresid$y
@@
 <<>>=
-draws2 <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + T + p)
-colnames(draws2) <- c("mu", "phi", "sigma", paste("beta", 0:(p-1), sep='_'),
-                      paste("h", 1:T, sep='_'))
+draws2 <- matrix(NA_real_, nrow = floor(draws / thinning),
+		 ncol = 3 + n + p)
+colnames(draws2) <- c("mu", "phi", "sigma",
+		      paste("beta", 0:(p-1), sep = '_'),
+                      paste("h", 1:n, sep='_'))
 betadraw <- c(0, 0)
-svdraw <- list(para = c(mu = -10, phi = .9, sigma = .2), latent = rep(-10, T))
-@
- \item Run the main sampler: Iteratively draw the latent volatilities (and AR-parameters) through conditioning on the regression parameters and calling \code{.svsample}, and the regression parameters through conditioning on the latent volatilities.
+svdraw <- list(para = c(mu = -10, phi = .9, sigma = .2),
+	       latent = rep(-10, n))
+@
+ \item Run the main sampler, i.e., iteratively draw
+  \begin{itemize}
+   \item the latent volatilities/parameters by conditioning on the regression parameters and calling \code{svsample2},
+   \item the regression parameters by conditioning on the latent volatilities and calling \code{rmvnorm}:
+  \end{itemize}
 <<eval=FALSE>>=
@@
  ytilde <- y - X %*% betadraw
- svdraw <- .svsample(ytilde, startpara=para(svdraw),
-                     startlatent=latent(svdraw), priormu=priormu,
-                     priorphi=priorphi, priorsigma=priorsigma)
+ svdraw <- svsample2(ytilde, startpara = para(svdraw),
+                     startlatent = latent(svdraw), priormu = priormu,
+                     priorphi = priorphi, priorsigma = priorsigma)
  normalizer <- as.numeric(exp(-latent(svdraw)/2))
@@
   draws2[i/thinning, 4:5] <- betadraw
-  draws2[i/thinning, 6:(T+5)] <- latent(svdraw)
+  draws2[i/thinning, 6:(n+5)] <- latent(svdraw)
  }
@@
  ytilde <- y - X %*% betadraw
- svdraw <- .svsample(ytilde, startpara=para(svdraw),
-                     startlatent=latent(svdraw), priormu=priormu,
-                     priorphi=priorphi, priorsigma=priorsigma)
+ svdraw <- svsample2(ytilde, startpara = para(svdraw),
+                     startlatent = latent(svdraw), priormu = priormu,
+                     priorphi = priorphi, priorsigma = priorsigma)
 
@@
 @
-\item Finally, visualize (some) posterior draws:
+\item Finally, summarize and visualize (some) posterior draws:
+<<eval=FALSE>>=
+colMeans(draws2[,4:8])
+@
+
+<<echo=FALSE>>=
+colMeans(draws2selection)
+@
+
 <<eval=FALSE>>=
@@
 
-<<hetero, fig=TRUE, include=FALSE, echo=FALSE, results=hide>>=
+<<hetero, fig=TRUE, include=FALSE, echo=FALSE, results=hide, width=7.5, height=8>>=
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
@@
  \begin{center}
-\includegraphics[width=.8\textwidth]{article-hetero}
+\includegraphics[width=.95\textwidth]{article-hetero}
 \caption{Trace plots and kernel density estimates
 %for draws from the marginal posterior distributions of the parameters
-in the regression model with heteroskedastic errors. Data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $h_1^\text{true} = -8.28$, $h_2^\text{true} = -8.50$, $T=1000$.}
+in the regression model with heteroskedastic errors. Data is simulated with $\betab^\text{true}=(0.1, 0.5)^\top$, $h_1^\text{true} = -8.28$, $h_2^\text{true} = -8.50$, $n=1000$.}
 \label{heterodraws}
@@
 \end{figure}
-<<eval=FALSE>>=
-colMeans(draws2[,4:8])
-@
-
-<<echo=FALSE>>=
-colMeans(draws2selection)
-@
+
 \end{itemize}
 
-\section{Real-world example}
-In the following, we aim for a comparison of the performance of the Bayesian normal linear model with homoskedastic errors from Section~\ref{homo} with the Bayesian normal linear model with SV errors from Section~\ref{hetero} using the \code{exrates} data set introduced in Section~\ref{prep}.
+It should be noted that even though \code{svsample2} is considerably faster than \code{svsample}, the cost of interpreting this function in each MCMC iteration is still rather high (which lies in the nature of \proglang{R} as an interpreted language).
+Thus, as of package version \code{0.8-0}, a single-step \code{update} is also made available at the \proglang{C}/\proglang{C++} level for samplers coded there.
+For details, please consult the \code{NEWS} file in the package source; for an application case using this approach, see \cite{kas-etal:ana}.
+
+\section{Illustrative predictive exercise}
+In the following, we compare the performance of the Bayesian normal linear model with homoskedastic errors from Section~\ref{homo} with the Bayesian normal linear model with SV errors from Section~\ref{hetero} using the \code{exrates} dataset introduced in Section~\ref{prep}.
+As a benchmark, we also include the Bayesian normal linear model with GARCH(1,1) errors given through Equation~\ref{reg} with
+\begin{eqnarray*}
+ \label{garch1} \Sigmab &=& \diag{\sigma^2_1,\dots,\sigma^2_n},\\
+ \label{garch2} \sigma^2_t &=& \alpha_0+\alpha_1\tilde y_{t-1}^2+\alpha_2\sigma_{t-1}^2,
+\end{eqnarray*}
+where time index $t=1,\dots,n$. In the second equation, $\tilde y_{t-1}$ denotes the past ``residual'', i.e., the $(t-1)$th element of $\tilde \yb = \yb - \Xb\betab$.
 
@@
 
-We again use the daily price of 1 EUR in USD from January 3, 2000, until April 4, 2012, denoted by $\pb = (p_1, p_2, \dots, p_T)^\top$.
-This time however, instead of using log-returns, we investigate the development of raw prices by regression: Let $\yb$ contain all raw observations except the very first one, and let $\Xb$ denote the design matrix containing ones in the first column and lagged raw prices in the second, i.e.
-\[
+We again use the daily price of 1 EUR in USD from January 3, 2000 to April 4, 2012, denoted by $\pb = (p_1, p_2, \dots, p_n)^\top$.
+This time however, instead of modeling log returns, we investigate the development of log levels by regression.
+Let $\yb$ contain the logarithm of all observations except the very first one, and let $\Xb$ denote the design matrix containing ones in the first column and lagged log prices in the second, i.e.,
+\begin{eqnarray*}
 \yb = 
 \begin{pmatrix}
- p_2\\
- p_3\\
+ \log p_2\\
+ \log p_3\\
  \vdots\\
- p_T
+ \log p_n
 \end{pmatrix}, \quad
@@
 \begin{pmatrix}
-1&p_1\\
-1&p_2\\
+1&\log p_1\\
+1&\log p_2\\
 \vdots&\vdots\\
-1&p_{T-1}
+1&\log p_{n-1}
 \end{pmatrix}.
-\]
+\label{regression}
+\end{eqnarray*}
+Note that this specification simply corresponds to an AR(1) model for the log prices with three different error specifications: homoskedastic, SV, GARCH(1,1).
+It is a generalization of directly dealing with demeaned log returns where $\beta_0$ is a priori assumed to be equal to the mean of the log returns and $\beta_1$ is held constant at $1$.
 %
-<<scatter, fig=TRUE, include=FALSE, results=hide, echo=FALSE, width=9>>=
-x <- exrates$USD[-length(exrates$USD)]
-y <- exrates$USD[-1]
+<<scatter, fig=TRUE, include=FALSE, results=hide, echo=FALSE, width=10.2>>=
+x <- log(exrates$USD[-length(exrates$USD)])
+y <- log(exrates$USD[-1])
 X <- matrix(c(rep(1, length(x)), x), nrow = length(x))
 par(mfrow=c(1,1), mar = c(2.9, 2.9, 2.7, .5), mgp = c(1.8,.6,0), tcl = -.4)
-plot(x,y, xlab=expression(p[t]), ylab=expression(p[t+1]),
-    main="Scatterplot of lagged daily raw prices of 1 EUR in USD",
+plot(x,y, xlab=expression(log(p[t])), ylab=expression(log(p[t+1])),
+    main="Scatterplot of lagged daily log prices of 1 EUR in USD",
     col="#00000022", pch=16, cex=1)
@@
 @
-%
-Clearly, we expect the posterior distribution of $\betab$ to spread around $(0,1)^\top$, which corresponds to a random walk.
-A scatterplot of $p_t$ against $p_{t+1}$, displayed in Figure~\ref{datascatter}, confirms this.
-\begin{figure}[!htp]
-\begin{center}
- \includegraphics[width=.75\textwidth]{article-scatter}
-\end{center}
-\caption{Scatterplot of daily raw prices at time $t$ against daily raw prices at time $t+1$. The solid line indicates the identity function $f(x) = x$.}
+
+Irrespectively of the error specification, we expect the posterior distribution of $\betab$ to spread around $(0,1)^\top$ which corresponds to a random walk.
+A scatterplot of $\log p_t$ against $\log p_{t+1}$, displayed in Figure~\ref{datascatter}, confirms this.
+\begin{figure}[!ht]
+\begin{center}
+ \includegraphics[width=\textwidth]{article-scatter}
+\end{center}
+\caption{Scatterplot of daily log prices at time $t$ against daily log prices at time $t+1$. The solid line indicates the identity function $f(x) = x$.}
 \label{datascatter}
@@
 
-We run both samplers for 100\,000 iterations and discard the first 10\,000 draws as burn-in.
-Prior hyperparameters are chosen as follows: $\bb{0} = (0, 0)^\top$, $\Bb{0} = \diag{10^{10}, 10^{10}}$, $c_0 = C_0 = 0.001$, $b_\mu = -10$, $B_\mu=2$, $a_0=20$, $b_0 = 1.5$, $B_\sigmapar=1$.
-However, due to the length of the dataset (and its obvious heteroskedasticity), the exact prior specification is not very influential.
-The two samplers yield slightly different posteriors for $\betab$, visualized in Figure~\ref{betapost}, where both the marginal posterior densities $p(\beta_0|\yb)$ and $p(\beta_1|\yb)$ as well as a scatterplot of draws from the joint posterior $\betab|\yb$ are displayed.
+To obtain draws from the posterior distributions for the models with homoskedastic/SV errors, samplers developed in Section~\ref{other} are used.
+For the GARCH-errors, a simple random walk Metropolis-Hastings sampler is employed.
+We run each of the three samplers for 100\,000 iterations and discard the first 10\,000 draws as burn-in.
+Starting values and tuning parameters are set using maximum likelihood estimates obtained through the \proglang{R} package \pkg{fGarch} \citep{r:fGarch}.
+
+Aiming for comparable results with minimal prior impact on predictive performance, the hyperparameters are chosen to yield vague priors: $\bb{0} = (0, 0)^\top$, $\Bb{0} = \diag{10^{10}, 10^{10}}$, $c_0 = C_0 = 0.001$, $b_\mu = 0$, $B_\mu=10^4$, $a_0=1$, $b_0 = 1$, $B_\sigmapar=1$.
+For the GARCH(1,1) parameter vector $\bm{\alpha} = (\alpha_0, \alpha_1, \alpha_2)^\top$, we pick independent flat priors on $\mathbb{R}^+$ for the components; the initial variance $\sigma_0^2$ is fixed to the empirical residual variance and $\tilde \y_0$ is assumed to be zero.
+Due to the length of the dataset (and its obvious heteroskedasticity), none of these specific choices seem to particularly influence the following analysis.
+Note, however, that for shorter series or series with less pronounced heteroskedasticity, sensible prior choices are of great importance as the likelihood carries less information in these cases.
+
+The samplers yield slightly different posteriors for $\betab$, visualized in Figure~\ref{betapost}.
+In the top panels, the estimated marginal posterior densities $p(\beta_0|\yb)$ and $p(\beta_1|\yb)$ are displayed; the bottom panel depicts a scatterplot of draws from the joint posterior $\betab|\yb$.
 
@@
 ## initialize some space to hold results:
-realres <- vector('list', 2)
-
+realres <- vector('list', 3)
+
+## AR(1)-SV:
 realres[[1]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
@@
  if (i%%1000 == 0) cat("Iteration", i, "done.\n")
- # draw latent volatilities and AR-parameters:
+ # draw latent volatilities and SV-parameters:
  ytilde <- y - X %*% betadraw
- svdraw <- .svsample(ytilde, startpara=para(svdraw),
+ svdraw <- svsample2(ytilde, startpara=para(svdraw),
 		      startlatent=latent(svdraw), priormu=priormu,
@@
 
-## standard sampler:
+## AR(1) homoskedastic:
 ## pre-calculate some values:
@@
 
+## AR(1)-GARCH(1,1):
+## allocate space:
+realres[[3]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
+colnames(realres[[3]]) <- c("a_0", "a_1", "b_1", paste("beta", 0:(p-1), sep='_'),
+			   paste("sigma", 1:n, sep='_'))
+
+## some auxiliary functions:
+loglik <- function(y, s2) {
+ sum(dnorm(y, 0, sqrt(s2), log=TRUE))
+}
+
+s2s <- function(y, para, y20, s20) {
+ s2 <- rep(NA_real_, length(y))
+ s2[1] <- para[1] + para[2]*y20 + para[3]*s20
+ for (i in 2:length(y)) s2[i] <- para[1] + para[2]*y[i-1]^2 + para[3]*s2[i-1]
+ s2	
+}
+
+## This is a very simple (and slow) random-walk MH sampler
+## which can most certainly be improved...
+
+## initial values:
+
+# crude approximation of residual variance:
+tmplm <- lm.fit(X, y)
+s20 <- var(resid(tmplm))
+
+y20 <- 0
+
+## starting values:
+## For the starting values for the parameters,
+## use ML results from fGarch::garchFit on the
+## residuals
+
+# tmpfit <- fGarch::garchFit(~garch(1,1), resid(tmplm), trace=FALSE)
+# startvals <- fGarch::coef(tmpfit)[c("omega", "alpha1", "beta1")]
+startvals <- c(1.532868e-07, 3.246417e-02, 9.644149e-01)
+
+## Random walk MH algorithm needs manual tuning for good mixing.
+## Gelman et al. (1996) suggest an acceptance rate of 31.6%
+## for spherical 3d multivariate normal (which we don't have :))
+## These values seem to work OK:
+# mhtuning <- tmpfit@fit$se.coef[c("omega", "alpha1", "beta1")] / 2
+mhtuning <- c(3.463851e-08, 2.226946e-03, 2.354829e-03)
+
+betadraw <- as.numeric(coef(tmplm))
+
+paradraw <- startvals
+paraprop <- rep(NA_real_, 3)
+
+s2draw <- rep(.1, length(y))
+s2prop <- rep(NA_real_, length(y))
+
+accepts <- 0
+
+## sampler:
+tim3 <- system.time(
+for (i in -(burnin-1):draws) {
+
+ if (i%%1000 == 0) cat("Iteration", i, "done.\n")
+
+ # draw volatilities and GARCH-parameters:
+ ytilde <- y - X %*% betadraw  # regression residuals
+
+ paraprop <- rnorm(3, paradraw, mhtuning)
+ s2prop <- s2s(ytilde, paraprop, y20, s20)
+
+ if (all(s2prop > 0)) { # s2 needs to be positive, otherwise reject
+  logR <- loglik(ytilde, s2prop) - loglik(ytilde, s2draw)  # log acceptance rate
+  if (log(runif(1)) < logR) {
+   paradraw <- paraprop
+   s2draw <- s2prop
+   if (i > 0) accepts <- accepts + 1
+  }
+ }
+
+ # draw the betas:
+ normalizer <- 1/sqrt(s2draw)
+ Xnew <- X * normalizer
+ ynew <- y * normalizer
+ Sigma <- solve(crossprod(Xnew) + B0inv)
+ mu <- Sigma %*% (crossprod(Xnew, ynew) + B0inv %*% b0)
+ betadraw <- as.numeric(mvtnorm::rmvnorm(1, mu, Sigma))
+
+ # store the results:
+ if (i > 0 & i %% thinning == 0) { 
+  realres[[3]][i/thinning, paras] <- paradraw
+  realres[[3]][i/thinning, latents] <- sqrt(s2draw)
+  realres[[3]][i/thinning, betas] <- betadraw
+ }
+}
+)[["elapsed"]]
+
 #Standardized residuals for model checking:
@@
 
-#heteroskedastic
+#heteroskedastic SV
 predmeans <- tcrossprod(X, realres[[1]][,4:5])
@@
 
-realresselection <- vector('list', 2)
+#heteroskedastic GARCH
+predmeans3 <- tcrossprod(X, realres[[3]][,4:5])
+standresidmeans3 <- rowMeans((y - predmeans3)/t(realres[[3]][,6:ncol(realres[[3]])]))
+
+realresselection <- vector('list', 3)
 realresselection[[1]] <- realres[[1]][,c('beta_0', 'beta_1')]
 realresselection[[2]] <- realres[[2]][,c('beta_0', 'beta_1')]
-#save(realresselection, standresidmeans, standresidmeans2, file = 'vignette_sampling_realworld.RData')
-}
-@
-
-<<betapost, echo=FALSE, result=hide, fig=TRUE, include=FALSE, width=9>>=
+realresselection[[3]] <- realres[[3]][,c('beta_0', 'beta_1')]
+#save(realresselection, standresidmeans, standresidmeans2, standresidmeans3, file = 'vignette_sampling_realworld.RData')
+}
+@
+
+<<betapost, echo=FALSE, result=hide, fig=TRUE, include=FALSE, width=8.7, height=7>>=
 smootherfactor <- 1.8
@@
 plot(density(realresselection[[1]][,"beta_0"], bw="SJ", adjust=smootherfactor),
-    main=bquote(paste("p(",beta[0],"|y)", sep="")),
-    xlab="", ylab="", xlim=c(-.0025, .0045))
-lines(density(realresselection[[2]][,"beta_0"], bw="SJ", adjust=smootherfactor), lty=2, col=2)
+    main=bquote(paste("p(",beta[0],"|",bold(y),")", sep="")),
+    xlab="", ylab="", xlim=c(-.0005, .001))
+lines(density(realresselection[[3]][,"beta_0"], bw="SJ", adjust=smootherfactor), lty=2, col=4)
+lines(density(realresselection[[2]][,"beta_0"], bw="SJ", adjust=smootherfactor), lty=3, col=2)
 #points(ols$coefficients[1],0, col=2)
 #lines(confint(ols)[1,], rep(0,2), col=2)
-legend("topleft", c("SV", "homosked."), col=1:2, lty=1:2)
+legend("topleft", c("SV", "GARCH", "homosked."), col=c(1,4,2), lty=1:3)
 
 plot(density(realresselection[[1]][,"beta_1"], bw="SJ", adjust=smootherfactor),
-    main=bquote(paste("p(",beta[1],"|y)", sep="")),
+    main=bquote(paste("p(",beta[1],"|",bold(y),")", sep="")),
     xlab="", ylab="", xlim=c(0.9965, 1.0022))
-lines(density(realresselection[[2]][,"beta_1"], bw="SJ", adjust=smootherfactor), lty=2, col=2)
+lines(density(realresselection[[3]][,"beta_1"], bw="SJ", adjust=smootherfactor), lty=2, col=4)
+lines(density(realresselection[[2]][,"beta_1"], bw="SJ", adjust=smootherfactor), lty=3, col=2)
 #points(ols$coefficients[2],0, col=2)
 #lines(confint(ols)[2,], rep(0,2), col=2)
-legend("topright", c("SV", "homosked."), col=1:2, lty=1:2)
-
-plotorder <- sample.int(2*nrow(realresselection[[1]]))
-cols <- rep(c("#00000055", "#ff000055"), each=nrow(realresselection[[1]]))[plotorder]
-pchs <- rep(1:2, each=nrow(realresselection[[1]]))[plotorder]
-beta0 <- c(realresselection[[1]][,"beta_0"], realresselection[[2]][,"beta_0"])[plotorder]
-beta1 <- c(realresselection[[1]][,"beta_1"], realresselection[[2]][,"beta_1"])[plotorder]
+legend("topright", c("SV", "GARCH", "homosked."), col=c(1,4,2), lty=1:3)
+
+plotorder <- sample.int(3*nrow(realresselection[[1]]))
+cols <- rep(c("#000000aa", "#0000ff66", "#ff000077"), each=nrow(realresselection[[1]]))[plotorder]
+pchs <- rep(1:3, each=nrow(realresselection[[1]]))[plotorder]
+
+beta0 <- c(realresselection[[1]][,"beta_0"],
+	   realresselection[[3]][,"beta_0"],
+	   realresselection[[2]][,"beta_0"])[plotorder]
+
+beta1 <- c(realresselection[[1]][,"beta_1"],
+	   realresselection[[3]][,"beta_1"],
+	   realresselection[[2]][,"beta_1"])[plotorder]
 
 plot(beta0, beta1, col=cols, pch=pchs,
-    xlab=bquote(paste("p(",beta[0],"|y)")),
-    ylab=bquote(paste("p(",beta[1],"|y)")),
+    xlab=bquote(paste("p(",beta[0],"|", bold(y), ")")),
+    ylab=bquote(paste("p(",beta[1],"|", bold(y), ")")),
     main="Scatterplot of posterior draws")
-legend("topright", c("SV", "homosked."), col=1:2, pch=1:2)
+legend("topright", c("SV", "GARCH", "homosked."), col=c(1,4,2), pch=1:3)
 @
@@
  \includegraphics[width=\textwidth]{article-betapost}
- \caption{Visualization of the posterior distributions $\betab|\yb$ for the model with SV residuals and the model with homoskedastic errors.
-Top panels: kernel density estimate of the univariate posterior marginal distributions.
-Bottom panel: bivariate scatterplot of posterior draws.}
+ \caption{Visualization of the posterior distributions $\betab|\yb$ for the model with SV regression residuals, the model with GARCH(1,1) regression residuals, and the model with homoskedastic regression residuals.
+Top panels: Kernel density estimates of the univariate posterior marginal distributions.
+Bottom panel: Bivariate scatterplot of posterior draws.}
   \label{betapost}
@@
 
-To assess the model fit, mean standardized resdiduals for both samplers are depicted in Figure~\ref{qqplot}.
-It stands out that the model with homoskedastic errors shows deficiencies in terms of heavy correlation amongst the residuals.
+To assess the model fit, mean standardized residuals are depicted in Figure~\ref{qqplot}.
+The model with homoskedastic errors shows deficiencies in terms of pronounced dependence amongst the residuals.
 This can clearly be seen in the top left panel, where mean standardized residuals are plotted against time.
-The bottom left panel shows the same plot for the model with SV errors, where this effect practically vanishes.
+The middle left panel shows the same plot for the model with GARCH errors where this effect is greatly diminished.
+The bottom left panel pictures that plot for the model with SV errors; here, this effect practically vanishes.
 Moreover, in the model with homoskedastic errors, the normality assumption about the unconditional error distribution is clearly violated.
-This can be seen by inspecting the quantile-quantile plot in the top right panel, where observed residuals show much heavier tails than one would expect from a normal distribution.
-On the contrary, standardized residuals obtained from the model with SV errors align almost perfectly.
-
-<<qqplot, echo=FALSE, result=hide, fig=TRUE, include=FALSE, width=9, height=7>>=
-par(mfrow=c(2,2), mar = c(3.1, 3.3, 2.0, .5), mgp = c(1.7,.5,0),
+This can be seen by inspecting the quantile-quantile plot in the top right panel, where observed residuals exhibit much heavier tails than one would expect from a normal distribution.
+The model with GARCH errors provides a better fit, however, heavy tails are still visible.
+Standardized residuals from the model with SV errors appear to be approximately normal with only few potential outliers.
+
+<<qqplot, echo=FALSE, result=hide, fig=TRUE, include=FALSE, width=8, height=7>>=
+par(mfrow=c(3,2), mar = c(3.1, 3.3, 2.0, .5), mgp = c(1.7,.5,0),
     tcl = -.4)
@@
        standresidmeans2, main="Residual Q-Q plot (homoskedastic errors)",
+       xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
+abline(0,1)
+plot(standresidmeans3, main="Residual scatterplot (GARCH errors)",
+     xlab="Time", ylab="Standardized residuals")
+qqplot(qnorm(ppoints(length(standresidmeans3)), mean=0, sd=1),
+       standresidmeans3, main="Residual Q-Q plot (GARCH errors)",
        xlab = "Theoretical N(0,1)-quantiles", ylab = "Empirical Quantiles")
@@
  \begin{center}
-  \includegraphics[width=.9\textwidth]{article-qqplot}
+  \includegraphics[width=\textwidth]{article-qqplot}
  \end{center}
  \caption{Visualization of mean standardized residuals.
-Top left panel shows a scatterplot against time for the model with homoskedastic errors, bottom left panel shows this plot for the model with SV errors.
+Top left panel shows a scatterplot against time for the model with homoskedastic errors, middle left panel shows this plot for the model with GARCH errors, and bottom left panel shows this plot for the model with SV errors.
 Quantile-Quantile plots of empirical quantiles against expected quantiles from a $\Normal{0,1}$-distribution are displayed on the panels on the right-hand side.}
@@
 
-\subsection{Predictive performance and model fit}
+\subsection{Measuring predictive performance and model fit}
 Within a Bayesian framework, a natural way of assessing the predictive performance of a given model is through its \emph{predictive density} (sometimes also referred to as \emph{posterior predictive distribution}).
-Collecting all \emph{unobservables}, i.e., parameters and possible latent variables, in a single vector $\unobs$, it is given through
+It is given through
 \begin{eqnarray}
@@
  p(\unobs|\yb^o_{[1:t]}) 
- \, \dif \unobs.
+ \, \dif \unobs,
 \end{eqnarray}
-Note that for the model with homoskedastic errors, $\unobs=(\betab,\sigmaeps)^\top$, while for the model with SV errors, $\unobs=(\betab, \allpara, \hb)^\top$.
+where $\unobs$ denotes the vector of all unobservables, i.e., parameters and possible latent variables.
+Note that for the model with homoskedastic errors, $\unobs=(\betab,\sigmaeps)^\top$; for the model with SV errors, $\unobs=(\betab, \allpara, \hb)^\top$; in the GARCH(1,1) case, $\unobs=(\betab,\bm{\alpha},\sigma_0^2,\tilde y_0^2)^\top$.
 By using the superscript $o$ in $\yb^o_{[1:t]}$, we follow \cite{gew-ami:com} and denote \emph{ex post} realizations (observations) for the set of points in time $\{1,2,\dots ,t\}$ of the \emph{ex ante} random values $\yb_{[1:t]} = (\y_1, \y_2, \dots,\y_t)^\top$.
-Integration is carried out over ${\bm K}$, which simply stands for the space of all possible values for $\unobs$.
+Integration is carried out over ${\bm K}$ which simply stands for the space of all possible values for $\unobs$.
 Equation~\ref{preddens} can be viewed as the integral of the likelihood function over the joint posterior distribution of the unobservables $\unobs$.
@@
 
-In the SV errors case, Equation~\ref{preddens} is a $(T+p+3)$-dimensional integral which cannot be solved analytically.
-Nevertheless, it may be evaluated at an arbitrary point $x$ through Monte Carlo integration:
+In the SV errors case, Equation~\ref{preddens} is a $(n+p+3)$-dimensional integral which cannot be solved analytically.
+Nevertheless, it may be evaluated at an arbitrary point $x$ through Monte Carlo integration,
 \begin{equation}
@@
 where $\unobs^{(m)}_{[1:t]}$ stands for the $m$\textsuperscript{th} draw from the respective posterior distribution up to time $t$.
-If Equation~\ref{preddensMC} is evaluated at $x=y_{t+1}^o$, we refer to it as the (one-step-ahead) \emph{predicitve likelihood} (at time $t+1$) denoted $\PL_{t+1}$.
+If Equation~\ref{preddensMC} is evaluated at $x=y_{t+1}^o$, we refer to it as the (one-step-ahead) \emph{predictive likelihood} at time $t+1$, denoted $\PL_{t+1}$.
 Moreover, draws from (\ref{preddens}) can be obtained by simulating values $\y_{t+1}^{(m)}$ from the distribution given through the density $p(\y_{t+1}|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]})$, the summands of Equation~\ref{preddensMC}.
 
-For model at hand, the predictive density and likelihood can thus be computed through the following
+For the model at hand, the predictive density and likelihood can thus be computed through the following
 
@@
  \begin{enumerate}
-  \item Reduce the data set to a training set $\yb^o_{[1:t]} = (\y_1^o,\dots,y_t^o)^\top$.
-  \item Run the posterior sampler using data from the training set only to obtain $M$ posterior draws $\unobs^{(m)}_{[1:t]}$.
-  \item[(3.)] Needed for the SV model only: Simulate $M$ values from the conditional distribution $\hsv_{t+1,[1:t]}|\yb^o_{[1:t]},\unobs_{[1:t]}$ by drawing $\hsv_{t+1,[1:t]}^{(m)}$ from a normal distribution with mean $\mupar_{[1:t]}^{(m)} + \phipar_{[1:t]}^{(m)}(\hsv^{(m)}_{t,[1:t]}-\mupar_{[1:t]}^{(m)})$ and standard deviation $\sigmaparextrasub{[1:t]}^{(m)}$ for $m=1,\dots,M$.
-  \item[4a.] To obtain $\PL_{t+1}$, average over $M$ densities of normal distributions with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model) or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model), each evaluated at $y^o_{t+1}$.
-  \item[4b.] To obtain $M$ draws from the predictive distribution, simulate from a normal distribution with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model) or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model) for $m=1,\dots,M$.
+  \item Reduce the dataset to a training set $\yb^o_{[1:t]} = (\y_1^o,\dots,y_t^o)^\top$.
+  \item Run the posterior sampler using data from the training set only to obtain $M$ posterior draws $\unobs^{(m)}_{[1:t]}$, $m=1,\dots, M$.
+  \item[(3a.)] Needed for the SV model only: Simulate $M$ values from the conditional distribution $\hsv_{t+1,[1:t]}|\yb^o_{[1:t]},\unobs_{[1:t]}$ by drawing $\hsv_{t+1,[1:t]}^{(m)}$ from a normal distribution with mean $\mupar_{[1:t]}^{(m)} + \phipar_{[1:t]}^{(m)}(\hsv^{(m)}_{t,[1:t]}-\mupar_{[1:t]}^{(m)})$ and standard deviation $\sigmaparextrasub{[1:t]}^{(m)}$ for $m=1,\dots,M$.
+  \item[(3b.)] Needed for the GARCH model only: Obtain $M$ draws from the conditional distribution $\sigma_{t+1,[1:t]}|\yb^o_{[1:t]},\unobs_{[1:t]}$ by computing $\sigma_{t+1,[1:t]}^{(m)} = \sqrt{\alpha_{0,[1:t]}^{(m)}+\alpha_{1,[1:t]}^{(m)}\left(\tilde y_t^o\right)^2+\alpha_{2,[1:t]}^{(m)}\left(\sigma_{t,[1:t]}^{(m)}\right)^2}$ for $m=1,\dots,M$.
+  \item[4a.] To obtain $\PL_{t+1}$, average over $M$ densities of normal distributions with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model), $\sigma_{t+1,[1:t]}^{(m)}$ (GARCH model), or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model), each evaluated at $y^o_{t+1}$, for $m=1,\dots, M$.
+  \item[4b.] To obtain $M$ draws from the predictive distribution, simulate from a normal distribution with mean $(1,\y^o_t)\times\betab^{(m)}_{[1:t]}$ and standard deviation $\exp\{\hsv_{t+1,[1:t]}^{(m)}/2\}$ (SV model), $\sigma_{t+1,[1:t]}^{(m)}$ (GARCH model), or $\sigmaepsextrasub{[1:t]}^{(m)}$ (homoskedastic model) for $m=1,\dots,M$.
  \end{enumerate}
@@
 
-To avoid strong dependence on the prior, the first 1000 days are used as training set only and the evaluation of the predictive distribution starts at $t=1001$, corresponding to December 4, 2003.
-The results for both models are displayed in Figure~\ref{predlik1}.
-In the top panel, the observed series along with the 98\% one-day-ahead predictive intervals are displayed.
-The bottom panel shows the log one-day-ahead predictive likelihood.
-
-Figure~\ref{predlik2} displays a zoomed-in version, depicting only the time span from January 2008 until August 2009.
-Note that while at the beginning of 2008 both credible intervals are very similar, there is a substantial difference one year later, where SV intervals become around twice as large compared to the corresponding homoskedastic analogs.
-According to the values of the log predictive likelihoods, SV errors can handle the inflated volatility during that time substantially better.
-Throughout 2009, the width of the intervals as well as the predictive likelihoods consolidate again.
-
-\begin{figure}[!htp]
-\begin{center}
- \includegraphics[width=\textwidth]{predlik1}
-\end{center}
-\caption{Top panel: Observed series (green) and symmetrical 98\% one-day-ahead predictive intervals for the model with homoskedastic errors (red) and the model with SV errors (black).
-Bottom panel: Log one-day-ahead predictive likelihoods for both models.}
-\label{predlik1}
-\end{figure}
-
-\begin{figure}[!htp]
-\begin{center}
- \includegraphics[width=\textwidth]{predlik2}
-\end{center}
-\caption{Zoomed version of Figure~\ref{predlik1}, showing only the period from January 2008 until August 2009.
-This time span is chosen to includes the beginning of the financial crisis.
-During that phase, predictive performance of the model with homoskedastic errors deteriorates substantially, while SV errors can capture the inflated volatility much better.}
-\label{predlik2}
-\end{figure}
-
 It is worth pointing out that log predictive likelihoods also carry an intrinsic connection to the log \emph{marginal likelihood}, defined through
-\begin{eqnarray}
+\begin{eqnarray*}
  \label{ML}
@@
  \, \dif \unobs.
-\end{eqnarray}
-This real number corresponds to the logarithm of the normalizing constant in the denominator of Bayes' law and is often
-used for evaluating model evidence. It can straightforwardly be decomposed into the sum of the one-step-ahead logarithms
-of the predictive likelihoods:
+\end{eqnarray*}
+This real number corresponds to the logarithm of the normalizing constant which appears in the denominator of Bayes' law and is often
+used for evaluating model evidence. It can straightforwardly be decomposed into the sum of the logarithms of the one-step-ahead predictive likelihoods:
 \[
-\log \ML = \log p(\yb^o) = \log \prod_{t=1}^T p(\y_t^o|\yb_{[1:t-1]}^o) = \sum_{t=1}^T \log \PL_{t}.
+\log \ML = \log p(\yb^o) = \log \prod_{t=1}^n p(\y_t^o|\yb_{[1:t-1]}^o) = \sum_{t=1}^n \log \PL_{t}.
 \]
 Thus, Algorithm \ref{predlikalg} provides a conceptually simple way of computing the marginal likelihood.
-However, these computations are quite costly in terms of CPU time, as they require an individual model fit for each of the $T$ points in time.
-On the other hand, due to the embarrassingly parallel nature of the task and because of today's comparably easy access to parallel computing environments, this burden becomes easily manageable.
-E.g., the computations for the analysis in this paper have been conducted in less than one hour, using 25 IBM dx360M3 nodes within a cluster of workstations.
+However, these computations are quite costly in terms of CPU time, as they require an individual model fit for each of the $n$ points in time.
+On the other hand, due to the embarrassingly parallel nature of the task and because of today's comparably easy access to distributed computing environments, this burden becomes manageable.
+E.g., the computations for the stochastic volatility analysis in this paper have been conducted in less than one hour, using 25 IBM dx360M3 nodes within a cluster of workstations.
 Implementation in \proglang{R} was achieved through the packages \pkg{parallel} \citep{r:r} and \pkg{snow} \citep{r:sno}.
@@
 Cumulative sums of $\log \PL_t$ also allow for model comparison through cumulative log predictive Bayes factors.
-Letting $\PL_t(A)$ denote the predictive likelihood of model $A$ at time $t$, and $\PL_t(B)$ the corresponding value of model $B$, the cumulative log predictive Bayes factor at time $u$ (and starting point $S$) in favor of model $A$ over model $B$ is simply given through
+Letting $\PL_t(A)$ denote the predictive likelihood of model $A$ at time $t$, and $\PL_t(B)$ the corresponding value of model $B$, the cumulative log predictive Bayes factor at time $u$ (and starting point $s$) in favor of model $A$ over model $B$ is straightforwardly given through
 \begin{equation}
  \label{clpbf}
-\log \left[ \frac{p_A(\yb^o_{[S+1:u]}|\yb^o_{[1:S]})}{p_B(\yb^o_{[S+1:u]}|\yb^o_{[1:S]})} \right] =  \sum_{t=S+1}^u \log \left[ \frac{\PL_t(A)}{\PL_t(B)} \right] =  \sum_{t=S+1}^u [\log \PL_t(A) - \log \PL_t(B)].
+\log \left[ \frac{p_A(\yb^o_{[s+1:u]}|\yb^o_{[1:s]})}{p_B(\yb^o_{[s+1:u]}|\yb^o_{[1:s]})} \right] =  \sum_{t=s+1}^u \log \left[ \frac{\PL_t(A)}{\PL_t(B)} \right] =  \sum_{t=s+1}^u [\log \PL_t(A) - \log \PL_t(B)].
 \end{equation}
 
-If the cumulative log predictive Bayes factor is positive at a given point in time, we have evidence in favor of model $A$, and vice versa.
-In this context, information up to time $S$ is regarded as prior information, while out-of-sample predictive evaluation starts at time $S+1$.
-Note that the usual log Bayes factor is a special case of Equation~\ref{clpbf} for $S=0$ and $u=T$.
-
-The top panel of Figure~\ref{predlik3} gives an overview of the model performance through visual inspection of observed residuals against their predicted distributions.
-For the purpose of this paper, residuals are simply defined as the deviation from the median of the predicted distribution.
-It stands out that predictive quantiles arising from the model with SV errors exhibit much more flexibility to ``adapt'' to the ``current state of the world'', while the simpler homoskedastic model barely exhibits this feature.
-While there is little difference in predicted residuals until the beginning of 2007, the model with SV residuals performs substantially better during the pre-crisis era (less volatility) and during the actual crisis (more volatility).
-This picture is confirmed through the cumulative log predictive Bayes factors displayed in the bottom panel.
-Note that the last point plotted equals around $223$, providing overwhelming overall evidence in favor of the model with SV errors, clearly rejecting the assumption of homoskedastic error terms.
-
-\begin{figure}[!ht]
+When the cumulative log predictive Bayes factor is positive at a given point in time, there is evidence in favor of model $A$, and vice versa.
+In this context, information up to time $s$ is regarded as prior information, while out-of-sample predictive evaluation starts at time $s+1$.
+Note that the usual (overall) log Bayes factor is a special case of Equation~\ref{clpbf} for $s=0$ and $u=n$.
+
+\subsection{Results}
+
+In order to reduce prior influence, the first 1000 days are used as a training set only and the evaluation of the predictive distribution starts at $t=1001$, corresponding to December 4, 2003.
+The homoskedastic model is compared to the model with SV residuals in Figure~\ref{predlik1}.
+In the top panel, the observed series along with the 98\% one-day-ahead predictive intervals are displayed.
+The bottom panel shows the log one-day-ahead predictive likelihood.
+According to these values, SV errors can handle the inflated volatility during that time substantially better.
+In the course of 2009, the width of the intervals as well as the predictive likelihoods consolidate again.
+Figure~\ref{predlik1} also contains a close-up of the one-year time span from September 2008 to August 2009.
+Both credible intervals are similar at the beginning and at the end of this interval.
+However, there is a substantial difference in early 2009, where SV intervals become around twice as large compared to the corresponding homoskedastic analogs.
+
+A visually barely distinguishable picture emerges when GARCH(1,1) errors are employed instead of SV errors, thus no separate figure is included in this article.
+
+\begin{figure}[!htp]
+\begin{center}
+ \includegraphics[width=\textwidth]{predlik1}
+\end{center}
+\caption{Top panel: Observed series (green) and symmetrical 98\% one-day-ahead predictive intervals for the model with homoskedastic errors (red) and the model with SV errors (black).
+The display also contains a close-up, showing only the period from September 2008 until August 2009.
+This time span is chosen to include the most noticeable ramifications of the financial crisis.
+During that phase, predictive performance of the model with homoskedastic errors deteriorates substantially, while SV errors can capture the inflated volatility much better.
+Bottom panel: Log one-day-ahead predictive likelihoods for both models.}
+\label{predlik1}
+\end{figure}
+
+The top panel of Figure~\ref{predlik3} visualizes the performance of the three models; it depicts observed regression residuals against their (one-day-ahead) predicted distributions.
+For the purpose of this image, observed regression residuals are simply defined as the deviation of the data from the median of the predicted distribution.
+It stands out that predictive quantiles arising from the models with heteroskedastic errors exhibit much more flexibility to adapt to the ``current state of the world'', while the simpler homoskedastic model barely exhibits this feature.
+
+Generally speaking, there is little difference in predicted residuals until the beginning of 2007.
+However, during the pre-crisis era (less volatility) and during the actual crisis (more volatility), the models catering for heteroskedasticity perform substantially better.
+It is interesting to see that predictive intervals for the models with SV errors and GARCH errors are very similar.
+
+In the bottom panel of Figure~\ref{predlik3}, the cumulative sums of the log predictive Bayes factors are displayed.
+The last points plotted equal to $176.52$ in the SV case and $166.37$ in the GARCH case; this provides overwhelming overall evidence in favor of a model catering for heteroskedasticity and ``decisive'' \citep{jef:the3ed} respectively ``very strong'' \citep{kas-raf:bay} evidence in favor of SV over vanilla GARCH(1,1) with a final cumulative predictive Bayes factor around $25\,000:1$.
+%and thus clearly rejects the assumption of homoskedastic error terms for this dataset.
+
+\begin{figure}[!htp]
 \begin{center}
@@
 \caption{Top panel: Observed residuals with respect to the median of the one-day-ahead predictive distribution along with 1\% and 99\% quantiles of the respective predictive distributions.
-It can clearly be seen that the variance of the predictive distribution in the SV model adjusts to heteroskedasticity, while the model with homoskedastic errors is much more restrictive.
+It can clearly be seen that the variances of the predictive distributions in the GARCH and SV models adjust to heteroskedasticity, while the model with homoskedastic errors is much more restrictive.
 Bottom panel: Cumulative log predictive Bayes factors in favor of the model with SV residuals.
-Values greater than zero mean that the model with SV residuals performs better out of sample up to this point in time.}
+Values greater than zero mean that the model with GARCH/SV residuals performs better out of sample up to this point in time.}
 \label{predlik3}
@@
 
-For an example of using \pkg{stochvol} for univariate SV updates within a factor SV framework, see \cite{kas-etal:ana}.
+It is worth noting that for other exchange rates contained in \code{exrates}, a qualitatively similar picture emerges.
+For an overview of cumulative log predictive likelihoods, see Table~\ref{exrates}.
+Detailed results about these currencies are available from the author.
+
+As pointed out above, the analysis of log returns can be viewed upon as a special case of the analysis of log levels where $\beta_1 \equiv 1$ is fixed a priori.
+When doing so, evidence in favor of heteroskedasticity is again striking.
+Once more, the model with SV scores highest; its demeaned predictive quantiles are almost indistinguishable from the ``varying $\beta_1$ case''.
+Predictive quantiles of the model with GARCH residuals again resemble those of the model with SV residuals very closely.
+%Also, predictive performance is similar; the cumulative log predictive Bayes factors amount to $177.36$ for SV vs.\ homoskedastic and $166.64$ for GARCH vs.\ homoskedastic at $t=n$.
+The sum of log predictive likelihoods for $t \in \{1001,1002,\dots,n\}$ is given in the last line of Table~\ref{exrates}.
+The values are slightly but consistently lower than for the models where $\beta_1$ is estimated from the data.
+The display of predictive intervals for the log returns is practically identical to Figure~\ref{predlik3} and thus omitted here.
+%Figure~\ref{predlik3a} displays the predictive intervals for the log returns and the comparison through predictive likelihoods and predictive Bayes factors.
+It is practically identical to Figure~\ref{predlik3}.
+
+Concluding this somewhat anecdotal prediction exercise, we would like to note that for a thorough and more universal statement concerning the real-world applicability and predictive accuracy of \pkg{stochvol}, further studies with different datasets and a richer variety of competing models, potentially including realized volatility measures, are called for.
+Such a voluminous undertaking is, however, beyond the scope of this paper.
+
+\begin{table}[!ht]
+ \centering
+ \begin{tabular}{lccc}
+  \hline
+  \hline
+ Currency & SV & GARCH & homoskedastic \\
+  \hline
+  AUD & 7938 & 7890 & 7554 \\
+  CAD & 7851 & 7844 & 7714 \\
+  CHF & 9411 & 9337 & 8303 \\
+  CZK & 9046 & 8995 & 8659 \\
+  DKK\tablefootnote{For exchange rates of the Danish krone with respect to the euro, we analyze \code{1000*log(exrates\$DKK)}, i.e., per mille (\textperthousand) log prices.
+This way, we avoid convergence issues when obtaining starting values for the MCMC sampler from \pkg{fGarch} that appear otherwise because the krone is pegged very closely to the euro.}
+  & 1473 & 1366 & 1178 \\
+  GBP & 8568 & 8549 & 8218 \\
+  HKD & 7907 & 7897 & 7728 \\
+  IDR & 7697 & 7662 & 7269 \\
+  JPY & 7607 & 7586 & 7280 \\
+  KRW & 7766 & 7749 & 7188 \\
+  MXN & 7536 & 7507 & 7055 \\
+  MYR & 8082 & 8064 & 7928 \\
+  NOK & 8648 & 8616 & 8331 \\
+  NZD & 7619 & 7587 & 7440 \\
+  PHP & 7890 & 7862 & 7654 \\
+  PLN & 8126 & 8094 & 7727 \\
+  RON & 9011 & 8880 & 8255 \\
+  RUB & 8664 & 8617 & 8146 \\
+  SEK & 9110 & 9101 & 8648 \\
+  SGD & 8540 & 8529 & 8308 \\
+  THB & 7867 & 7844 & 7692 \\
+  \hline
+  USD & 7878 & 7868 & 7701 \\
+  USD [AR(0)]\tablefootnote{These results refer to the analysis of log returns by means of a intercept-only model, i.e., AR(0).}
+  & 7876 & 7865 & 7699 \\
+  \hline
+  \hline
+ \end{tabular}
+ \caption{Final cumulative predictive log likelihoods for AR(1) models with different error assumptions for the logarithm of several EUR exchange rates included in \pkg{stochvol}'s dataset \code{exrates}.
+SV is strongly favored in all cases.}
+ \label{exrates}
+\end{table}
 
 \section{Conclusion}
-Aim of this paper was to introduce to the functionality of the \proglang{R} package \pkg{stochvol}, which provides a fully Bayesian simulation-based approach for inference in stochastic volatility models.
-The typical workflow occurring when using \pkg{stochvol} was presented by analyzing exchange rate data in the package's \code{exrates} data set.
+The aim of this article was to introduce the reader to the functionality of \pkg{stochvol}.
+This \proglang{R} package provides a fully Bayesian simulation-based approach for statistical inference in stochastic volatility models.
+The typical application workflow of \pkg{stochvol} was illustrated through the analysis of exchange rate data contained in the package's \code{exrates} dataset.
 Furthermore, it was shown how the package can be used as a ``plug-in'' tool for other MCMC samplers.
@@
 
-In the real-world example, raw exchange rates from EUR to USD were analyzed.
-Out-of-sample analysis through cumulative predictive Bayes factors clearly showed that already for a simple linear autoregressive model for the raw exchange rates, SV residuals substantially improve prediction performance, especially in turbulent times.
+In the predictive example, both log levels of exchange rates from EUR to USD and log returns thereof were analyzed.
+For this dataset, out-of-sample analysis through cumulative predictive Bayes factors clearly showed that modeling regression residuals heteroskedastically substantially improves predictive performance, especially in turbulent times.
+A direct comparison of SV and vanilla GARCH(1,1) indicated that the former performs better in terms of predictive accuracy.
+
+\section*{Acknowledgments}
+
+The author would like to thank Sylvia Fr\"{u}hwirth-Schnatter, Hedibert Freitas Lopes, Karin Dobernig, and two anonymous referees for helpful comments and suggestions.
 

### 2015-01-28 inst/doc/article.Rnw
version 1.0.0

@@
 
-%\VignetteIndexEntry{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol}
+%\VignetteIndexEntry{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol}
 %\VignetteKeyword{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
@@
 %\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
-\Plaintitle{Dealing With Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting
+\Plaintitle{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting
 
@@
 \Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
-\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
+%\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
 %% at least one keyword must be supplied
@@
   WU Vienna University of Economics and Business\\
+%  Building D4, Level 4\\
   Welthandelsplatz 1\\
   1020 Vienna, Austria\\
-  E-mail: \email{gregor.kastner@wu.ac.at}
-%  URL: \url{http://eeecon.uibk.ac.at/~zeileis/}
+%  Telephone: +43/1/31336-5593\\
+%  Fax: +43/1/31336-90-5593\\
+  E-mail: \email{gregor.kastner@wu.ac.at}\\
+  URL: \url{http://www.wu.ac.at/statmath/en/faculty_staff/faculty/gkastner/}
 }
@@
 @
+
+\section*{Preface}
+This vignette corresponds to a forthcoming article of the same name in the Journal of Statistical Software.
+At the time of this writing, the vignette and the published article coincide; nevertheless, this version might receive minor updates over time.
+To cite, please use \cite{kas:dea}.
+Further information about citing \pkg{stochvol} can be obtained in \proglang{R} by installing the package, e.g., through \code{install.packages("stochvol")}, and typing \code{citation("stochvol")}.
 
@@
 Even though several papers \citep[e.g.,][]{jac-etal:bayJBES, ghy-etal:sto, kim-etal:sto} provide early evidence in favor of using SV, these models have found comparably little use in applied work.
-This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard packages implementing these methods.
+This obvious discrepancy is discussed in \cite{bos:rel} who points out two reasons: the variety (and potential incompatibility) of estimation methods for SV models -- whereas the many variants of the GARCH model have basically a single estimation method -- and the lack of standard software packages implementing these methods.
 
@@
 is proposed.
-The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use packages for efficiently estimating SV models.
+The paper at hand and the corresponding package \pkg{stochvol} \citep{r:sto} for \proglang{R} \citep{r:r} are crafted to cope with the latter problem: the apparent lack of ready-to-use software packages for efficiently estimating SV models.
 
@@
 
-The SV model (in its centered parameterization) is thus given through
+The SV model can thus be conveniently expressed in hierarchical form. In its centered parameterization, it is given through
 \begin{eqnarray}  
@@
 We refer to $\allpara = (\mupar, \phipar, \sigmapar)^\top$ as the vector of \emph{parameters}: the \emph{level} of log-variance $\mupar$, the \emph{persistence} of log-variance $\phipar$, and the \emph{volatility} of log-variance $\sigmapar$.
-The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in state equations (\ref{c2}) and (\ref{c3}) is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process).
-Note that the initial state $\hsv_0$ is distributed according to the stationary distribution of the autoregressive process of order one.
-
-\subsection{Prior distributions}
+The process $\hsvm=(\hsv_{0}, \hsv_{1}, \ldots,\hsv_{n})$ appearing in Equation~\ref{c2} and Equation~\ref{c3} is unobserved and usually interpreted as the latent time-varying \emph{volatility process} (more precisely, the log-variance process).
+Note that the initial state $\hsv_0$ appearing in Equation~\ref{c3} is distributed according to the stationary distribution of the autoregressive process of order one.
+
+\subsection{Prior distribution}
 \label{priors}
@@
 
-\subsection{MCMC Sampling}
+\subsection{MCMC sampling}
 
@@
 This is commonly referred to as a reparameterization issue and dealt with extensively in scholarly literature; for an early reference see, e.g., \cite{hil-smi:par}.
-In the case of the SV model, a move of this kind can be achieved by moving the level $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state equation (\ref{c2}) to the observation equation (\ref{c1}) through a simple reparameterization of the latent process $\hsvm$.
+For the model at hand, a move of this kind can be achieved by transferring the level of log-variance $\mupar$ and/or the volatility of log-variance $\sigmapar$ from the state process (Equations~\ref{c2}~and~\ref{c3}) to the observation process (Equation~\ref{c1}) through a simple reparameterization of $\hsvm$.
 However, in the case of the SV model, it turns out that no single superior parameterization exists.
@@
 All these steps are described in more detail below, along with a worked example.
-For a stepwise incorporation of SV effects into other MCMC samplers, see Section \ref{other}.
+For a stepwise incorporation of SV effects into other MCMC samplers, see Section~\ref{other}.
 
@@
 ret <- logret(exrates$USD, demean = TRUE)
-par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
-plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
-plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log returns")
-@
-
+par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, 0.5), mgp = c(2, 0.6, 0))
+plot(exrates$date, exrates$USD, type = "l",
+  main = "Price of 1 EUR in USD")
+plot(exrates$date[-1], ret, type = "l", main = "Demeaned log returns")
+@
 \begin{figure}[!ht]
@@
 The current default value for this parameter is $1000$ which has turned out to suffice in most situations.
-Nevertheless, the user is encouraged to check convergence carefully; see Section \ref{check} for more details.
+Nevertheless, the user is encouraged to check convergence carefully; see Section~\ref{check} for more details.
 The amount of iterations which are run after burn-in can be specified through the parameter \code{draws}, currently defaulting to $10\,000$.
@@
 res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
-                priorsigma = .1)
+  priorsigma = 0.1)
 @
@@
 if (usePreCalcResultsExtra) {
- load('vignette_sampling_draws1.RData')
+ load("vignette_sampling_draws1.RData")
 } else {
  res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
- #save(res, file = 'vignette_sampling_draws1.RData')
-}
-@
-
-\begin{CodeOutput}
+ #save(res, file = "vignette_sampling_draws1.RData")
+}
+@
+
+\begin{Soutput}
 Calling GIS_C MCMC sampler with 11000 iter. Series length is 3139.
@@
 
-Timing (elapsed): 17.935 seconds.
-613 iterations per second.
+Timing (elapsed): 12.92 seconds.
+851 iterations per second.
 
@@
 Summarizing posterior draws... Done!
-\end{CodeOutput}
+\end{Soutput}
 
@@
  \begin{enumerate}
-  \item[(1)] \code{volplot}: Plots posterior quantiles of the latent volatilities in percent, i.e., $100\exp(h_t/2)$, over time.
+  \item[(1)] \code{volplot}: Plots posterior quantiles of the latent volatilities in percent, i.e., empirical quantiles of the posterior distribution of $100\exp(h_t/2)$, over time.
 Apart from the mandatory \code{svsample}-object itself, this function takes several optional arguments.
@@
 
-\begin{figure}[!ht]
+\begin{figure}[!htp]
 \begin{center}
@@
 <<usd4, fig=TRUE, include=FALSE, width=11, height=7>>=
-res <- updatesummary(res, quantiles = c(.01, .1, .5, .9, .99))
+res <- updatesummary(res, quantiles = c(0.01, 0.1, 0.5, 0.9, 0.99))
 volplot(res, forecast = 100, dates = exrates$date[-1])
@@
 \begin{center}
- \includegraphics[width=.96\textwidth]{article-usd4}
+ \includegraphics[width=\textwidth]{article-usd4}
  \caption{As above, now with medians (black line) and 1\%/10\%/90\%/99\% quantiles (gray lines). This behavior can be achieved through a preceding call of \code{updatesummary}.}
@@
 \begin{center}
- \includegraphics[width=.96\textwidth]{article-usd5}
+ \includegraphics[width=\textwidth]{article-usd5}
  \caption{Trace plots of posterior draws for the parameters $\mupar, \phipar, \sigmapar$.}
@@
 n <- 1000
-beta.true <- c(.1, .5)
+beta.true <- c(0.1, 0.5)
 sigma.true <- 0.01
@@
 draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
-colnames(draws1) <- c(paste("beta", 0:(p-1), sep='_'), "sigma")
+colnames(draws1) <- c(paste("beta", 0:(p-1), sep = "_"), "sigma")
 sigma2draw <- 1
@@
 \item Run the main sampler: Iteratively draw from the conditional bivariate Gaussian distribution $\betab|\yb,\sigmaepsextrasuper2$, e.g., through the use of \pkg{mvtnorm} \citep{r:mvt}, and the conditional Inverse Gamma distribution $\sigmaepsextrasuper2|\yb,\betab$.
-<<eval=FALSE>>=
+<<>>=
 for (i in -(burnin-1):draws) {
@@
     sigma2draw * preCov))
- tmp <- C0 + .5 * (crossprod(y - X %*% betadraw) +
- 		  crossprod((betadraw - b0), B0inv) %*%
-		   (betadraw - b0))
+  tmp <- C0 + 0.5 * (crossprod(y - X %*% betadraw) +
+    crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
   sigma2draw <- 1 / rgamma(1, preDf, rate = tmp)
- if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
-}
-@
-<<echo=FALSE>>=
-for (i in -(burnin-1):draws) {
- betadraw <- as.numeric(mvtnorm::rmvnorm(1, preMean, sigma2draw*preCov))
- tmp <- C0 + .5*(crossprod(y - X%*%betadraw) +
- 		 crossprod((betadraw - b0), B0inv) %*% (betadraw - b0))
- sigma2draw <- 1/rgamma(1, preDf, rate = tmp)
- if (i > 0) draws1[i,] <- c(betadraw, sqrt(sigma2draw))
-}
-@
-\item Finally, summarize and visualize the posterior draws:
+  if (i > 0) draws1[i, ] <- c(betadraw, sqrt(sigma2draw))
+}
+@
+\item Calculate posterior means in order to obtain point estimates for the parameters:
 <<>>=
@@
 @
+
+\item Visualize the draws through \pkg{coda}'s native \code{plot} method:
 
@@
 \begin{center}
-\includegraphics[width=.98\textwidth]{article-homo}
+\includegraphics[width=\textwidth]{article-homo}
 \caption{Trace plots and kernel density estimates for some draws from the marginal posterior distributions in the regression model with heteroskedastic errors.
@@
 colnames(draws2) <- c("mu", "phi", "sigma",
-		      paste("beta", 0:(p-1), sep = '_'),
-                      paste("h", 1:n, sep='_'))
+  paste("beta", 0:(p-1), sep = "_"), paste("h", 1:n, sep = "_"))
 betadraw <- c(0, 0)
-svdraw <- list(para = c(mu = -10, phi = .9, sigma = .2),
+svdraw <- list(para = c(mu = -10, phi = 0.9, sigma = 0.2),
   latent = rep(-10, n))
@@
     priorphi = priorphi, priorsigma = priorsigma)
- normalizer <- as.numeric(exp(-latent(svdraw)/2))
+  normalizer <- as.numeric(exp(-latent(svdraw) / 2))
   Xnew <- X * normalizer
@@
 if (usePreCalcResults) {
- load('vignette_sampling_draws2.RData')
+ load("vignette_sampling_draws2.RData")
 } else {
@@
 @
-\item Finally, summarize and visualize (some) posterior draws:
+\item Finally, visualize and summarize (some) posterior draws:
+
 <<eval=FALSE>>=
-colMeans(draws2[,4:8])
-@
-
-<<echo=FALSE>>=
-colMeans(draws2selection)
-@
-
-<<eval=FALSE>>=
-plot(coda::mcmc(draws2[,4:7]))
+plot(coda::mcmc(draws2[, 4:7]))
 @
@@
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
-plot(coda::mcmc(draws2selection[,1:4]))
+plot(coda::mcmc(draws2selection[, 1:4]))
 @
@@
  \begin{center}
-\includegraphics[width=.95\textwidth]{article-hetero}
+\includegraphics[width=\textwidth]{article-hetero}
 \caption{Trace plots and kernel density estimates
@@
 \end{figure}
+
+<<eval=FALSE>>=
+colMeans(draws2[, 4:8])
+@
+
+<<echo=FALSE>>=
+colMeans(draws2selection)
+@
 
@@
 if (usePreCalcResults) {
- load('vignette_sampling_realworld.RData')
+ load("vignette_sampling_realworld.RData")
 } else {
@@
 ## initialize some space to hold results:
-realres <- vector('list', 3)
+realres <- vector("list", 3)
 
@@
 realres[[1]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
-colnames(realres[[1]]) <- c("mu", "phi", "sigma", paste("beta", 0:(p-1), sep='_'),
-			   paste("h", 1:n, sep='_'))
+colnames(realres[[1]]) <- c("mu", "phi", "sigma", paste("beta", 0:(p-1), sep="_"),
+			   paste("h", 1:n, sep="_"))
 
@@
 realres[[2]] <- matrix(as.numeric(NA), nrow=floor(draws/thinning), ncol=p+1)
-colnames(realres[[2]]) <- c(paste("beta", 0:(p-1), sep='_'), "sigma")
+colnames(realres[[2]]) <- c(paste("beta", 0:(p-1), sep="_"), "sigma")
 
@@
 realres[[3]] <- matrix(NA_real_, nrow = floor(draws/thinning), ncol = 3 + n + p)
-colnames(realres[[3]]) <- c("a_0", "a_1", "b_1", paste("beta", 0:(p-1), sep='_'),
-			   paste("sigma", 1:n, sep='_'))
+colnames(realres[[3]]) <- c("a_0", "a_1", "b_1", paste("beta", 0:(p-1), sep="_"),
+			   paste("sigma", 1:n, sep="_"))
 
@@
 
-realresselection <- vector('list', 3)
+realresselection <- vector("list", 3)
 realresselection[[1]] <- realres[[1]][,c('beta_0', 'beta_1')]
@@
 If Equation~\ref{preddensMC} is evaluated at $x=y_{t+1}^o$, we refer to it as the (one-step-ahead) \emph{predictive likelihood} at time $t+1$, denoted $\PL_{t+1}$.
-Moreover, draws from (\ref{preddens}) can be obtained by simulating values $\y_{t+1}^{(m)}$ from the distribution given through the density $p(\y_{t+1}|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]})$, the summands of Equation~\ref{preddensMC}.
+Moreover, draws from the posterior predictive distribution can be obtained by simulating values $\y_{t+1}^{(m)}$ from the distribution given through the density $p(\y_{t+1}|\yb^o_{[1:t]}, \unobs^{(m)}_{[1:t]})$, the summands of Equation~\ref{preddensMC}.
 
@@
 \]
-Thus, Algorithm \ref{predlikalg} provides a conceptually simple way of computing the marginal likelihood.
+Thus, Algorithm~\ref{predlikalg} provides a conceptually simple way of computing the marginal likelihood.
 However, these computations are quite costly in terms of CPU time, as they require an individual model fit for each of the $n$ points in time.
@@
 \begin{center}
- \includegraphics[width=\textwidth]{predlik1}
+ \includegraphics[width=.984\textwidth]{predlik1}
 \end{center}
@@
 
-The top panel of Figure~\ref{predlik3} visualizes the performance of the three models; it depicts observed regression residuals against their (one-day-ahead) predicted distributions.
-For the purpose of this image, observed regression residuals are simply defined as the deviation of the data from the median of the predicted distribution.
-It stands out that predictive quantiles arising from the models with heteroskedastic errors exhibit much more flexibility to adapt to the ``current state of the world'', while the simpler homoskedastic model barely exhibits this feature.
-
-Generally speaking, there is little difference in predicted residuals until the beginning of 2007.
-However, during the pre-crisis era (less volatility) and during the actual crisis (more volatility), the models catering for heteroskedasticity perform substantially better.
-It is interesting to see that predictive intervals for the models with SV errors and GARCH errors are very similar.
-
-In the bottom panel of Figure~\ref{predlik3}, the cumulative sums of the log predictive Bayes factors are displayed.
-The last points plotted equal to $176.52$ in the SV case and $166.37$ in the GARCH case; this provides overwhelming overall evidence in favor of a model catering for heteroskedasticity and ``decisive'' \citep{jef:the3ed} respectively ``very strong'' \citep{kas-raf:bay} evidence in favor of SV over vanilla GARCH(1,1) with a final cumulative predictive Bayes factor around $25\,000:1$.
-%and thus clearly rejects the assumption of homoskedastic error terms for this dataset.
-
 \begin{figure}[!htp]
 \begin{center}
- \includegraphics[width=\textwidth]{predlik3}
+ \includegraphics[width=.984\textwidth]{predlik3}
 \end{center}
@@
 
+The top panel of Figure~\ref{predlik3} visualizes the performance of the three models; it depicts observed regression residuals against their (one-day-ahead) predicted distributions.
+For the purpose of this image, observed regression residuals are simply defined as the deviation of the data from the median of the predicted distribution.
+It stands out that predictive quantiles arising from the models with heteroskedastic errors exhibit much more flexibility to adapt to the ``current state of the world'', while the simpler homoskedastic model barely exhibits this feature.
+
+Generally speaking, there is little difference in predicted residuals until the beginning of 2007.
+However, during the pre-crisis era (less volatility) and during the actual crisis (more volatility), the models catering for heteroskedasticity perform substantially better.
+It is interesting to see that predictive intervals for the models with SV errors and GARCH errors are very similar.
+
+In the bottom panel of Figure~\ref{predlik3}, the cumulative sums of the log predictive Bayes factors are displayed.
+The last points plotted equal to $176.52$ in the SV case and $166.37$ in the GARCH case; this provides overwhelming overall evidence in favor of a model catering for heteroskedasticity and ``decisive'' \citep{jef:the3ed} respectively ``very strong'' \citep{kas-raf:bay} evidence in favor of SV over vanilla GARCH(1,1) with a final cumulative predictive Bayes factor around $25\,000:1$.
+%and thus clearly rejects the assumption of homoskedastic error terms for this dataset.
+
 It is worth noting that for other exchange rates contained in \code{exrates}, a qualitatively similar picture emerges.
 For an overview of cumulative log predictive likelihoods, see Table~\ref{exrates}.
-Detailed results about these currencies are available from the author.
-
-As pointed out above, the analysis of log returns can be viewed upon as a special case of the analysis of log levels where $\beta_1 \equiv 1$ is fixed a priori.
-When doing so, evidence in favor of heteroskedasticity is again striking.
-Once more, the model with SV scores highest; its demeaned predictive quantiles are almost indistinguishable from the ``varying $\beta_1$ case''.
-Predictive quantiles of the model with GARCH residuals again resemble those of the model with SV residuals very closely.
-%Also, predictive performance is similar; the cumulative log predictive Bayes factors amount to $177.36$ for SV vs.\ homoskedastic and $166.64$ for GARCH vs.\ homoskedastic at $t=n$.
-The sum of log predictive likelihoods for $t \in \{1001,1002,\dots,n\}$ is given in the last line of Table~\ref{exrates}.
-The values are slightly but consistently lower than for the models where $\beta_1$ is estimated from the data.
-The display of predictive intervals for the log returns is practically identical to Figure~\ref{predlik3} and thus omitted here.
-%Figure~\ref{predlik3a} displays the predictive intervals for the log returns and the comparison through predictive likelihoods and predictive Bayes factors.
-It is practically identical to Figure~\ref{predlik3}.
-
-Concluding this somewhat anecdotal prediction exercise, we would like to note that for a thorough and more universal statement concerning the real-world applicability and predictive accuracy of \pkg{stochvol}, further studies with different datasets and a richer variety of competing models, potentially including realized volatility measures, are called for.
-Such a voluminous undertaking is, however, beyond the scope of this paper.
 
 \begin{table}[!ht]
+ \begin{small}
  \centering
@@
  \end{tabular}
- \caption{Final cumulative predictive log likelihoods for AR(1) models with different error assumptions for the logarithm of several EUR exchange rates included in \pkg{stochvol}'s dataset \code{exrates}.
+ \caption{Final cumulative predictive log likelihoods for AR(1) models with different error assumptions, applied to the logarithm of several EUR exchange rates.% included in \pkg{stochvol}'s dataset \code{exrates}.
  SV is strongly favored in all cases.}
  \label{exrates}
+ \end{small}
 \end{table}
+
+As pointed out above, the analysis of log returns can be viewed upon as a special case of the analysis of log levels where $\beta_1 \equiv 1$ is fixed a priori.
+When doing so, evidence in favor of heteroskedasticity is again striking.
+Once more, the model with SV scores highest; its demeaned predictive quantiles are almost indistinguishable from the ``varying $\beta_1$ case''.
+Predictive quantiles of the model with GARCH residuals again resemble those of the model with SV residuals very closely.
+%Also, predictive performance is similar; the cumulative log predictive Bayes factors amount to $177.36$ for SV vs.\ homoskedastic and $166.64$ for GARCH vs.\ homoskedastic at $t=n$.
+The sum of log predictive likelihoods for $t \in \{1001,1002,\dots,n\}$ is given in the last line of Table~\ref{exrates}.
+The values are slightly but consistently lower than for the models where $\beta_1$ is estimated from the data.
+The display of predictive intervals for the log returns is practically identical to Figure~\ref{predlik3} and thus omitted here.
+%Figure~\ref{predlik3a} displays the predictive intervals for the log returns and the comparison through predictive likelihoods and predictive Bayes factors.
+It is practically identical to Figure~\ref{predlik3}.
+
+Concluding this somewhat anecdotal prediction exercise, we would like to note that for a thorough and more universal statement concerning the real-world applicability and predictive accuracy of \pkg{stochvol}, further studies with different datasets and a richer variety of competing models, potentially including realized volatility measures, are called for.
+Such a voluminous undertaking is, however, beyond the scope of this paper.
 

### 2015-01-28 R/utilities_svdraws.R

@@ -239,7 +239,7 @@ plot.svdraws <- function(x, forecast = NULL, dates = NULL, show0 = FALSE,
  invisible(x)
 }
 
-predict.svdraws <- function(object, steps = 1, ...) {
+predict.svdraws <- function(object, steps = 1L, ...) {
  if (!is(object, "svdraws")) stop("Argument 'object' must be of class 'svdraws'.")
  steps <- as.integer(steps)
  if (steps < 1) stop("Argument 'steps' must be greater or equal to 1.")

### 2015-01-28 inst/doc/article.R

@@ -24,9 +24,10 @@ set.seed(123)
 library("stochvol")
 data("exrates")
 ret <- logret(exrates$USD, demean = TRUE)
-par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, .5), mgp = c(2, .6, 0))
-plot(exrates$date, exrates$USD, type = 'l', main = "Price of 1 EUR in USD")
-plot(exrates$date[-1], ret, type = 'l', main = "Demeaned log returns")
+par(mfrow = c(2, 1), mar = c(1.9, 1.9, 1.9, 0.5), mgp = c(2, 0.6, 0))
+plot(exrates$date, exrates$USD, type = "l",
+  main = "Price of 1 EUR in USD")
+plot(exrates$date[-1], ret, type = "l", main = "Demeaned log returns")
 
 
 ###################################################
@@ -38,25 +39,25 @@ plot(sim)
 
 
 ###################################################
-### code chunk number 5: article.Rnw:227-229 (eval = FALSE)
+### code chunk number 5: article.Rnw:233-235 (eval = FALSE)
 ###################################################
 ## res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
-##                 priorsigma = .1)
+##   priorsigma = 0.1)
 
 
 ###################################################
-### code chunk number 6: article.Rnw:232-238
+### code chunk number 6: article.Rnw:238-244
 ###################################################
 if (usePreCalcResultsExtra) {
- load('vignette_sampling_draws1.RData')
+ load("vignette_sampling_draws1.RData")
 } else {
  res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1), priorsigma = .1)
- #save(res, file = 'vignette_sampling_draws1.RData')
+ #save(res, file = "vignette_sampling_draws1.RData")
 }
 
 
 ###################################################
-### code chunk number 7: article.Rnw:260-261
+### code chunk number 7: article.Rnw:266-267
 ###################################################
 summary(res, showlatent = FALSE)
 
@@ -70,7 +71,7 @@ volplot(res, forecast = 100, dates = exrates$date[-1])
 ###################################################
 ### code chunk number 9: usd4
 ###################################################
-res <- updatesummary(res, quantiles = c(.01, .1, .5, .9, .99))
[…]

### 2015-05-15 inst/doc/heavytails.Rnw
version 1.1.0

%\VignetteIndexEntry{Heavy-Tailed Innovations in the R Package stochvol}
%\VignetteKeyword{Student's t distribution}
%\VignetteKeyword{data augmentation}
%\VignetteKeyword{EUR/CHF exrange rates}

\documentclass[article, nojss]{jss}

\usepackage{amsmath, amssymb}
\usepackage{bm}

\interfootnotelinepenalty=10000

\author{Gregor Kastner\\WU Vienna University of Economics and Business\\[1em]First version, May 15, 2015}
\title{Heavy-Tailed Innovations in the \proglang{R} Package \pkg{stochvol}}

%% for pretty printing and a nice hypersummary also set:
\Plainauthor{Gregor Kastner} %% comma-separated
\Plaintitle{Heavy-Tailed Innovations in the R Package stochvol} %% without formatting

%% an abstract and keywords
\Abstract{We document how sampling from a conditional Student's $t$ distribution is implemented in \pkg{stochvol}. Moreover, a simple example using EUR/CHF exchange rates illustrates how to use the augmented sampler. We conclude with results and implications.}
\Keywords{Student's $t$ distribution, data augmentation, EUR/CHF exchange rates}

%% The address of (at least) one author should be given
%% in the following format:
\Address{
  Gregor Kastner\\
  Institute for Statistics and Mathematics\\
  Department of Finance, Accounting and Statistics\\
  WU Vienna University of Economics and Business\\
  Welthandelsplatz 1, Building D4, Level 4\\
  1020 Vienna, Austria\\
  E-mail: \email{gregor.kastner@wu.ac.at}\\
  URL: \url{http://www.wu.ac.at/statmath/en/faculty_staff/faculty/gkastner/}
}

\input{macros.tex}
\graphicspath{{extrafig/}}
%% need no \usepackage{Sweave.sty}

\begin{document}

<<preliminaries, echo=FALSE, results=hide>>=
options(prompt = 'R> ', continue = '+  ')
@

\section*{Preface}
This note serves as a preliminary add-on to the more elaborate article ``Dealing with Stochastic Volatility in Time Series using the \proglang{R} package \pkg{stochvol}'' \citep{kas:dea}. It discusses and relaxes the restriction to conditionally normal errors in the vanilla stochastic volatility (SV) model.

\section{The SV model with Student's $t$ errors}

Several authors have suggested to use non-normal conditional innovation distributions for stochastic volatility modeling. Examples include the Student's $t$ distribution \citep{har-etal:mul}, the extended Generalized Inverse Gaussian \citep{sil-etal:ext}, (semi-)parametric innovations \citep{jen-mah:bay, del-gri:bay}, or the GH skew Student's $t$ distribution \citep{nak-omo:sto}. In the following, we describe how the estimation of the SV model with Student's $t$ errors is implemented in the \proglang{R} \citep{r:r} package \pkg{stochvol} \citep{r:sto}.

Let $\yb = (y_1, y_2, \dots, y_n)^\top$ be a vector of returns with mean zero. The SV model with Student's $t$ errors (in short SV-$t$) is given through
\begin{eqnarray}  
 \label{m1} y_{t}|\hsv_t,\df &\sim&  \Tdist{\df}{0,\exp \hsv_t}, \\
  \label{m2} \hsv_{t}|\hsv_{t-1},\mupar, \phipar, \sigmapar&\sim& \Normal{\mupar +  \phipar (\hsv_{t-1}- \mupar ),  \sigmapartwo},\\% \etat_{t},% \qquad  \etat_{t} \sim \Normal{0,1},
 \label{m3} \hsv_0|\mu,\phi,\sigmapar &\sim& \Normal{\mu,\sigmapartwo/(1-\phi^2)},
\end{eqnarray}
i.e., conditionally on $\hsv_t$, the data is assumed to follow a zero-mean non-standardized Student's $t$ distribution with $\df$ degrees of freedom and variance $(\df \exp \hsv_t)/(\df-2)$ for $\df > 2$. Following \cite{chi-etal:mar}, we assume that a priori the degrees of freedom parameter $\df \sim \Uniform{a,b}$, i.e., follows a uniform distribution with support on the real interval $(a,b)$. All other prior components are chosen as in \cite{kas:dea}.

\section{Usage}
Estimating a stochastic volatility model with conditional $t$ errors via \pkg{stochvol} is very similar to estimating a model with standard Gaussian errors, differing only through specifying a non-\code{NA} argument \code{priornu}. This triggers the sampler specified in Section~\ref{sec:alg}. To provide an example, we investigate the historical daily EUR/CHF exchange rates and display these in Figure~\ref{fig:data}.

<<data, fig=TRUE, include=FALSE, width=10, height=6, echo=TRUE>>=
library(stochvol)
data(exrates)
par(mfrow = c(2, 1), mar = c(1.7, 1.7, 1.7, 0.1), mgp = c(1.6, 0.6, 0))
plot(exrates$date, exrates$CHF, type = 'l', main = 'Price of 1 EUR in CHF')
dat <- logret(exrates$CHF, demean = TRUE)
plot(exrates$date[-1], dat, type = 'l', main = 'Demeaned log returns')
@

\begin{figure}[!ht]
\begin{center}
 \includegraphics[width=\textwidth]{heavytails-data}
 \caption{Levels and demeaned log returns of EUR in CHF.}
 \label{fig:data}
\end{center}
\end{figure}

First, we fit a standard SV model to the data. Then, by specifying the argument \code{priornu} (a two-element vector containing the lower and upper bounds of the uniform prior for $\nu$), we can trigger the sampler allowing for heavy-tailed conditional innovations.

<<normalerr, fig=TRUE, include=FALSE, width=10, height=6, echo=TRUE>>=
res <- svsample(dat, priormu = c(-12, 1), priorphi = c(20, 1.1),
  priorsigma = 0.1)
plot(res, showobs = FALSE)
@

<<terr, fig=TRUE, include=FALSE, width=10, height=7, echo=TRUE>>=
rest <- svsample(dat, priormu = c(-12, 1), priorphi = c(20, 1.1),
  priorsigma = 0.1, priornu = c(2, 100))
plot(rest, showobs = FALSE)
@

Results are displayed in Figure \ref{fig:res}, where the top panel (rows 1 to 3) corresponds to the output from the standard SV model and the bottom panel (rows 4 to 7) contains the output from the SV-$t$ model. Row~4 depicts $\exp(h_t/2)$ for $t \in \{1,\dots,n\}$; row~5 shows the time varying standard deviations given through $\sqrt{\nu/(\nu-2)}\exp(h_t/2)$ for $t \in \{1,\dots,n\}$; row~6 portrays trace plots and row~7 outlines the corresponding smoothed kernel density estimates for the four parameters $\mu$, $\phi$, $\sigma$, and $\nu$. It is worth noting that $\nu$ is estimated to lie between $6$ and $18$ with high posterior probability, indicating evidence for the presence of heavy tails even after catering for stochastic volatility. The extra flexibility of the SV-$t$ sampler seems to allow for increased persistence $\phi$ and smaller variance of log-volatility $\sigma^2$, resulting in smoother time-varying volatility estimates.

\begin{figure}[p]
\begin{center}
 \includegraphics[width=\textwidth]{heavytails-normalerr}
 \vspace{2em}

 \includegraphics[width=\textwidth]{heavytails-terr}
 \caption{Standard output of the \code{plot} method when applied to an \code{svdraws} object containing posterior draws from an SV model with Gaussian errors (top) and Student's $t$ errors (bottom).}
 \label{fig:res}
\end{center}
\end{figure}

We investigate one-day-ahead out-of-sample predictive performance of an AR(1) model with (a) homoskedastic, (b) SV, (c) SV-$t$, (d) GARCH(1,1) errors, applied to the raw exchange rate data. Details about this procedure are provided in Chapter~5 of \cite{kas:dea}. The results, summarized in Figure~\ref{fig2}, speak in favor of the SV-$t$ model for this dataset.

\begin{figure}[p]
\begin{center}
 \includegraphics[width=\textwidth]{predlik_terr}
 \caption{Log predictive one-day-ahead Bayes factors in favor of SV, SV-$t$, and GARCH errors over the homoskedastic model. The final log predictive Bayes factors aggregate to 1107.44 (SV), 1115.36 (SV-$t$), and 1033.53 (GARCH), respectively, thus providing strong evidence for the SV-$t$ model.}
 \label{fig2}
\end{center}
\end{figure}

\section{Necessary modifications in the sampling scheme}
\label{sec:alg}

The Student's $t$ distribution appearing in Equation \ref{m1} can be conveniently expressed as a scale mixture of normal distributions,
\begin{eqnarray*}
y_{t}|\hsv_t,\taug_t &\sim&  \Normal{0, \taug_t \exp{\hsv_{t}}},%\error_{t},% \qquad \error_{t} \sim \Normal{0,1},
 \\
 \taug_t|\df &\sim& \Gammainv{\df/2, \df/2},
\end{eqnarray*}
where $\Normal{\mu, \sigmapartwo}$ denotes the normal distribution with mean $\mu$ and variance $\sigmapartwo$, and $\Gammainv{a,b}$ denotes the inverse gamma distribution with shape and scale parameters $a$ and $b$, respectively.

Treating $\taugm = (\taug_1, \dots, \taug_n)^\top$ as latent data and letting $\tilde y_t = y_t/\sqrt{\taug_t}$ for $t \in \{1,\dots,n\},$ we have
\[
\tilde y_t|\hsv_t \sim \Normal{0, \exp{h_t}},
\]
and the AWOL sampler described in \cite{kas-fru:anc} can directly be applied to the transformed data. To obtain draws from the newly introduced variables $\taugm$ and $\df$, two additional steps are required.

\subsection{Sampling the auxiliary variables}

It is easy to see that the marginal posterior is given through
\[
\taug_t|y_t,\hsv_t,\df \sim \Gammainv{\frac{\df + 1}{2},\frac{\df + y_t^2\exp(-\hsv_t)}{2}},
\]
independently for each $t \in \{1, \dots, n\}$. Obtaining draws from this distribution is straightforward.

\subsection{Sampling the degrees of freedom parameter}

The full conditional posterior of the degrees of freedom parameter, $\df|\cdot$, only depends on $\taugm$. Its density is given through the product of $n$ univariate truncated inverse gamma densities which may be written as
\begin{eqnarray}
 \label{dfdens}
p(\df|\cdot) = p(\df|\taugm) \propto
\left(\frac{\df}{2}\right)^{n\df/2}
\Gamma\!\left(\frac{\df}{2}\right)^{-n}
\left(\prod_{t=1}^n \taug_t\right)^{-\df/2}
\exp\left\{-\frac{\df}{2}\sum_{t=1}^n\frac{1}{\taug_t}\right\}
\end{eqnarray}
for $\df \in (a,b)$ and zero elsewhere.

For obtaining draws from this distribution, we use an independence Metropolis-Hastings update. We follow \cite{chi-gre:bay}, who introduced the idea of specifying an independence proposal through numerical maximization of the log-density. For the problem at hand, we consequently aim for optimizing
\begin{eqnarray}
 \label{logdfdens}
\log p(\df|\taugm) = \frac{n\df}{2}\log(\df/2) -
n\log\Gamma(\df/2)-
\frac{\df}{2}\sum_{t=1}^n\!\left(\log\taug_t+\frac{1}{\taug_t}\right) + C,
\end{eqnarray}
with first and second derivatives given through
\begin{eqnarray}
 \label{d1}
 \frac{\partial \log p(\df|\taugm)}{\partial \df} &=&
 \frac{n}{2}\left(1 + \log(\df/2) -\psi^{(0)}(\nu/2)\right) -
 \frac{1}{2}\sum_{t=1}^n\!\left(\log\taug_t + \frac{1}{\taug_t}\right),\\
\label{d2}
\frac{\partial^2 \log p(\df|\taugm)}{\partial \df^2} &=&
\frac{n}{2\df} - \frac{n}{4}\psi^{(1)}(\nu/2),
\end{eqnarray}
where $\psi^{(m)}$ denotes the polygamma function of order $m$. Using the above, it is easy to numerically find
\begin{eqnarray*}
\hat \df &=& \arg\max_{\df} \log p(\df|\taugm),\\
B_{\hat \df} &=& \left.-1\middle/\frac{\partial^2 \log p(\df|\taugm)}{\partial \df^2}\right|_{\df=\hat\df},
\end{eqnarray*}
and a proposal candidate $\df_\text{prop}$ may be drawn from a normal distribution with mean $\hat\df$ and variance $B_{\hat \df}$ (the Laplace approximation). Letting $\phi(x| \hat\df, B_{\hat \df})$ denote the corresponding density function, the acceptance probability is equal to $\min\{1,R\}$ with
\[
R = \frac{p(\df_\text{prop}|\taugm)}{p(\df_\text{old}|\taugm)} \times
\frac{\phi(\df_\text{old}| \hat \df, B_{\hat \df})}{\phi(\df_\text{prop}| \hat \df, B_{\hat \df})}.
\
\]

\section{Conclusion}
We have shown how a simply data augmentation trick can be utilized to generalize the core sampler in \pkg{stochvol} in order to cater for potentially heavier-tailed innovation distributions. However, several caveats are called for:
\begin{itemize}
 \item Even though the uniform prior for $\nu$ has been used widely, more robust alternatives are probably preferred, cf.\ \cite{fru-pay:bay} and the references therein.
 \item Sampling the degrees of freedom parameter $\nu$ conditionally on $\bm{\tau}$ can be very inefficient if $\nu$ becomes large (and thus the plain vanilla SV model suffices). We recommend to resort to the original SV sampler in this case.
 \item Leaving aside the additional computational burden, it is trivial to incorporate this extension into samplers employing \pkg{stochvol} as part of a larger MCMC scheme \citep[e.g.][]{hub:den, kas-etal:ana, dov-etal:doe}. Nevertheless, at the current stage of development, this should be conducted with caution by carefully investigating the convergence of the posterior draws.
\end{itemize}

\section*{Acknowledgments}
The author would like to thank the attendants of the Spring 2015 Brown Bag Seminar of the Institute for Statistics and Mathematics, WU Vienna University of Economics and Business, in particular Sylvia Fr\"{u}hwirth-Schnatter, Mark Jensen, and Kurt Hornik, for stimulating comments and suggestions.

\bibliography{mybib}

\end{document}

### 2015-05-15 R/simulation.R

@@ -1,4 +1,4 @@
-svsim <- function(len, mu = -10, phi = 0.98, sigma = 0.2) {
+svsim <- function(len, mu = -10, phi = 0.98, sigma = 0.2, nu = Inf) {
  
  # Some error checking
  if (any(is.na(len)) | !is.numeric(len) | length(len) != 1 | any(len < 1)) {
@@ -19,19 +19,28 @@ if (!is.numeric(sigma) | length(sigma) != 1 | sigma <= 0) {
   stop("Argument 'sigma' (volatility of latent variable) must be a single number > 0.")
  }
 
+if (!is.numeric(nu) || length(nu) != 1 || nu <= 2) {
+ stop("Argument 'nu' (degrees of freedom for the conditional error) must be a single number > 2.")
+}
+
  h <- rep(as.numeric(NA), len)
  h0 <- rnorm(1, mean=mu, sd=sigma/sqrt(1-phi^2))
- nu <- rnorm(len)
+ innov <- rnorm(len)
 
  # simulate w/ simple loop
- h[1] <- mu + phi*(h0-mu) + sigma*nu[1]
- for (i in seq(2, len = len-1)) h[i] <- mu + phi*(h[i-1]-mu) + sigma*nu[i]
+ h[1] <- mu + phi*(h0-mu) + sigma*innov[1]
+ for (i in seq(2, len = len-1)) h[i] <- mu + phi*(h[i-1]-mu) + sigma*innov[i]
 
- y <- rnorm(len, mean=0, sd=sqrt(exp(h)))  # "log-returns"
+ if (is.infinite(nu)) {
+  y <- exp(h / 2) * rnorm(len)  # "log-returns"
+ } else {
+  y <- exp(h / 2) * rt(len, df = nu)  # "log-returns"
+ }
  ret <- list(y = y, vol = exp(h/2), vol0 = exp(h0/2),
 	     para = list(mu = mu,
 			 phi = phi,
 			 sigma = sigma))
+ if (is.finite(nu)) ret$para$nu <- nu
  class(ret) <- "svsim"
  ret
 }
@@ -39,11 +48,13 @@ if (!is.numeric(sigma) | length(sigma) != 1 | sigma <= 0) {
 print.svsim <- function(x, ...) {
  cat("\nSimulated time series consisting of", length(x$y), "observations.\n
 Parameters: level of latent variable                  mu =", x$para$mu, "
-	    persistence of latent variable           phi =", x$para$phi, "
[…]

### 2015-05-15 R/utilities_svdraws.R

@@ -30,6 +30,9 @@ runtime <- function(x) {
 
 updatesummary <- function(x, quantiles = c(.05, .5, .95), esspara = TRUE, esslatent = FALSE) {
  
+ # NEW: Check if conditional t errors are used
+ if (ncol(x$para) == 4) terr <- TRUE else terr <- FALSE
+
  summaryfunction <- function(x, quants = quantiles, ess = TRUE) {
   if (ess) {
    c(mean = mean(x), sd = sd(x), quantile(x, quantiles),
@@ -51,6 +54,12 @@ updatesummary <- function(x, quantiles = c(.05, .5, .95), esspara = TRUE, esslat
  res$latent <- cbind(res$latent, "sd(exp(h_t/2))" = apply(tmp, 2, sd))
  
  res$latent0 <- c(summaryfunction(x$latent0, ess = esslatent), "mean(exp(h_t/2))" = mean(exp(x$latent0/2)), "sd(exp(h_t/2))" = sd(exp(x$latent0/2)))
+
+ if (terr && x$thinning$para == x$thinning$latent) {
+  tmp <- sqrt((exp(x$latent) * x$para[,"nu"]) / (x$para[,"nu"] - 2))
+  res$sd <- t(apply(tmp, 2, summaryfunction, ess = esslatent))
+  rownames(res$sd) <- gsub("h", "sd", rownames(res$latent))
+ }
  
  x$summary <- res
  x
@@ -73,6 +82,8 @@ Prior distributions:
 mu        ~ Normal(mean = ", x$priors$mu[1], ", sd = ", x$priors$mu[2], ")
 (phi+1)/2 ~ Beta(a0 = ", x$priors$phi[1], ", b0 = ", x$priors$phi[2], ")
 sigma^2   ~ ", x$priors$sigma, " * Chisq(df = 1)\n", sep="")
+
+ if (length(x$priors) == 4) cat("nu        ~ Unif(lower = ", x$priors$nu[1], ", upper = ", x$priors$nu[2], ")\n", sep="")
 
  if (exists("para", x)) {
   cat("\nPosterior draws of parameters (thinning = ", x$mcp[3], "):\n", sep='')
@@ -101,142 +112,29 @@ print.svdraws <- function(x, showpara = TRUE, showlatent = TRUE, ...) {
  invisible(x)
 }
 
[…]

### 2015-05-15 R/wrappers.R

@@ -1,6 +1,6 @@
 # R wrapper function for the main MCMC loop
 
-svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), priorphi = c(5, 1.5), priorsigma = 1, thinpara = 1, thinlatent = 1, thintime = 1, quiet = FALSE, startpara, startlatent, expert, ...) {
+svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), priorphi = c(5, 1.5), priorsigma = 1, priornu = NA, thinpara = 1, thinlatent = 1, thintime = 1, quiet = FALSE, startpara, startlatent, expert, ...) {
  
  # Some error checking for y
  if (is(y, "svsim")) {
@@ -42,7 +42,15 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
  if (!is.numeric(priorsigma) | length(priorsigma) != 1 | priorsigma <= 0) {
   stop("Argument 'priorsigma' (scaling of the chi-squared(df = 1) prior for sigma^2) must be a single number > 0.")
  }
- 
+
+ if (any(is.na(priornu))) {
+  priornu <- NA
+ } else {
+  if (!is.numeric(priornu) || length(priornu) != 2 || priornu[1] >= priornu[2] || priornu[1] < 0) {
+   stop("If not NA, argument 'priornu' (lower and upper bounds for the uniform prior for the df) must be numeric and of length 2. Moreover, 0 <= priornu[1] < priornu[2].")
+  }
+ }
+
  # Some error checking for thinpara
  if (!is.numeric(thinpara) | thinpara < 1) {
   stop("Argument 'thinpara' (thinning parameter for mu, phi, and sigma) must be a single number >= 1.")
@@ -160,10 +168,14 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
 
  # Some input checking for startpara
  if (missing(startpara)) {
-  startpara <- list(mu = -10, phi = .9, sigma = .3) 
+  if (any(is.na(priornu))) {
+   startpara <- list(mu = -10, phi = .9, sigma = .3)
+  } else {
+   startpara <- list(mu = -10, phi = .9, sigma = .3, nu = mean(priornu))
+  }
  } else {
   if (!is.list(startpara))
[…]

### 2015-05-15 inst/doc/article.R

@@ -1,4 +1,5 @@
 ### R code from vignette source 'article.Rnw'
+### Encoding: UTF-8
 
 ###################################################
 ### code chunk number 1: article.Rnw:1-7
@@ -39,14 +40,14 @@ plot(sim)
 
 
 ###################################################
-### code chunk number 5: article.Rnw:233-235 (eval = FALSE)
+### code chunk number 5: article.Rnw:240-242 (eval = FALSE)
 ###################################################
 ## res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
 ##   priorsigma = 0.1)
 
 
 ###################################################
-### code chunk number 6: article.Rnw:238-244
+### code chunk number 6: article.Rnw:245-251
 ###################################################
 if (usePreCalcResultsExtra) {
  load("vignette_sampling_draws1.RData")
@@ -57,7 +58,7 @@ if (usePreCalcResultsExtra) {
 
 
 ###################################################
-### code chunk number 7: article.Rnw:266-267
+### code chunk number 7: article.Rnw:273-274
 ###################################################
 summary(res, showlatent = FALSE)
 
@@ -103,7 +104,7 @@ plot(myresid, ret)
 
 
 ###################################################
-### code chunk number 14: article.Rnw:398-404
+### code chunk number 14: article.Rnw:405-411
 ###################################################
 set.seed(123456)
 n <- 1000
@@ -114,7 +115,7 @@ y <- rnorm(n, X %*% beta.true, sigma.true)
 
 
 ###################################################
-### code chunk number 15: article.Rnw:408-414
+### code chunk number 15: article.Rnw:415-421
 ###################################################
 burnin <- 100
 draws <- 5000
@@ -125,7 +126,7 @@ C0 <- 0.001
 
 
 ###################################################
-### code chunk number 16: article.Rnw:417-421
+### code chunk number 16: article.Rnw:424-428
 ###################################################
 p <- ncol(X)
 preCov <- solve(crossprod(X) + B0inv)
@@ -134,7 +135,7 @@ preDf <- c0 + n/2 + p/2
 
 
 ###################################################
-### code chunk number 17: article.Rnw:424-427
+### code chunk number 17: article.Rnw:431-434
 ###################################################
 draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
 colnames(draws1) <- c(paste("beta", 0:(p-1), sep = "_"), "sigma")
@@ -142,7 +143,7 @@ sigma2draw <- 1
 
 
[…]

### 2015-05-15 inst/include/update.h

@@ -3,13 +3,17 @@
 #if defined(COMPILING_STOCHVOL)
 
 // just declare, will be defined in sampler.cpp
-// a single MCMC update:
+// a single MCMC update (normal errors):
 void update(const Rcpp::NumericVector &, double *, double *, double &,
             double *, int *, const bool, const double,
 	    const double, const double, const double, const double,
 	    const double, const double, const double, const double,
 	    const bool, const bool, const double, const int, const int,
 	    const bool);
+
+// a single MCMC update (t errors):
+void update_terr(const Rcpp::NumericVector &,
+                 double *, double &, const double, const double);
 
 #else
 

### 2015-05-15 src/densities.h

@@ -37,9 +37,25 @@ inline double logacceptrateRW(double xnew, double xold, double Bsigma, int T, do
 
 // proportion of two beta-distributions with same parameters
 // (evaluated at two different points)
-
 inline double propBeta(double x, double y, double a, double b) {
  return pow(x/y, a-1)*pow((1-x)/(1-y), b-1);
 }
+
+// full conditional non-normalized posterior log-density of the
+// degrees of freedom parameter nu
+inline double logdnu(double nu, double sumtau, int n) {
+ return .5 * nu * (n*log(.5*nu) - sumtau) - n*lgamma(.5*nu);
+}
+
+// first derivative of logdnu
+inline double dlogdnu(double nu, double sumtau, int n) {
+ return .5 * (n * ( 1 + log(.5*nu) - Rf_digamma(.5*nu)) - sumtau);
+}
+
+// second derivative of logdnu
+inline double ddlogdnu(double nu, int n) {
+ return .25 * n * (2/nu - Rf_trigamma(.5*nu));
+}
+
 
 #endif

### 2015-05-15 src/progutils.cpp

@@ -2,14 +2,17 @@
 
 /* Contains the following code modules:
 
-   a) Some helper functions such as progress bar tools and return value
+   a) some helper functions such as progress bar tools and return value
       prettifier
 
-   b) Some functions related to the Cholesky decomposition used for 
+   b) some functions related to the Cholesky decomposition used for 
       sampling AWOL and efficiently solving the systems of linear
       equations
 
    c) function for inverse transform sampling
+
+   d) a very basic Newton-Raphson algorithm for finding the root
+      of dlogdnu (defined in densities.h)
 */
 
 // a)
@@ -18,12 +21,17 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
                    const Rcpp::NumericVector & phi,
 		   const Rcpp::NumericVector & sigma,
 		   const Rcpp::NumericMatrix & hstore,
-		   const Rcpp::NumericVector & h0store) {
-
- Rcpp::NumericMatrix res(mu.length(), 3); 
+		   const Rcpp::NumericVector & h0store,
+		   const Rcpp::NumericVector & nustore,
+		   const Rcpp::NumericMatrix & taustore) {
+ int paracols;
+ if (nustore.size() > 0) paracols = 4; else paracols = 3;
+ 
+ Rcpp::NumericMatrix res(mu.length(), paracols); 
  res(Rcpp::_,0) = mu;
  res(Rcpp::_,1) = phi;
  res(Rcpp::_,2) = sigma;
+ if (nustore.size() > 0) res(Rcpp::_,3) = nustore;
 
 /* res.attr("dimnames") = Rcpp::List::create(
    R_NilValue, 
@@ -33,6 +41,7 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
    Rcpp::_["para"] = res,
    Rcpp::_["latent"] = hstore,
    Rcpp::_["latent0"] = h0store);
+//   Rcpp::_["tau"] = taustore);
 
  return val;
 }
@@ -133,3 +142,27 @@ void invTransformSampling(const double * const mixprob, int * r, int T) {
  r[j] = index;
  }
 }
+
+// d)
+// find the root of a function (Newton-Raphson)
+double newtonRaphson(double startval, double sumtau, int n,
+                     double lower, double upper, double tol,
+		     int maxiter) {
+ double x = startval;
+ double error = R_PosInf;
[…]

### 2015-05-15 src/progutils.h

@@ -3,17 +3,21 @@
 
 /* Contains the following code modules:
 
-   a) Some helper functions such as progress bar tools and return value
+   a) some helper functions such as progress bar tools and return value
       prettifier
 
-   b) Some functions related to the Cholesky decomposition used for 
+   b) some functions related to the Cholesky decomposition used for 
       sampling AWOL and efficiently solving the systems of linear
       equations
 
    c) function for inverse transform sampling
+
+   d) a very basic Newton-Raphson algorithm for finding the root
+      of dlogdnu (defined in densities.h)
 */
 
 #include <Rcpp.h>
+#include "densities.h"
 
 // a)
 // Sums up results and prepares return value
@@ -21,7 +25,9 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
                    const Rcpp::NumericVector & phi,
 		   const Rcpp::NumericVector & sigma,
 		   const Rcpp::NumericMatrix & hstore,
-		   const Rcpp::NumericVector & h0store);
+		   const Rcpp::NumericVector & h0store,
+		   const Rcpp::NumericVector & nustore,
+		   const Rcpp::NumericMatrix & taustore);
 
 // sets up the progress bar
 int progressbar_init(int N);
@@ -37,7 +43,9 @@ void progressbar_finish(int N);
 
 // to store (some) values of h
 inline void store_h(double * h, double * hstore, int timethin,
-                    int hstorelength, double h0, double * h0store,
+                    int hstorelength,
+                    double * tau, double * taustore, 
+		    double h0, double * h0store,
 		    Rcpp::NumericVector curpara,
 		    bool centered_baseline) {
 
@@ -48,7 +56,7 @@ inline void store_h(double * h, double * hstore, int timethin,
   for (int j = 0; j < hstorelength; j++) hstore[j] = curpara[0] + curpara[2]*h[timethin*j];
   *h0store = curpara[0] + curpara[2]*h0;
  }
-
+ //for (int j = 0; j < hstorelength; j++) taustore[j] = tau[timethin*j];
 }
 
 // b)
@@ -65,4 +73,9 @@ void backwardAlg(const Rcpp::NumericVector & chol_diag, const Rcpp::NumericVecto
 // draws length(r) RVs, expects the non-normalized CDF mixprob
 void invTransformSampling(const double * const, int * r, int T);
 
+// d)
+// find the root of a function (Newton-Raphson)
[…]

### 2015-05-15 src/sampler.cpp

@@ -19,10 +19,10 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
   const SEXP MHsteps_in, const SEXP B011_in, const SEXP B022_in,
   const SEXP mhcontrol_in, const SEXP gammaprior_in,
   const SEXP truncnormal_in, const SEXP offset_in,
-  const SEXP dontupdatemu_in) {
+  const SEXP dontupdatemu_in, const SEXP priordf_in) {
 
  //RNGScope scope;       // just in case no seed has been set at R level
- GetRNGstate(); // "by hand" because RNGScope isn't save if return
+ GetRNGstate(); // "by hand" because RNGScope isn't safe if return
                 // variables are declared afterwards
 
  // convert SEXP into Rcpp-structures (no copy at this point)
@@ -64,13 +64,19 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
  // offset:
  double offset		= as<double>(offset_in);
 
+ // t-errors:
+ bool terr;
+ NumericVector priordf(priordf_in);
+
+ if (ISNA(priordf(0))) terr = false; else terr = true;
+
  // indicator telling us whether assume mu = 0 fixed
  bool dontupdatemu      = as<bool>(dontupdatemu_in);
 
  // moment-matched IG-prior
  double c0 = 2.5;
  double C0 = 1.5*Bsigma;
- 
+
  // pre-calculation of a posterior parameter
  double cT = -10000; 
  if (Gammaprior) {  
@@ -105,10 +111,19 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
 
  NumericMatrix mixprob(10, T);  // mixture probabilities
  NumericVector data = log(y*y + offset);  // commonly used transformation
+ NumericVector datastand = clone(data);  // standardized "data" (different for t-errors)
  NumericVector curpara(3);  // holds mu, phi, sigma in every iteration
  curpara[0] = mu[0];
  curpara[1] = phi[0];
  curpara[2] = 1/sqrt(sigma2inv[0]);
+ 
+ double nu;
+ if (terr) nu = as<double>(startpara["nu"]);
+ NumericVector tau(T, 1.);
+
+ NumericVector nustore(terr * (N+1), nu);
[…]

### 2015-05-15 src/sampler.h

@@ -18,7 +18,7 @@ RcppExport SEXP sampler(const SEXP, const SEXP, const SEXP,
   const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
   const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
   const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
-  const SEXP, const SEXP, const SEXP, const SEXP);
+  const SEXP, const SEXP, const SEXP, const SEXP, const SEXP);
 
 // Step (b): sample mu, phi, sigma - __CENTERED__ version:
 Rcpp::NumericVector regressionCentered(

### 2015-05-23 inst/doc/article.R
version 1.1.1

@@ -165,13 +165,13 @@ colMeans(draws1)
 ### code chunk number 20: homo
 ###################################################
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
-plot(coda::mcmc(draws1))
+plot(coda::mcmc(draws1), show.obs = FALSE)
 
 
 ###################################################
 ### code chunk number 21: article.Rnw:459-460 (eval = FALSE)
 ###################################################
-## plot(coda::mcmc(draws1))
+## plot(coda::mcmc(draws1), show.obs = FALSE)
 
 
 ###################################################
@@ -266,14 +266,14 @@ for (i in -(burnin-1):draws) {
 ###################################################
 ### code chunk number 27: article.Rnw:566-567 (eval = FALSE)
 ###################################################
-## plot(coda::mcmc(draws2[, 4:7]))
+## plot(coda::mcmc(draws2[, 4:7]), show.obs = FALSE)
 
 
 ###################################################
 ### code chunk number 28: hetero
 ###################################################
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
-plot(coda::mcmc(draws2selection[, 1:4]))
+plot(coda::mcmc(draws2selection[, 1:4]), show.obs = FALSE)
 
 
 ###################################################

### 2015-05-23 src/progutils.cpp

@@ -152,7 +152,7 @@ double newtonRaphson(double startval, double sumtau, int n,
  double error = R_PosInf;
  double xnew;
  bool converged = false;
- for (int i; i < maxiter; i++) {
+ for (int i = 0; i < maxiter; i++) {
   xnew = x - dlogdnu(x, sumtau, n)/ddlogdnu(x, n);
   if (xnew > upper) xnew = upper; else if (xnew < lower) xnew = lower;
   error = fabs(xnew - x);

### 2015-07-15 inst/doc/article.Rnw
version 1.2.0

@@
 %\VignetteIndexEntry{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol}
-%\VignetteKeyword{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
+
+%\VignetteKeyword{Bayesian inference}
+%\VignetteKeyword{Markov chain Monte Carlo (MCMC)}
+%\VignetteKeyword{heteroskedasticity}
+%\VignetteKeyword{SV}
+%\VignetteKeyword{GARCH}
+%\VignetteKeyword{forecasting}
+%\VignetteKeyword{predictive density}
+%\VignetteKeyword{predictive Bayes factor}
+%\VignetteKeyword{exchange rates}
 
@@
 \title{Dealing with Stochastic Volatility in Time Series Using the \proglang{R} Package \pkg{stochvol}}
-%\Shorttitle{The \proglang{R} package \pkg{stochvol} (DRAFT VERSION)} %% a short title (if necessary)
 \Plaintitle{Dealing with Stochastic Volatility in Time Series Using the R Package stochvol} %% without formatting
@@
 }
-\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
-%\Plainkeywords{Bayesian inference, Markov chain Monte Carlo (MCMC), auxiliary mixture sampling, ancillarity-sufficiency interweaving strategy (ASIS), state-space model, heteroskedasticity, financial time series}
+\Keywords{Bayesian inference, Markov chain Monte Carlo (MCMC), heteroskedasticity, SV, GARCH, forecasting, predictive density, predictive Bayes factor, exchange rates}
 %% at least one keyword must be supplied
@@
   WU Vienna University of Economics and Business\\
-%  Building D4, Level 4\\
-  Welthandelsplatz 1\\
+  Welthandelsplatz 1 / D4 / level 4\\
   1020 Vienna, Austria\\
@@
 This vignette corresponds to a forthcoming article of the same name in the Journal of Statistical Software.
-At the time of this writing, the vignette and the published article coincide; nevertheless, this version might receive minor updates over time.
+At the time of this writing, the vignette and the published article coincide; nevertheless, this version might receive minor updates as time goes by.
 To cite, please use \cite{kas:dea}.
-Further information about citing \pkg{stochvol} can be obtained in \proglang{R} by installing the package, e.g., through \code{install.packages("stochvol")}, and typing \code{citation("stochvol")}.
+Further information about citing \pkg{stochvol} can be obtained in \proglang{R} by installing the package, e.g., through \code{install.packages("stochvol")}, and calling \code{citation("stochvol")}.
 
@@
 par(mfrow = c(1, 3))
-paradensplot(res)
+paradensplot(res, showobs = FALSE)
 @
@@
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
-plot(coda::mcmc(draws1))
+plot(coda::mcmc(draws1), show.obs = FALSE)
 @
@@
 <<eval=FALSE>>=
-plot(coda::mcmc(draws1))
+plot(coda::mcmc(draws1), show.obs = FALSE)
 @
@@
 <<eval=FALSE>>=
-plot(coda::mcmc(draws2[, 4:7]))
+plot(coda::mcmc(draws2[, 4:7]), show.obs = FALSE)
 @
@@
 par(mar = c(3.1, 1.8, 1.9, .5), mgp = c(1.8, .6, 0))
-plot(coda::mcmc(draws2selection[, 1:4]))
+plot(coda::mcmc(draws2selection[, 1:4]), show.obs = FALSE)
 @
@@
  \end{tabular}
- \caption{Final cumulative predictive log likelihoods for AR(1) models with different error assumptions, applied to the logarithm of several EUR exchange rates.% included in \pkg{stochvol}'s dataset \code{exrates}.
+ \caption{Final cumulative predictive log likelihoods for AR(1) models with different error assumptions, applied to the logarithm of several EUR exchange rates.
+% included in \pkg{stochvol}'s dataset \code{exrates}.
  SV is strongly favored in all cases.}

### 2015-07-15 R/utilities_svdraws.R

@@ -32,6 +32,7 @@ updatesummary <- function(x, quantiles = c(.05, .5, .95), esspara = TRUE, esslat
  
  # NEW: Check if conditional t errors are used
  if (ncol(x$para) == 4) terr <- TRUE else terr <- FALSE
+ if (exists("betas", x)) regression <- TRUE else regression <- FALSE
 
  summaryfunction <- function(x, quants = quantiles, ess = TRUE) {
   if (ess) {
@@ -60,6 +61,8 @@ updatesummary <- function(x, quantiles = c(.05, .5, .95), esspara = TRUE, esslat
   res$sd <- t(apply(tmp, 2, summaryfunction, ess = esslatent))
   rownames(res$sd) <- gsub("h", "sd", rownames(res$latent))
  }
+
+ if (exists("beta", x)) res$beta <- t(apply(x$beta, 2, summaryfunction, ess = esspara))
  
  x$summary <- res
  x
@@ -161,22 +164,15 @@ predict.svdraws <- function(object, steps = 1L, ...) {
  sigma <- object$para[,"sigma"][usepara]
  hlast <- object$latent[,dim(object$latent)[2]][uselatent]
 
- # NEW: check if df parameter nu was sampled as well
- if (ncol(object$para) == 4) {
-  nu <- object$para[,"nu"][usepara]
- } else {
-  nu <- Inf  # corresponds to conditional normality
- }
-
  mythin <- max(thinpara, thinlatent)
  len <- length(sigma)
  volpred <- mcmc(matrix(as.numeric(NA), nrow=len, ncol=steps), start=mythin, end=len*mythin, thin=mythin)
  
- volpred[,1] <- mu + phi*(hlast - mu) + sigma*rt(len, df = nu)
+ volpred[,1] <- mu + phi*(hlast - mu) + sigma*rnorm(len)
  if (steps > 1)
   for (i in (seq.int(steps-1) + 1))
-   volpred[,i] <- mu + phi*(volpred[,i-1] - mu) + sigma*rt(len, df = nu)
- 
+   volpred[,i] <- mu + phi*(volpred[,i-1] - mu) + sigma*rnorm(len)
+
  class(volpred) <- c("svpredict", "mcmc")
  lastname <- dimnames(object$latent)[[2]][dim(object$latent)[2]]
  lastnumber <- as.integer(gsub("h_", "", lastname))
[…]

### 2015-07-15 R/wrappers.R

@@ -1,6 +1,6 @@
 # R wrapper function for the main MCMC loop
 
-svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), priorphi = c(5, 1.5), priorsigma = 1, priornu = NA, thinpara = 1, thinlatent = 1, thintime = 1, quiet = FALSE, startpara, startlatent, expert, ...) {
+svsample <- function(y, draws = 10000, burnin = 1000, designmatrix = NA, priormu = c(0, 100), priorphi = c(5, 1.5), priorsigma = 1, priornu = NA, priorbeta = c(0, 10000), thinpara = 1, thinlatent = 1, thintime = 1, quiet = FALSE, startpara, startlatent, expert, ...) {
  
  # Some error checking for y
  if (is(y, "svsim")) {
@@ -17,24 +17,49 @@ svsample <- function(y, draws = 10000, burnin = 1000, priormu = c(0, 100), prior
  } else myoffset <- 0
 
  # Some error checking for draws
- if (!is.numeric(draws) | draws < 1) {
+ if (!is.numeric(draws) || length(draws) != 1 || draws < 1) {
   stop("Argument 'draws' (number of MCMC iterations after burn-in) must be a single number >= 1.")
  } else {
   draws <- as.integer(draws)
  }
 
  # Some error checking for burnin
- if (!is.numeric(burnin) | burnin < 0) {
+ if (!is.numeric(burnin) || length(draws) != 1 || burnin < 0) {
   stop("Argument 'burnin' (burn-in period) must be a single number >= 0.")
  } else {
   burnin <- as.integer(burnin)
  }
  
+ # Some error checking for designmatrix
+ if (any(is.na(designmatrix))) {
+  designmatrix <- matrix(NA)
+ } else {
+  if (any(grep("ar[0-9]+$", as.character(designmatrix)[1]))) {
+   order <- as.integer(gsub("ar", "", as.character(designmatrix)))
+   if (length(y) <= order + 1) stop("Time series 'y' is to short for this AR process.")
+   designmatrix <- matrix(rep(1, length(y) - order), ncol = 1)
+   colnames(designmatrix) <- c("const")
+   if (order >= 1) {
+    for (i in 1:order) {
+     oldnames <- colnames(designmatrix)
+     designmatrix <- cbind(designmatrix, y[(order-i+1):(length(y)-i)])
[…]

### 2015-07-15 inst/doc/article.R

@@ -40,14 +40,14 @@ plot(sim)
 
 
 ###################################################
-### code chunk number 5: article.Rnw:240-242 (eval = FALSE)
+### code chunk number 5: article.Rnw:239-241 (eval = FALSE)
 ###################################################
 ## res <- svsample(ret, priormu = c(-10, 1), priorphi = c(20, 1.1),
 ##   priorsigma = 0.1)
 
 
 ###################################################
-### code chunk number 6: article.Rnw:245-251
+### code chunk number 6: article.Rnw:244-250
 ###################################################
 if (usePreCalcResultsExtra) {
  load("vignette_sampling_draws1.RData")
@@ -58,7 +58,7 @@ if (usePreCalcResultsExtra) {
 
 
 ###################################################
-### code chunk number 7: article.Rnw:273-274
+### code chunk number 7: article.Rnw:272-273
 ###################################################
 summary(res, showlatent = FALSE)
 
@@ -87,7 +87,7 @@ paratraceplot(res)
 ### code chunk number 11: usd6
 ###################################################
 par(mfrow = c(1, 3))
-paradensplot(res)
+paradensplot(res, showobs = FALSE)
 
 
 ###################################################
@@ -104,7 +104,7 @@ plot(myresid, ret)
 
 
 ###################################################
-### code chunk number 14: article.Rnw:405-411
+### code chunk number 14: article.Rnw:404-410
 ###################################################
 set.seed(123456)
 n <- 1000
@@ -115,7 +115,7 @@ y <- rnorm(n, X %*% beta.true, sigma.true)
 
 
 ###################################################
-### code chunk number 15: article.Rnw:415-421
+### code chunk number 15: article.Rnw:414-420
 ###################################################
 burnin <- 100
 draws <- 5000
@@ -126,7 +126,7 @@ C0 <- 0.001
 
 
 ###################################################
-### code chunk number 16: article.Rnw:424-428
+### code chunk number 16: article.Rnw:423-427
 ###################################################
 p <- ncol(X)
 preCov <- solve(crossprod(X) + B0inv)
@@ -135,7 +135,7 @@ preDf <- c0 + n/2 + p/2
 
 
 ###################################################
-### code chunk number 17: article.Rnw:431-434
+### code chunk number 17: article.Rnw:430-433
 ###################################################
 draws1 <- matrix(NA_real_, nrow = draws, ncol = p + 1)
[…]

### 2015-07-15 src/auxmix.h

@@ -1,7 +1,7 @@
 #ifndef _AUXMIX_H_
 #define _AUXMIX_H_
 
-#include <Rcpp.h>
+#include <RcppArmadillo.h>
 
 // Some constants relating to the approximation of log(chisq) trough
 // normal mixture (from Omori et al., 2007), and

### 2015-07-15 src/progutils.cpp

@@ -23,7 +23,8 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
 		   const Rcpp::NumericMatrix & hstore,
 		   const Rcpp::NumericVector & h0store,
 		   const Rcpp::NumericVector & nustore,
-		   const Rcpp::NumericMatrix & taustore) {
+		   const Rcpp::NumericMatrix & taustore,
+		   const Rcpp::NumericMatrix & betastore) {
  int paracols;
  if (nustore.size() > 0) paracols = 4; else paracols = 3;
  
@@ -40,7 +41,8 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
  Rcpp::List val = Rcpp::List::create(
    Rcpp::_["para"] = res,
    Rcpp::_["latent"] = hstore,
-   Rcpp::_["latent0"] = h0store);
+   Rcpp::_["latent0"] = h0store,
+   Rcpp::_["beta"] = betastore);
 //   Rcpp::_["tau"] = taustore);
 
  return val;

### 2015-07-15 src/progutils.h

@@ -16,7 +16,7 @@
       of dlogdnu (defined in densities.h)
 */
 
-#include <Rcpp.h>
+#include <RcppArmadillo.h>
 #include "densities.h"
 
 // a)
@@ -27,7 +27,8 @@ Rcpp::List cleanUp(const Rcpp::NumericVector & mu,
 		   const Rcpp::NumericMatrix & hstore,
 		   const Rcpp::NumericVector & h0store,
 		   const Rcpp::NumericVector & nustore,
-		   const Rcpp::NumericMatrix & taustore);
+		   const Rcpp::NumericMatrix & taustore,
+		   const Rcpp::NumericMatrix & betastore);
 
 // sets up the progress bar
 int progressbar_init(int N);

### 2015-07-15 src/sampler.cpp

@@ -12,14 +12,16 @@ using namespace Rcpp; // avoid to type Rcpp:: every time
 
 // RcppExport is an alias for 'extern "C"'
 RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
-  const SEXP burnin_in, const SEXP bmu_in, const SEXP Bmu_in,
+  const SEXP burnin_in, const SEXP X_in,
+  const SEXP bmu_in, const SEXP Bmu_in,
   const SEXP a0_in, const SEXP b0_in, const SEXP Bsigma_in,
   const SEXP thin_in, const SEXP timethin_in, const SEXP startpara_in,
   const SEXP startvol_in, const SEXP quiet_in, const SEXP para_in,
   const SEXP MHsteps_in, const SEXP B011_in, const SEXP B022_in,
   const SEXP mhcontrol_in, const SEXP gammaprior_in,
   const SEXP truncnormal_in, const SEXP offset_in,
-  const SEXP dontupdatemu_in, const SEXP priordf_in) {
+  const SEXP dontupdatemu_in, const SEXP priordf_in,
+  const SEXP priorbeta_in) {
 
  //RNGScope scope;       // just in case no seed has been set at R level
  GetRNGstate(); // "by hand" because RNGScope isn't safe if return
@@ -27,11 +29,26 @@ RcppExport SEXP sampler(const SEXP y_in, const SEXP draws_in,
 
  // convert SEXP into Rcpp-structures (no copy at this point)
  NumericVector y(y_in), startvol(startvol_in);
+ arma::vec armay(y.begin(), y.length(), false);
+
+ NumericMatrix X(X_in);
+
+ int T = y.length();
+ int p = X.ncol();
+ arma::mat armaX(X.begin(), T, p, false);
+ 
+ // should we model the mean as well?
+ bool regression;
+ if (ISNA(X(0,0))) regression = false; else regression = true;
+ NumericVector priorbeta(priorbeta_in);
+
+ // prior mean and precision matrix for the regression part (currently fixed)
+ arma::vec priorbetamean(p); priorbetamean.fill(priorbeta(0));
+ arma::mat priorbetaprec(p, p, arma::fill::zeros);
+ priorbetaprec.diag() += 1./(priorbeta(1)*priorbeta(1));
+ 
  List startpara(startpara_in);
 
- // length of time series
- int T = y.length();
- 
  // number of MCMC draws
  int burnin = as<int>(burnin_in);
  int draws  = as<int>(draws_in);
[…]

### 2015-07-15 src/sampler.h

@@ -3,7 +3,7 @@
 
 // Main sampling steps and helper functions
 
-#include <Rcpp.h>
+#include <RcppArmadillo.h>
 #define COMPILING_STOCHVOL
 #include <update.h>  // this header declares update
                      // (which is also available to other packages)
@@ -18,7 +18,8 @@ RcppExport SEXP sampler(const SEXP, const SEXP, const SEXP,
   const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
   const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
   const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
-  const SEXP, const SEXP, const SEXP, const SEXP, const SEXP);
+  const SEXP, const SEXP, const SEXP, const SEXP, const SEXP,
+  const SEXP, const SEXP);
 
 // Step (b): sample mu, phi, sigma - __CENTERED__ version:
 Rcpp::NumericVector regressionCentered(

### 2016-02-18 inst/doc/heavytails.Rnw
version 1.2.3

@@
   1020 Vienna, Austria\\
+  Telephone: +43/1/31336-5593\\
+  Fax: +43/1/31336-90-5593\\
   E-mail: \email{gregor.kastner@wu.ac.at}\\
-  URL: \url{http://www.wu.ac.at/statmath/en/faculty_staff/faculty/gkastner/}
+  URL: \url{http://statmath.wu.ac.at/~kastner/}
 }

### 2016-02-18 R/utilities_svdraws.R

@@ -1,5 +1,6 @@
 logret <- function(x, demean = FALSE) {
  tmp <- diff(log(x))
+ if (is(tmp, "xts")) tmp <- tmp[-1]
  if (all(isTRUE(demean))) tmp <- tmp - mean(tmp)
  tmp
 }
@@ -123,11 +124,11 @@ residuals.svdraws <- function(object, type = "mean", ...) {
  if (object$thinning$time != 1) warning("Not every point in time has been stored ('thintime' was set to a value unequal to 1 during sampling), thus only some residuals have been extracted.")
   
  if (type == "mean") {
-  res <- rowMeans(object$y[seq(1, length(object$y), by=object$thinning$time)] / exp(t(object$latent)/2))
+  res <- rowMeans(as.numeric(object$y)[seq(1, length(object$y), by=object$thinning$time)] / exp(t(object$latent)/2))
  }
  
  if (type == "median") {
-  res <- apply(object$y[seq(1, length(object$y), by=object$thinning$time)] / exp(t(object$latent)/2), 1, median)
+  res <- apply(as.numeric(object$y)[seq(1, length(object$y), by=object$thinning$time)] / exp(t(object$latent)/2), 1, median)
  }
 
  names(res) <- sub("h", "r", colnames(object$latent))

### 2020-11-17 R/exports.R
version 3.0.2

@@ -40,7 +40,7 @@
 #' \code{length(y)} times the number of covariates. If there are
 #' no covariates then this should be \code{matrix(NA)}
 #' @param priorspec a \code{priorspec} object created by
-#' \link{specify_priors}
+#' \code{\link{specify_priors}}
 #' @param thinpara single number greater or equal to 1, coercible to integer.
 #' Every \code{thinpara}th parameter draw is kept and returned. The default
 #' value is 1, corresponding to no thinning of the parameter draws i.e. every
@@ -54,14 +54,14 @@
 #' @param startpara named list, containing the starting values
 #' for the parameter draws. It must contain
 #' elements
-#'  * mu: an arbitrary numerical value
-#'  * phi: real number between \code{-1} and \code{1}
-#'  * sigma: a positive real number
-#'  * nu: a number larger than \code{2}; can be \code{Inf}
-#'  * rho: real number between \code{-1} and \code{1}
-#'  * beta: a numeric vector of the same length as the number of covariates
-#'  * latent0: either \code{'stationary'} or a positive number denoting the
-#'      prior variance of \code{h0}
+#' \itemize{
+#'   \item{mu: an arbitrary numerical value}
+#'   \item{phi: real number between \code{-1} and \code{1}}
+#'   \item{sigma: a positive real number}
+#'   \item{nu: a number larger than \code{2}; can be \code{Inf}}
+#'   \item{rho: real number between \code{-1} and \code{1}}
+#'   \item{beta: a numeric vector of the same length as the number of covariates}
+#'   \item{latent0: a single number, the initial value for \code{h0}}}
 #' @param startlatent vector of length \code{length(y)},
 #' containing the starting values for the latent log-volatility draws.
 #' @param keeptau Logical value indicating whether the 'variance inflation
@@ -70,12 +70,13 @@
 #' normal disturbance had to be 'upscaled' by a mixture factor and when the
 #' series behaved 'normally'.
 #' @param print_settings List of three elements:
-#'  * quiet: logical value indicating whether the progress bar and other
-#' informative output during sampling should be omitted
-#'  * n_chains: number of independent MCMC chains
-#'  * chain: index of this chain
+#'  \itemize{
[…]

### 2020-11-17 R/utilities_svdraws.R

@@ -174,8 +174,14 @@ runtime <- function (x) {
 #' @rdname extractors
 #' @export
 sampled_parameters <- function (x) {
+  pr <- if (inherits(x, "svdraws")) {
+    priors(x)
+  } else if (inherits(x, "sv_priorspec")) {
+    x
+  } else {
+    stop("Argument 'x' must be either an svdraws object or an sv_priorspec object.")
+  }
   res <- c()
-  pr <- priors(x)
   if (!inherits(pr$mu, "sv_constant")) {
     res <- c(res, "mu")
   }
@@ -446,7 +452,12 @@ print.svdraws <- function (x, showpara = TRUE, showlatent = FALSE, ...) {
 print.summary.svdraws <- function (x, showpriorbeta = !is.null(x$beta), ...) {
   cat("\nSummary of 'svdraws' object\n\n")
   print(x$priors, showbeta = showpriorbeta)
-  cat("\nStored ", x$mcp[2]-x$mcp[1]+x$mcp[3], " MCMC draws after a burn-in of ", x$mcp[1]-x$mcp[3], ".\n", sep="")
+  cat("\nStored ", (x$mcp[2] - x$mcp[1] + 2*x$mcp[3] - 1) %/% x$mcp[3], " MCMC draws after a burn-in of ", x$mcp[1]-x$mcp[3], ".\n", sep="")
+  if (x$mcp[3] == 1) {
+    cat("No thinning.\n")
+  } else {
+    cat("Thinning: ", x$mcp[3], ".\n", sep="")
+  }
   
   if (exists("para", x)) {
     cat("\nPosterior draws of SV parameters (thinning = ", x$mcp[3], "):\n", sep='')
@@ -679,13 +690,13 @@ summary.svdraws_roll <- function (object, ...) {
   have_predlik <- !is.null(object[[1]]$predictive_likelihood)
   have_predquant <- length(object[[1]]$quantiles) > 0
   if (have_predlik) {
-    cat("Predictive likelihoods\n\n")
-    sapply(object, function (xx) xx$predictive_likelihood)
-    cat("\n\n")
+    cat(" Summary of predictive likelihoods\n")
+    print(summary(sapply(object, function (xx) xx$predictive_likelihood)))
+    cat("\n")
   }
   if (have_predquant) {
-    cat("Summary of predicted quantiles\n\n")
-    sapply(object, function (xx) xx$predicted_quantile)
+    cat(" Summary of predicted quantiles\n")
[…]

### 2020-11-17 R/wrappers.R

@@ -354,7 +354,7 @@ svsample <- function(y, draws = 10000, burnin = 1000, designmatrix = NA,
   assert_positive(n_chains, "Parameter for number of chains")
 
   ## expert
-  expert <- validate_and_process_expert(expert)
+  expert <- validate_and_process_expert(expert, priorspec)
   correct_model_misspecification <- expert$correct_model_misspecification
   interweave <- expert$interweave
   fast_sv <- expert$fast_sv
@@ -594,9 +594,17 @@ svsample <- function(y, draws = 10000, burnin = 1000, designmatrix = NA,
   res
 }
 
-#' @rdname svsample_cpp
+#' Default Values for the Expert Settings
+#' 
+#' These functions define meaningful expert settings for
+#' argument \code{expert} of \code{\link{svsample}} and its derivatives.
+#' The result of \code{get_default_fast_sv} should be provided as
+#' \code{expert$fast_sv} and \code{get_default_general_sv} as \code{expert$general_sv}
+#' when relevant.
+#' @rdname expert_defaults
+#' @seealso \code{link{svsample}}, \code{\link{specify_priors}}, \code{\link{svsample_roll}}, \code{\link{svsample_fast_cpp}}, \code{\link{svsample_general_cpp}}
 #' @export
-default_fast_sv <- 
+get_default_fast_sv <- function () {
   list(baseline_parameterization = "centered",  # "centered" or "noncentered"
        proposal_phi = "immediate acceptance-rejection",  # "immediate acceptance-rejection" or "repeated acceptance-rejection"
        proposal_sigma2 = "independence",  # "independence" or "log random walk"
@@ -608,14 +616,56 @@ default_fast_sv <-
        update = list(latent_vector = TRUE, parameters = TRUE, mixture_indicators = TRUE),
        init_indicators = 5,
        init_tau = 1)
-#' @rdname svsample_cpp
+}
+
+#' @rdname expert_defaults
+#' @param priorspec a \code{priorspec} object created by
+#' \code{\link{specify_priors}}
 #' @export
-default_general_sv <-
-  list(multi_asis = 5,  # positive integer
+get_default_general_sv <- function (priorspec) {
+  if (!inherits(priorspec, "sv_priorspec")) {
+    stop("Argument 'priorspec' must be an sv_priorspec object.")
+  }
+  multi_asis <- 5
+  list(multi_asis = multi_asis,  # positive integer
        starting_parameterization = "centered",  # "centered" or "noncentered"
[…]

### 2020-11-17 inst/doc/article2.R

@@ -110,97 +110,7 @@ par(opar)
 #    beta = sv_multinormal(mean = 0, sd = 10000, dim = 1))
 #  svsample(CHF_logret, priorspec = ps)
 
-## ----fsvprepdata, echo=TRUE, fig.width=10, fig.height=5-----------------------
-library("factorstochvol")
-data("exrates", package = "stochvol")
-m <- 6
-n <- 1000
-y <- 100 * logret(tail(exrates[, seq_len(m)], n + 1), demean = TRUE)
-plot.ts(y)
-
-## ----preorder, echo=TRUE------------------------------------------------------
-preorder(y, factors = 2)
-
-## ----findrestrict, echo=TRUE--------------------------------------------------
-findrestrict(y, factors = 2)
-
-## ----runmodel, echo=2:3-------------------------------------------------------
-set.seed(1)
-library(factorstochvol)
-res <- fsvsample(y, factors = 2, draws = 10000, thin = 10, quiet = TRUE)
-
-## ----printrres, echo = TRUE---------------------------------------------------
-res
-
-## ----covn, echo = TRUE--------------------------------------------------------
-dim(cov_n <- covmat(res))
-
-## ----logdetcovn, echo = 2:4, fig.width = 10, fig.height=3.5-------------------
-logdet <- function (x) log(det(x))
-opar <- par(mfrow = c(1, 2), mgp = c(1.7, 0.5, 0), mar = c(3, 3, 1, 1))
-logdet_n <- apply(cov_n[,,,1], 3, logdet)
-ts.plot(logdet_n)
-acf(logdet_n, main = "")
-par(opar)
-
-## ----covess, echo = TRUE------------------------------------------------------
-round(apply(cov_n, 1:2, coda::effectiveSize))
-
-## ----corimageplot, echo=2-----------------------------------------------------
-opar <- par(mfrow = c(1, 3), xpd = TRUE)
-corimageplot(res, these = seq(1, n, length.out = 3), plotCI = "circle",
-  plotdatedist = 2, date.cex = 1.1)
-par(opar)
-
-## ----voltimeplot, echo=2:3, fig.width = 10, fig.height = 3, cache.rebuild = TRUE----
-opar <- par(mgp = c(1.7, 0.5, 0), mar = c(2, 1.5, 1, 0.5))
[…]

### 2020-11-17 inst/include/adaptation.hpp

@@ -1,9 +1,32 @@
+/*
+ * R package stochvol by
+ *     Gregor Kastner Copyright (C) 2013-2020
+ *     Darjus Hosszejni Copyright (C) 2019-2020
+ *  
+ *  This file is part of the R package stochvol: Efficient Bayesian
+ *  Inference for Stochastic Volatility Models.
+ *  
+ *  The R package stochvol is free software: you can redistribute it
+ *  and/or modify it under the terms of the GNU General Public License
+ *  as published by the Free Software Foundation, either version 2 or
+ *  any later version of the License.
+ *  
+ *  The R package stochvol is distributed in the hope that it will be
+ *  useful, but WITHOUT ANY WARRANTY; without even the implied warranty
+ *  of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU
+ *  General Public License for more details.
+ *  
+ *  You should have received a copy of the GNU General Public License
+ *  along with the R package stochvol. If that is not the case, please
+ *  refer to <http://www.gnu.org/licenses/>.
+ */
+
 #ifndef _STOCHVOL_ADAPTATION_H_
 #define _STOCHVOL_ADAPTATION_H_
 
 #include <RcppArmadillo.h>
 #include <vector>
-#include "type_definitions.h"
+#include "type_definitions.hpp"
 
 namespace stochvol {
   // Encapsulate adaptation logic
@@ -14,12 +37,12 @@ namespace stochvol {
   public:
     // Useful structs
     struct Storage {
-      const double gamma;
-      const double scale;
-      const double rate_acceptance;
+      double gamma;
+      double scale;
+      double rate_acceptance;
     };
 
-    // Constructor
+    // Constructor / initializes from zero knowledge
     Adaptation (
         const int _dim,
         const int _memory_size,
@@ -27,33 +50,156 @@ namespace stochvol {
         const double _target_acceptance = 0.234,
         const double _lambda = 0.1,
         const double _scale = 0.1,
-        const double _C_in = 0.99)  // _C is probably a macro on Solaris
-      : dim{_dim},
-        target_acceptance{_target_acceptance},
+        const double _C_in = 0.99)  // _C is reserved name in the C++ standard
+      : target_acceptance{_target_acceptance},
         lambda{_lambda},
         alpha{calculate_alpha(_lambda)},
         C{_C_in},
         scale{_scale},
         state(_dim, _batch_size),
-        draws_batch(dim, _batch_size),
+        draws_batch(_dim, _batch_size),
[…]

### 2020-11-17 inst/include/stochvol.h

@@ -1,9 +1,33 @@
+/*
+ * R package stochvol by
+ *     Gregor Kastner Copyright (C) 2013-2020
+ *     Darjus Hosszejni Copyright (C) 2019-2020
+ *  
+ *  This file is part of the R package stochvol: Efficient Bayesian
+ *  Inference for Stochastic Volatility Models.
+ *  
+ *  The R package stochvol is free software: you can redistribute it
+ *  and/or modify it under the terms of the GNU General Public License
+ *  as published by the Free Software Foundation, either version 2 or
+ *  any later version of the License.
+ *  
+ *  The R package stochvol is distributed in the hope that it will be
+ *  useful, but WITHOUT ANY WARRANTY; without even the implied warranty
+ *  of MERCHANTABILITY or FITNESS FOR A PARTICULAR PURPOSE. See the GNU
+ *  General Public License for more details.
+ *  
+ *  You should have received a copy of the GNU General Public License
+ *  along with the R package stochvol. If that is not the case, please
+ *  refer to <http://www.gnu.org/licenses/>.
+ */
+
 #ifndef stochvol_H_
 #define stochvol_H_
 
 #include <RcppArmadillo.h>
-#include "adaptation.hpp"
 #include "type_definitions.h"
+#include "adaptation.hpp"
+#include "expert.hpp"
 
 namespace stochvol {
 

### 2020-11-17 inst/include/type_definitions.h

@@ -1,290 +1,7 @@
-#ifndef _TYPE_DEFINITIONS_H_
-#define _TYPE_DEFINITIONS_H_
+#ifndef _TYPE_DEFINITIONS_H_OLD_STOCHVOL_
+#define _TYPE_DEFINITIONS_H_OLD_STOCHVOL_
 
-#include <RcppArmadillo.h>
-#include <vector>
+#include "type_definitions.hpp"
+#include "expert.hpp"
 
-namespace stochvol {
-
-  enum class Parameterization {CENTERED, NONCENTERED};
-
-  enum class Proposal {RWMH, MALA};
-
-  // Class specifying the prior distributions
-  // Implementation stays at the "end user functions"
-  struct PriorSpec {
-    // Recognized distributions
-    struct Constant {
-      double value;
-      Constant (const double _v) : value{_v} {}
-    };
-    struct Normal {
-      double mean, sd;
-      Normal (const double _m, const double _s) : mean{_m}, sd{_s} {}
-    };
-    struct MultivariateNormal {
-      arma::vec mean;
-      arma::mat precision;
-    };
-    struct Gamma {
-      double shape, rate;
-      Gamma (const double _s, const double _r) : shape{_s}, rate{_r} {}
-    };
-    struct InverseGamma {
-      double shape, scale;
-      InverseGamma (const double _sh, const double _sc) : shape{_sh}, scale{_sc} {}
-    };
-    struct Beta {
-      double alpha, beta;
-      Beta (const double _a, const double _b) : alpha{_a}, beta{_b} {}
-    };
-    struct Exponential {
-      double rate;
-      Exponential (const double _r) : rate{_r} {}
-    };
-    struct Infinity {};
-
-    // Parameter classes
-    struct Mu {
-      enum {CONSTANT, NORMAL} distribution;
-      union {
-        Constant constant;
-        Normal normal;
-      };
-
-      Mu (const Constant& _c) : distribution{CONSTANT}, constant{_c} {}
-      Mu (const Normal& _n) : distribution{NORMAL}, normal{_n} {}
-    };
-
-    struct Phi {
-      enum {CONSTANT, BETA, NORMAL} distribution;
-      union {
-        Constant constant;
-        Beta beta;
-        Normal normal;
-      };
-
-      Phi (const Constant& _c) : distribution{CONSTANT}, constant{_c} {}
-      Phi (const Beta& _b) : distribution{BETA}, beta{_b} {}
[…]

### 2020-11-17 src/sampling_main.cc

@@ -29,10 +29,10 @@
  */
 
 #include <RcppArmadillo.h>
+#include <expert.hpp>
 #include <adaptation.hpp>
 #include "sampling_main.h"
 #include "single_update.h"
-#include <type_definitions.h>
 #include "utils_main.h"
 #include "utils_latent_states.h"
 #include "utils.h"
@@ -123,11 +123,11 @@ List svsample_fast_cpp(
 
   // storage
   const bool keep_r = expert_in["store_indicators"];
-  const int para_draws = draws / thinpara + 1;
+  const int para_draws = draws / thinpara;
   Rcpp::NumericMatrix para_store(5, para_draws);
   Rcpp::NumericMatrix beta_store(p, is_regression * para_draws);
   const int latent_length = T / thintime;  // thintime must be either 1 or T
-  const int latent_draws = draws / thinlatent + 1;
+  const int latent_draws = draws / thinlatent;
   Rcpp::NumericVector latent0_store(latent_draws);
   Rcpp::NumericMatrix latent_store(latent_length, latent_draws);
   Rcpp::NumericMatrix tau_store(latent_length, keep_tau * latent_draws);
@@ -156,8 +156,8 @@ List svsample_fast_cpp(
       ::R_CheckUserInterrupt();
     }
 
-    const bool thinpara_round = (thinpara > 1) and (i % thinpara != 0);  // is this a parameter thinning round?
-    const bool thinlatent_round = (thinlatent > 1) and (i % thinlatent != 0);  // is this a latent thinning round?
+    const bool parasave_round = (i - 1) % thinpara == thinpara - 1,  // is this a parameter saving round?
+               latentsave_round = (i - 1) % thinlatent == thinlatent - 1;  // is this a latent saving round?
 
     // print a progress sign every "show" iterations
     if (verbose and i % show == 0) {
@@ -206,19 +206,21 @@ List svsample_fast_cpp(
 
     // store draws
     double correction_weight_i = 1;
-    if ((not thinpara_round or not thinlatent_round) and correct_model_specification) {
+    if ((parasave_round or latentsave_round) and correct_model_specification) {
       correction_weight_i = fast_sv::compute_correction_weight(data_demean, log_data2_normal, h, 1/exp_h_half_inv);
     }
-    if (i >= 1 and not thinpara_round) {
-      save_para_sample(i / thinpara - 1, mu, phi, sigma, nu, beta, para_store, beta_store, is_regression);
[…]

### 2020-11-17 src/utils_main.h

@@ -33,7 +33,7 @@
 // Helper functions for the main sampler
 
 #include <RcppArmadillo.h>
-#include <type_definitions.h>
+#include <expert.hpp>
 #include <adaptation.hpp>
 
 namespace stochvol {
@@ -83,8 +83,7 @@ inline
 void clamp_log_data2(
     arma::vec& log_data2) {
   // -100 ~= log(4e-44)
-  static const auto my_clamp { [](double& value) { value = std::max(value, -100.0); } };
-  std::for_each(log_data2.begin(), log_data2.end(), my_clamp);
+  std::for_each(log_data2.begin(), log_data2.end(), [](double& value) { value = std::max(value, -100.0); });
 }
 
 namespace fast_sv {
@@ -306,6 +305,16 @@ ExpertSpec_GeneralSV list_to_general_sv(
 ExpertSpec_FastSV list_to_fast_sv(
     const Rcpp::List& list,
     const bool interweave);
+
+// Transform an R adaptation list into its corresponding
+// C++ object
+Adaptation list_to_adaptation (
+    const Rcpp::List& list);
+
+// Transform an R adaptation collection list into
+// its corresponding C++ object
+AdaptationCollection list_to_adaptationcollection (
+    const Rcpp::List& list);
 
 }
 
